---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: axis scale
# author: jonathan
# revised: 15/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Essential/axis-scale.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Essential/axis-scale.Rmd")
library(ggplot2)
library(gridExtra)

# variable 1
values1 <- 1:100
# variable 2
values2 <- values1 * (rnorm(100, mean = 1, sd = .1))

# dataframe
df <- data.frame(values1, values2)
```
The clarity of a plot sometimes relies on proper scaling of the axes. Very often, the scale is linear; `ggplot()` displays that by default. But your data might be  represented in a more meaningful way if the Y-axis is set to logarithmic, for example. 

In ggplot, the scale can be set to `log10`, `log2`, `sqrt` (square root), and `reverse` (to invert the axis). Here we will see how to do this using a simple example.

Let’s start with the code for a reference plot. This plot is stored in the object `baseplot` so that we can reuse it throughout the whole tutorial:

```{r plot}
baseplot <- ggplot(df, aes(values1, values2)) +
  geom_point(size = 2)
baseplot
```
<br/><br/>

### log10

Here, we use `scale_y_continuous()` to transform the Y-axis along with the argument `trans = "log10"` to set the log10 scale:

```{r log10}
baseplot +
  scale_y_continuous(trans = "log10")
```
<br/><br/>

### log2

Use `scale_y_continuous()` with `trans = "log2"` to set the log2 scale:

```{r log2}
baseplot +
  scale_y_continuous(trans = "log2")
```
<br/><br/>

### square root (sqrt)

Use `scale_y_continuous()` with `trans = "sqrt"` to set the square root axis scale:

```{r sqrt}
baseplot +
  scale_y_continuous(trans = "sqrt")
```
<br/><br/>

### reverse

Use `scale_y_continuous()` with `trans = "reverse"`  to invert the Y-axis scale:

```{r rev}
baseplot +
  scale_y_continuous(trans = "reverse")
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
