---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: axis lines
# author: jonathan
# revised: 15/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Essential/axis-lines.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Essential/axis-lines.Rmd")
library(ggplot2)

# variable 1
values1 <- 1:100
# variable 2
values2 <- values1 * (rnorm(100, mean = 1, sd = .1))

# dataframe
df <- data.frame(values1, values2)
```
ggplot does not draw axis lines by default. If you desperately miss them, this post is for you. Adding X- and Y-axis lines to your plot implies that you modify the *theme* of the plot itself. Nothing very complicated about it, as you will see further below.

Let’s start with the code for a reference plot. This plot is stored in the object `baseplot` so that we can reuse it throughout the whole tutorial:
```{r plot}
baseplot <- ggplot(df, aes(values1, values2)) +
  geom_point(size = 2)
baseplot
```
<br/><br/>

To add both X- and Y-axis lines to your plot, add an extra line of code and write `theme(axis.line = element_line())`. In the parentheses, you may add additional arguments for modifying the size/thickness (`size=`), color (`color=`) and pattern of the line (`linetype=`):

```{r lines}
baseplot +
  theme(axis.line = element_line(linetype = "solid", size = 1, colour = "darkblue"))
```
<br/><br/>

If you wish to use different parameters for each axis, use `axis.line.x` and `axis.line.y` to operate on the X- and Y-axis separately:
```{r lines2}
baseplot +
  theme(
    axis.line.y = element_line(size = 1, colour = "darkblue", linetype = "solid"),
    axis.line.x = element_line(size = 1.25, colour = "red", linetype = "dotted")
  )
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
