---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: Color scale for continuous variables
# author: jonathan
# revised: 15/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Essential/color-scale-continuous.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Essential/color-scale-continuous.Rmd")
library(ggplot2)
library(gridExtra)
# variable 1
values1 <- c(rnorm(100, mean = 50, sd = 15), rnorm(100, mean = 75, sd = 15), rnorm(100, mean = 100, sd = 15))
# variable 2
values2 <- c(rnorm(100, mean = 20, sd = 10), rnorm(100, mean = 30, sd = 10), rnorm(100, mean = 40, sd = 10))
# variable 3
values3 <- c(rnorm(100, mean = 2, sd = 1), rnorm(100, mean = 3, sd = 1), rnorm(100, mean = 4, sd = 1))
# dataframe
df <- data.frame(values1, values2, values3)
```
Here we will see how to add/change colors to a plot when working with *continuous* variables. To illustrate this, we will use a scatter plot with 3 continuous variables: *values1* which is plotted along the X-axis, *values2* which is plotted along the Y-axis, and *values3* which range will be represented with colors. Here is the code for this reference plot without colors (i.e. *values3* is not involved yet):
```{r code}
ggplot(df, aes(values1, values2)) +
  geom_point()
```
<br/><br/>

### Bringing colors to the dots

To bring colors to the dots as a function of *values3*, we use the argument `color=` in `aes()`, either directly *within* `ggplot()` or *within `geom_point()`*. In the following example, we use `geom_point(aes(color=...))`. Note that we also use `size=2` to increase the size of the dots:
```{r color}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(color = values3), size = 2)
```

The settings by default tell ggplot to use a blue gradient to represent the variable `values3`. This gradient, represented in the legend to the right of the plot, is indexed on `values3`. The lower the value of `values3`, the darker the color of the dot.
<br/><br/>

### Changing the color palette

Instead of a relatively simple blue gradient, you may want to use a palette of nicer colors. The palettes provided by `viridis` or `brewer/distiller` are among the most frequently used. Note that the functions that will apply for continuous variables are:

+ `scale_color_viridis_c()`,
+ `scale_color_distiller()`.

Their use is further explained [HERE](https://biostats.w.uib.no/choosing-colors-for-your-plot/).
<br/><br/>

### Using a gradient based on two given colors

Now, if you do not find the exact set of colors that you want, you may use the color/gradient of your choice by adding `scale_color_gradient()` with the arguments `low=` and `high=` to the original code. The function will produce a two-color gradient ranging from the color indicated by `low=` to the color indicated by `high=`. Whether you indicate the colors by their [R name](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf){target="_blank"} or by their [Hex code](https://htmlcolorcodes.com/) is up to you. Here are two examples:
```{r gradient1, eval=FALSE}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(color = values3), size = 2) +
  scale_color_gradient(low = "yellow", high = "darkblue") # left plot, colors coded with R names
ggplot(df, aes(values1, values2)) +
  geom_point(aes(color = values3), size = 2) +
  scale_color_gradient(low = "#AF7AC5", high = "#E74C3C") # right plot, colors coded with Hex codes
```
```{r gradient2, echo=FALSE, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values1, values2)) +
    geom_point(aes(color = values3), size = 2) +
    scale_color_gradient(low = "yellow", high = "darkblue"), # left plot, colors coded with R names
  ggplot(df, aes(values1, values2)) +
    geom_point(aes(color = values3), size = 2) +
    scale_color_gradient(low = "#AF7AC5", high = "#E74C3C"), # right plot, colors coded with Hex codes
  nrow = 1
)
```
<br/><br/>

### Using a gradient based on two given colors and a breaking point

`scale_fill_gradient2()` is a function that also produces a gradient of colors, but this time based on three colors defined by `low=`, `mid=` and `high=`. This color `mid=` is special in the way that you can use it to highlight a breaking point in your data (encoded by the argument `midpoint=`). For example, if the average of the variable `values3` has a special meaning for the interpretation of the data, you may set `midpoint = mean(values3)`. Thus all values close to the average will get a color close to the one defined by `mid=`. Here is the code and the result:
```{r gradient3}
ggplot(df, aes(values1, values2, values3)) +
  geom_point(aes(color = values3), size = 2) +
  scale_color_gradient2(low = "yellow", mid = "darkblue", high = "red", midpoint = mean(values3))
```
<br/><br/>



***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
