---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Linear regression
# author: jonathan
# revised: 3/2/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/regression/linear-regression.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/regression/linear-regression.Rmd")
library(ggplot2)
```

**Linear regression** helps you simplify a dataset by modelling and drawing a straight line representing this dataset. It is often used to find a relationship between a *continuous* response variable and a *continuous* predictor variable. Examples are numerous: finding the relationship between bodyweight and height is one of them. 

Let’s use the following dataset as an example:
```{r dataframe}
# variable1
bodyweight <- c(70, 75, 72, 58, 80, 80, 48, 56, 103, 51)

# variable2
size <- c(177, 178, 167, 153, 174, 177, 152, 134, 191, 136)

# dataframe
my.dataframe <- data.frame(bodyweight, size)
```
<br/><br/>

As usual, everything starts with a plot. A scatter plot of the dataset is usually a good beginning to assess relationship between continuous variables:
```{r scatter}
ggplot(my.dataframe, aes(size, bodyweight)) +
  geom_point(color = "blue", size = 2.5)
```
<br/><br/>

### Fitting a linear model with `lm()`

We may try to fit a linear model with the function `lm()` which we have already encountered when performing [analysis of variance (ANOVA)](https://biostats.w.uib.no/one-way-anova/){target="_blank"}.

```{r lm}
lm(bodyweight ~ size)
```
The output gives us everything we need to draw a line in our original plot:

 + the intercept is -56.2716,  
 + the slope is 0.7661.
 
Using the function `summary()`, we get much more from the linear model. We get of course the slope and intercept in the column *Estimates* in the table *coefficients*, but we also get the range, interquartile range and median, among other useful values:
 
```{r lm-summary}
summary(lm(bodyweight ~ size))
```
Of interest is the (adjusted) R-squared (R^2^) at the bottom of the output, which describes how well the model matches the data (NB: be careful when interpreting R-squared, see this [blogpost](https://blog.minitab.com/blog/adventures-in-statistics-2/regression-analysis-how-do-i-interpret-r-squared-and-assess-the-goodness-of-fit){target="_blank"} for some info).
<br/><br/>

### Plotting the line

Having these values, we may thus add  this line "manually" to our previous scatter plot:
```{r scatter-and-line}
ggplot(my.dataframe, aes(size, bodyweight)) +
  geom_point(color = "blue", size = 2.5) +
  geom_abline(slope = 0.7661, intercept = -56.2716)
```
<br/><br/>

Note that the function `geom_smooth()` in `ggplot2` can draw the line directly on top of the scatter plot. Make sure that you use the argument `method="lm"`:
```{r scatter-and-smoother}
ggplot(my.dataframe, aes(size, bodyweight)) +
  geom_point(color = "blue", size = 2.5) +
  geom_smooth(method = "lm")
```

The gray area along the line is the confidence interval. If this is irrelevant, it may be removed with the extra argument `se=FALSE`:
```{r scatter-smoother-noSE}
ggplot(my.dataframe, aes(size, bodyweight)) +
  geom_point(color = "blue", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE)
```


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
