---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Test for homogeneity of variances – Levene’s test and the Fligner Killeen test
# author: jonathan
# revised: 30/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/useful-tests/homogeneity.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/useful-tests/homogeneity.Rmd")
library(ggplot2)
library(car)
```

2 tests are commonly used to check for **homogeneity of variance**: [**Fisher’s F test**](https://biostats.w.uib.no/1-comparing-two-variances/){target="_blank"} and **Levene’s test**. Fisher’s F test, which is introduced [here](https://biostats.w.uib.no/1-comparing-two-variances/){target="_blank"}, is restricted to comparison of two variances/groups while Levene's test can assess more than two variances/groups. Like normality, verifying homogeneity of variance is a requirement for running several of the well-known statistical tests such as [Student's t-test](https://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"} and [ANOVA](https://biostats.w.uib.no/topics/basic-stats-in-r/analysis-of-variance-anova/){target="_blank"}. 

Here we will see how to perform Levene's test. In this test, the null hypothesis `H0` is that all variances are equal. Note that this test is meant to be used with *normally distributed data*, but can tolerate relatively low deviation from normality. The corresponding function in R is `leveneTest()` and the syntax is `leveneTest(response ~ predictor, data, center)` where `response` is the response variable, `predictor` is the predictor variable that contains the groups to compare and `data` is the dataframe. One can also add the optional argument `center = mean` or `center = median` to define how the function will compute the center of each group. `mean` gives the original Levene's test; the default, `median`, provides a more robust test called **Brown-Forsythe test** for homogeneity of variance. 

`leveneTest()` is part of the package `car`, so start with activating it:
```{r car, eval=FALSE}
library(car)
```
<br/><br/>

Here we will consider the same example as introduced [here](http://biostats.w.uib.no/one-way-anova/){target="_blank"}:
```{r dataframe}
# response variable
size <- c(25, 22, 28, 24, 26, 24, 22, 21, 23, 25, 26, 30, 25, 24, 21, 27, 28, 23, 25, 24, 20, 22, 24, 23, 22, 24, 20, 19, 21, 22)

# predictor variable
location <- c(rep("ForestA", 10), rep("ForestB", 10), rep("ForestC", 10))

# dataframe
my.dataframe <- data.frame(size, location)
```
<br/><br/>
We may first visualize the data with a boxplot:
```{r boxplot, echo=FALSE}
ggplot(my.dataframe, aes(x = location, y = size)) +
  geom_boxplot()
```
<br/><br/> 

### Running the test

The syntax adapted to our example is as follow: 
```{r Levene}
leveneTest(size ~ location, my.dataframe, center = mean)
```
The test reveals a p-value greater than 0.05, indicating that there is no significant difference between the group variances in `location`.

<br/><br/> 

### Alternative

The **Fligner-Killeen test** does a rather similar job, meaning that it checks for homogeneity of variance, but is a better option when data are non-normally distributed or when problems related to outliers in the dataset cannot be resolved.

The function is `fligner.test()` and the syntax is `fligner.test(response ~ predictor, data)` which is very similar `leveneTest( )`.
```{r Fligner}
fligner.test(size ~ location, my.dataframe)
```
and again the p-value is greater than 0.05; the variances are homogeneous.


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
