---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Test for normality – Shapiro-Wilks test
# author: jonathan
# revised: 30/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/useful-tests/shapiro.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/useful-tests/shapiro.Rmd")
library(ggplot2)
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22)
df <- data.frame(size)
```

The **Shapiro-Wilk test** is a test of **normality** that assesses whether a sample is likely to originate from a normal distribution. Verifying normality of distribution is a requirement for running several of the well-known statistical tests such as [Student's t-test](https://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"} and [ANOVA](https://biostats.w.uib.no/topics/basic-stats-in-r/analysis-of-variance-anova/){target="_blank"}. 

In this test, the null hypothesis `H0` states that the sample comes from a normally distributed population.

The function to use in R is `shapiro.test()` and the syntax is `shapiro.test(sample)` where `sample` is the vector containing the data.

Let’s take an example where we measured the size of blue ground beetles (*Carabus intricatus*) at a given location. Here is the code for the vector containing the data:
```{r data}
# sample data
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22)
```

<br/><br/>
To visualize the distribution of the sample, we may use a [Q-Q plot](https://biostats.w.uib.no/creating-a-q-q-plot/){target="_blank"} with a quantile-quantile line:
```{r QQ, echo=FALSE}
ggplot(df, aes(sample=size)) +
  stat_qq(size = 2, color = "blue") + 
  stat_qq_line(size=1.25, color="red")
```  
<br/><br/>
The rather good alignment of the dots in this plot shows that the distribution is close to normality, but this needs to be verified with a test.
<br/><br/>

### Running the test
We use `shapiro.test()` the following way:
```{r shapiro}
shapiro.test(size)
```
Apparently, the p-value is rather high, well over 0.05. The null hypothesis `H0` is NOT rejected, meaning that the sample is very likely to be normally distributed.


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
