```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: median
# author: jonathan
# revised: 8/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Descr/median.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/median.Rmd")
```

The **median** is another measure of the central tendency of your sample. The median is the number “in the middle” of the dataset, the middle value of a series of values sorted from the smallest to the largest.

Providing that your dataset contain an **odd number *n*** of entries, the median is the value of the **((n+1)/2)^th^ entry** in the dataset. If the dataset contains an **even number *n*** of entries, then the median is the ***average* of the values of the (n/2)^th^ and ((n/2)+1)^th^ entries**.

In R, the function that calculates the median is called `median()`:

```{r median}
my.dataset <- c(1,2,3,4,5,6,7,8,9,10)
median(my.dataset)
```
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
