```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: minimum, maximum and range
# author: jonathan
# revised: 8/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Descr/minmaxrange.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/minmaxrange.Rmd")
```

The **minimum** and **maximum** of a dataset are the smallest and the largest entries, respectively. No surprise here…

The **range** is the difference between the maximum and the minimum, and defines the spread of the data.

Note that a range may be expressed as a single value (the actual difference between maximum and minimum) or by writing `minimum-maximum`. For instance, in the series (1, 2, 3, 4, 5, 6, 7, 8, 9, 10), the range is 9 or 1-10.

In R, the functions to be used are `min()`, `max()` and `range()`. However, `range()` returns a vector that contains the minimum and the maximum, respectively. To get the actual difference between these values, you may use `diff(range())`.

```{r minmaxrange}
my.dataset <- c(1,2,3,4,5,6,7,8,9,10)
min(my.dataset)
max(my.dataset)
range(my.dataset)
diff(range(my.dataset))
```
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
