---
title: "Viz1-TheCode"
output:
  learnr::tutorial:
    highlight: tango
    css: css/style-apps.css
runtime: shiny_prerendered
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(learnr)
library(ggplot2)

```

The following exercises are based on the data frame `iris`, which contains the variables `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width`, and `Species`. This data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are *Iris setosa*, *versicolor*, and *virginica*. Reference: https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris

## Make a simple scatter plot

Add the required R code to plot `Petal.Length` against `Petal.Width`:

```{r ex1, exercise=TRUE}
ggplot(data = , aes(x = , y = )) +
  geom_point() +
  theme_minimal()

```

```{r ex1-hint}
# you have to set `iris` as the source of the data
# you have to set `Petal.Width` and `Petal.Length` as `x` and `y`

```

```{r ex1-solution}
# SOLUTION

ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) +
  geom_point() +
  theme_minimal()

```

## Add colors to represent species

Add the required R code to colour each species in a different colour.

```{r ex2, exercise=TRUE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, colour = )) +
  geom_point()  +
  theme_minimal()

```

```{r ex2-hint}
# you have to set `Species` as the variable that will define the colour in aes()

```

```{r ex2-solution}
# SOLUTION

ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) +
  geom_point() +
  theme_minimal()

```

## Change the shapes 

Add the required R code to plot the data as triangles.
Use [this reference table](https://lh3.googleusercontent.com/proxy/oqsNQ84iXiqOyhYu4Tpv7wqCtbvUn12MOF-CoOyvYzX9zMkbpFwgtWkTx3WBX2I5kuBOAx6xnzhJSYoKxG6h7_Q9os9XBEHyVCQPwBZkwRab7FsVqF2z-CR9CaRRwmGxnmUSIS_83Ws) if you need the code for the shapes.

```{r ex3, exercise=TRUE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, colour = Species)) +
  geom_point()  +
  theme_minimal()

```

```{r ex3-hint}
# you have to set `shape = 17` in geom_point()

```

```{r ex3-solution}
# SOLUTION

ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) +
  geom_point(shape = 17) +
  theme_minimal()

```