---
title: "Working with dates and times"
description: Manipulating dates and times with the lubridate package
output:
  bookdown::html_document2: 
    highlight: tango
    toc: yes
    toc_float: yes
    css: ../css/style-chapters.css
---

``` {r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, rows.print = 3)
library("tidyverse")
library("lubridate")
library("conflicted")
conflict_prefer(name = "filter", winner = "dplyr")

temperature <- read_csv("Florida_2020-04-20_2020-04-23_1589018968.csv")

temperature2 <- temperature %>% 
  mutate(
    date_time = paste(Dato, Tid), #combine
    date_time = ymd_hms(date_time)#convert
    ) 

source("../Templates/biostats_theme.R")
```



## Date and times

Many data sets have dates and times.

In this tutorial, you will learn how to

- make dates and times in R
- manipulate dates and times
- use dates and times in data.frames with `dplyr`
- plot dates and times with `ggplot2`

## Date and time formats

Converting text strings to dates can be tricky.

There are many ways of writing dates
 
- Saturday 5th December 2020
- 12-5-2020
- 5.12.20
- 5 Dec 20
- 2020/12/5

With or without times

- Saturday 5th December 2020 2:23 PM
- 12-5-2020 14:23:01

And in a variety of languages

- januar
- January
- janvier
- 一月

R can cope will all of these formats, but when entering data into a spreadsheet, I strongly recommend using the [international standard format](https://en.wikipedia.org/wiki/ISO_8601) for dates and times.

- Year-Month-Day Hour:Minute:Second
- 2020-12-5 12:23:13

## Time and date formats in R

It is usually more convenient to have times and dates in special format rather than as a text.
This allows them to be manipulated, sorted, and plotted properly.

The formats for dates without times is `Date`.
Internally, this is the number of day since 1970-01-01.

The format for dates with times is `POSIXct`.
Internally, this is the number of seconds since 1970-01-01.

Sometimes it is useful to have a time-only format, this is provided by the `hms` package.

## Converting dates and times

This can be done using base R.

```{r base-R-convert}
dat <- "26 October 2016 14:39:10 CEST" #date with time-zone
as.POSIXct(dat, format = "%d %B %Y %H:%M:%S") # might return NA if run on your computer - see locales below
```

Here the `%d` represents the day, `%B` represents the month written as the full word (`%b` would work if the month was the abbreviated) and `%Y` represents the 4-digit year. 
There is a complete alphabet of codes you can use. 
See `?strptime` for the complete list.

## Converting dates and times with `lubridate`

It is difficult to remember all the codes and easy to get the format wrong.
Fortunately, the `lubridate` package makes converting dates much easier.

** No more `"%d %B %y"`**

With `lubridate`, you don't need to remember the alphabet of codes and get the format **exactly** right, you just need to select the function with the day, month and year in the correct order.

```{r lubridate-convert}
library("lubridate")
dat <- "26 October 2016 14:39:10"
dmy_hms(dat)
```

`dmy_hms` expects the order to be days-months-years followed by hours:minutes:seconds. 
Provided the elements are in this order, any format can be used.

```{r lubridate-dmy_hms}
dmy_hms("26-10-16 14.39.10")
dmy_hms("26th Oct 2016 14 39 10")
dmy_hms("261016143910")#this would be a really bad way to store date-times
```

`lubridate` can even cope when the format is mixed, provided the elements are in the same order.

```{r lubridate-dmy_hms-mixed}
dmy_hms(c("26-10-16 14.39.10", "26th Oct 2016 14 39 10", "261016143910"))
```



### Different formats

There are many other functions in the `lubridate` package for coping with dates with elements in different orders.

For example

```{r lubridate-formats}
dmy_hm("26th Oct 2016 14.39") # no seconds
ydm_hms("2016 26th Oct 14 39 10") # year first then day
mdy_hms("Oct 26th 2016 14 39 10") # month first (US standard)
dmy("26-10-2016") #date only. Gives date format.
```

Look at the help file for more functions and arguments.

If you have an invalid format or an impossible date, you will get a warning.

```{r bad-date}
dmy("28-F-2020") # Invalid format - use February, Feb or 2
dmy("31-Feb-2020") # Impossible date
```



### Time zones

The default timezone for the `lubridate` functions is UTC (approximately Greenwich Mean Time).
This is normally fine unless you are dealing with local times in multiple time-zones, or because you need to allow for daylight saving time.
You can change the time zone with the argument "tz".

```{r timezone}
mdy_hm("July 1 2001 2:14", tz = "Europe/Oslo")
```

The function `OlsonNames()` returns a vector of valid time zones.
See the help file for details.

```{r}
OlsonNames()[1:10]
```

## Languages

R uses locales to know what language to expect the date to be written in. 
See `?locales`. 
Locales are a bit of a pain.

You can find your current locale with 

```{r get-locale}
Sys.getlocale(category = "LC_TIME")
```

Unfortunately, the locales available depend on your operating system.

You can change the locale to Bokmål for the whole session with 

```{r date-locales}
Sys.setlocale(category = "LC_TIME", locale = "nb_NO.utf8") #on linux/Mac, "nn_NO.utf8" for nynorsk
#Sys.setlocale(category = "LC_TIME", locale = "Norwegian Bokmål_Norway.1252") # should work on windows
dmy("1 januar 2020")
```

Or use the `locale` argument of `dmy`.

To see the available locales on your computer, run 

```{r available-locales, eval = FALSE}
system("locale -a", intern = TRUE) #on linux/Mac
```
On Windows, you need to go to the Region setting in the Control Panel.

```{r date-locale-revert, echo = FALSE, include = FALSE}
Sys.setlocale("LC_TIME", "en_DK.utf8")
```


## Date arithmetic

Dates and times can be used in calculations

```{r date-arithmetic}
ymd("2020-5-18") - 5 #base unit is days for dates
ymd("2020-5-18") - period(5, unit = "months")
ymd("2020-5-18") - ymd("2020-5-13")
```

## Extracting elements of the date

Given a date or time, we can extract different elements, for example the month, day or hour.

```{r}
month(today())
yday(today())# day of year
hour(now())
```



## Dates and `dplyr`

As an ecologist, I spend a lot of time processing climate data  for comparison with ecological data.
The data are typically in R in a data.frame, so `dplyr` is useful for manipulating them.
This section shows how to use `lubridate` together with `dplyr`.

The dataset `temperature` includes a few days of air temperature from the [GFI weather station](https://veret.gfi.uib.no/), Bergen.

The dataset has three columns, `Dato`, `Tid` and `Lufttemperatur`. 
I have imported the data with `readr::read_csv` so the first two columns have automatically been converted to date and time formats. 
You can tell this because the class of each column is given below the column name.

```{r print-temperature}
temperature
```

### Using `mutate` to make a new column

I want to make a single date-time column. 
To do this I need to combine the data and time columns with `paste`, and then use a `lubridate` function for the conversion. 

```{r date-time-temperature}
temperature2 <- temperature %>% 
  mutate(
    date_time = paste(Dato, Tid), #combine
    date_time = ymd_hms(date_time)#convert
    ) 

temperature2
```

`lubridate` functions `make_date` and `make_datetime` are useful if the year, month, day etc. are in different columns.

```{r make-date}
tibble(year = 2020, month = 10, day = 3) %>% 
  mutate(date = make_date(year = year, month = month, day = day))
```


### Using `filter` to select rows

I often need to `filter` data to remove bad data or restrict the dates and times in the dataset.

For example, to calculate the mean temperature of the afternoon of the 21st April, we first need to `filter` the data. 
`between` is a useful helper function here to avoid having to write a more complicated logical condition.

```{r filter-temperature}
temperature2 %>% 
  filter(between(date_time, 
                 left = ymd_hm("2020-04-21 12:00"),
                 right = ymd_hm("2020-04-21 17:00"))) %>% 
  summarise(mean_temp = mean(Lufttemperatur))

```


### Using `group_by` and `summarise`

I often want to summarise climate data to calculate mean monthly temperatures. 
We can do an analogous task with the `temperature` dataset to calculate mean hourly temperatures.

```{r mean-hourly}
temperature2 %>% 
  mutate(hour = hour(date_time)) %>% 
  group_by(hour) %>% 
  summarise(mean_temperature = mean(Lufttemperatur))
```


## Plotting dates

We can plot the temperature data with `ggplot`.

```{r}
ggplot(temperature2, aes(x = date_time, y = Lufttemperatur)) + 
  geom_line()
```

`ggplot` recognises that the x-axis variable is a date-time and plots the axis with appropriate labels. 
The language will depend on the locale (see `Sys.getlocale()`). 
I don't like the way the dates have been formatted so I can use `scale_x_datetime` to alter the format. 
The codes from `?strptime` that you learnt at the start of this tutorial finally become useful!

```{r}
ggplot(temperature2, aes(x = date_time, y = Lufttemperatur)) + 
  geom_line() +
  scale_x_datetime(date_labels = "%A %d %B") #see also date_breaks to change label positions
```


We can also plot times. 

```{r}
ggplot(temperature2, aes(x = Tid, y = Lufttemperatur, group = Dato)) + 
  geom_line()
```

Here the aesthetic `group = Dato` make `ggplot` draw a different line for each day.

I don't think the breaks in this plot were chosen very well so I can use `scale_x_time` to set the breaks (in seconds) and the format for the times.

```{r}
ggplot(temperature2, aes(x = Tid, y = Lufttemperatur, group = Dato)) + 
  geom_line() + 
  scale_x_time(breaks = seq(0, 24, 6) * 3600, labels = scales::time_format("%H:%M"))
```

