---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Combining a histogram and a density plot
# author: jonathan
# revised: 25/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Multiple/histogram-density.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Multiple/histogram-density.Rmd")
library(ggplot2)
ID <- 1:200
data <- rnorm(200, mean=65, sd=15)
```

Sometimes a density plot and a histogram of frequency are found combined in the same chart. After all, both represent data distribution in their own specific way. Here we will use `ggplot()` to draw this combined chart for a rather simple dataset.

Before going any further, if you are not so familiar with histograms and density plots, have a quick look at these pages:

+ [Creating a histogram of frequency](https://biostats.w.uib.no/creating-a-histogram-of-frequency/){target="_blank"},
+ [Creating a density plot](https://biostats.w.uib.no/creating-a-density-plot/){target="_blank"}.

The dataframe for this tutorial is as follows:
```{r dataframe}
# dataframe
df <- data.frame(ID, data)
# structure of the dataframe
str(df)
```
<br/><br/>

Now, let’s look at the code for the chart. Here we shall create a plot with two layers and thus two geometries. We shall use:

+ `geom_density()` for drawing the density plot, 
+ `geom_histogram()` for drawing the histogram. 

However we must add the argument `aes(y = ..density..)` in `geom_histogram()` for the histogram to take the same dimension as the density plot on the Y-axis and thus show up correctly:
```{r plot}
ggplot(df, aes(x = data)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, fill = "grey") +
  geom_density() 
```
<br/><br/>

Omitting `aes(y = ..density..)` would result in a proper histogram under an almost flat line which is the density plot on a *microscopic* scale:
```{r bad scale}
ggplot(df, aes(x = data)) +
  geom_histogram(binwidth = 5, fill = "grey") +
  geom_density() 
```
<br/><br/>

Here is the original chart with a slightly better look:
```{r scatterplot with smoother}
ggplot(df, aes(x = data)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, colour= "black", fill = "white") +
  geom_density(fill="blue", alpha = .2) 
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
