---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Friedman test
# author: jonathan
# revised: 30/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/useful-tests/friedman.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/useful-tests/friedman.Rmd")
library(ggplot2)
```

**Friedman rank sum test (or simply Friedman test)** is a non-parametric test that may be used as an alternative to [repeated measures ANOVA](https://biostats.w.uib.no/repeated-measures-anova/){target="_blank"} when, for example, you suspect or you know that the assumption of normality of distribution is not met. 

In this test, the null hypothesis `H0` states that there is no difference between the means of the groups.

The function to use in R is `friedman.test()`. It is possible to use the following two syntaxes `friedman.test(response, predictor-group, predictor-block)` and `friedman.test(response ~ predictor-group | predictor-block)`. In both cases, `response` is response variable, `predictor-group` is the predictor variable corresponding to the grouping factor and `predictor-block` is the predictor variable corresponding to the blocking factor. Note that the position of the variables in the function is important; inverting `predictor-group` and `predictor-block` will most likely result in a different test outcome.  

Let's take an example where 5 rats are weighed 4 times with intervals of 4 weeks (week8 to 20). Here is the code for the dataframe:

```{r dataframe}
# response variable
rat.weight <- c(164, 164, 158, 159, 155, 220, 230, 226, 227, 222, 261, 275, 264, 280, 272, 306, 326, 320, 330, 312)

# predictor variables
rat.ID <- as.factor(rep(c("rat1", "rat2", "rat3", "rat4", "rat5"), 4))
time.point <- as.factor(c(rep("week08", 5), rep("week12", 5), rep("week16", 5), rep("week20", 5)))

# dataframe
my.dataframe <- data.frame(rat.ID, time.point, rat.weight)
```
<br/><br/>

Let's visualize the data with a multiple line chart:
```{r plot, echo=FALSE}
ggplot(my.dataframe, aes(x = time.point, y = rat.weight, group = rat.ID, color = rat.ID)) +
  geom_line(size = 1.25) +
  geom_point(size = 2) +
  scale_color_viridis_d()
```
<br/><br/>

### Running the test

In such an experimental design where subjects (rats) are being tested in a time series, the time points are the groups and the subjects are the blocks. We thus write:
```{r friedman}
friedman.test(rat.weight, time.point, rat.ID)
```

The null hypothesis `H0` stating that there is no difference between the means of the groups is rejected due to the low p-value.

As mentioned earlier, there is an alternative syntax for the function. Here is the code:
```{r friedman2}
friedman.test(rat.weight ~ time.point | rat.ID)
```
As you see, the result is identical.



<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
