---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Test for sphericity – Mauchly’s test
# author: jonathan
# revised: 29/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/useful-tests/mauchly.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/useful-tests/mauchly.Rmd")
```
Statistical tests such as [repeated measures ANOVA](https://biostats.w.uib.no/repeated-measures-anova/){target="_blank"} assume **sphericity**, meaning that **the variances of the differences between all possible pairs of groups are equal**. Whereas assumption of normality of distribution may be overlooked in some cases due to the robustness of ANOVA, the assumption of sphericity in repeated measures ANOVA *must* be met. Violating this assumption may severely compromise the interpretation of the test results.

**Mauchly’s sphericity test** checks this assumption for you. The null hypothesis `H0` in this test states that there is no difference in variances for all pairwise group comparisons. In R, the main function is `mauchly.test()`. However, running it is not that straight forward. It involves building a matrix with the dataset to be tested, whereas your data is usually nicely arranged in a dataframe.

Let's see how this is done using the same example as for [repeated measures ANOVA](https://biostats.w.uib.no/repeated-measures-anova/){target="_blank"}. Here, 5 rats are weighed 4 times with intervals of 4 weeks (week8 to 20). This is the code for the dataframe:

```{r dataframe}
# response variable
rat.weight <- c(164, 164, 158, 159, 155, 220, 230, 226, 227, 222, 261, 275, 264, 280, 272, 306, 326, 320, 330, 312)

# predictor variable
time.point <- as.factor(c(rep("week08", 5), rep("week12", 5), rep("week16", 5), rep("week20", 5)))

# individual ID
rat.ID <- as.factor(rep(c("rat1", "rat2", "rat3", "rat4", "rat5"), 4))

# dataframe
my.dataframe <- data.frame(rat.ID, time.point, rat.weight)
```
<br/><br/>

We start by creating a matrix called `neo` that contains the data `rat.weight`. In that matrix, the 5 individuals will be shown in rows, and time points in column. Here is the code:

```{r matrix}
neo <- matrix(rat.weight, nrow = 5, ncol = 4)
neo
```
<br/><br/>

Now that the matrix is in place, we can used `mauchly.test()` (in combination with `lm()`) to test for sphericity:
```{r mauchly}
mauchly.test(lm(neo ~ 1), X = ~1)
```

The output returns a p-value over 0.05 meaning that the null hypothesis stating that there is no difference in variances for all pairwise group comparisons is accepted. 

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
