---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Installing ggplot2
# author: jonathan
# revised: 11/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Starting/installing-ggplot2.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Starting/installing-ggplot2.Rmd")
library(ggplot2)
```

Like any other package in R, if you have not installed/used it before on your machine, you will need to install it using the `install.packages()` command.
 
Simply type the following code in your console:
```{r install, eval=FALSE}
install.packages("ggplot2")
```
<br/><br/>
Then use the following line to activate ggplot2 everytime you restart R:
```{r load, eval=FALSE}
library(ggplot2)
```
<br/><br/>
Always remember that you will find plenty of help in the documentation of the package. To access it, type:
```{r doc, eval=FALSE}
?ggplot2
```
A window with links and references will pop up in Rstudio.


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
