---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Key components of ggplot()
# author: jonathan
# revised: 14/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Starting/key-components.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Starting/key-components.Rmd")
library(ggplot2)
```

Any plot designed with the function `ggplot()` has 3 necessary components:

+ the data to display, which must be contained in a **dataframe**,
+ **æsthetics**, introduced by the function `aes()` that takes care of positioning/mapping the variables on the plot,
+ **geometry**, introduced by the function `geom()` that takes care of drawing the graph (bar graph, line, scatter plot, etc).

Simplified a lot, these 3 components can be seen as *layers* which, added on top of each other, will form the resulting plot. The code for the “simplest plot ever” has the following syntax:
```{r plot, eval=FALSE}
ggplot(data = dataframe, aes(x = variable1, y = variable2)) +
  geom()
```
Note that both the dataframe and `aes()` are inside `ggplot()` while `geom()` is added to `ggplot()`.



### Example

Let's take a simple example where we draw a *scatter plot* based on the variables `var1` and `var2` which are contained in the dataframe `df`. Here is the code for the variables and dataframe:

```{r variables}
# variable1
var1 <- c(1,2,3,4,5,6,7,8,9,10)

# variable2
var2 <- c(45,12,48,79,65,32,78,95,12,75)

# dataframe
df <- data.frame(var1, var2)
```

The code for the scatter plot is:
```{r scatter}
ggplot(data = df, aes(x = var1, y = var2)) +   # maps var1 and var2 contained in df
  geom_point()                                 # draws the dots of the scatter plot
```

### Simplifying the code

It is possible to write the code in a simpler manner by removing `data =`, `x =` and `y =`. However, make sure that there is no ambiguity about the order of the variables written in `aes()` since the first variable will be defined as `x` by convention. Any inversion will result *at best* in an inverted plot.

The code for the scatter plot in the example above may be simplified the following way:
```{r scatter2, eval=FALSE}
ggplot(df, aes(var1, var2)) +   
  geom_point()                             
```

However, inverting the variables results in something rather different:
```{r scatter3}
ggplot(df, aes(var2, var1)) +   
  geom_point()                             
```
<br/><br/>

### Changing your mind about which plot type to use?

Let’s now assume that you change your mind and want to draw a bar plot instead of a scatter plot. The only thing you need to change in the code is the layer that is responsible for drawing, aka the geometry. You shall thus replace `geom_point()` with a more appropriate function such as `geom_col()`. The code becomes:
```{r bar}
ggplot(data = df, aes(x = var1, y = var2)) +   
  geom_col()                                   # draws bars instead of dots
```
<br/><br/>

The relative simplicity in the syntax and structure of `ggplot()` will allow you to not only rapidly modify layers, but also add other layers, thus creating [multiple plots](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/overlay-plots-multiple-plots/){target="_blank"}, [faceted plots](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/faceted-plots/){target="_blank"}, etc. 
<br/><br/>

### What happens if you omit one of the three layers?

#### Omitting the geometry

Here we omit the function `geom_point()`. `ggplot()` is left with the dataframe and the æsthetics, but it does not know what to do with it. As a result, it shows you the axes, scales, grid and background, but nothing more:

```{r omit1}
ggplot(data = df, aes(x = var1, y = var2))
```
<br/><br/>

#### Omitting the dataframe

Here we omit the dataframe `df`. Doing so, `ggplot()` knows that you want to draw a scatter plot, that the data is in two variables called `var1` and `var2`, but it does not find the data frame that contains them. It just ends with an error message:
```{r omit2, echo=TRUE, error=TRUE}
ggplot(aes(x = var1, y = var2)) +
  geom_point()
```
<br/><br/>

#### Omitting the æsthetics

Here we omit the function `aes()` and its content. Doing so, `ggplot()` knows that the data is in the dataframe `df`, that you want to draw a scatter plot, but it cannot map the data. It ends again with an error message:
```{r omit3, echo=TRUE, error=TRUE, eval=TRUE}
ggplot(data = df) +
  geom_point()
```

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
