---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: comparing two proportions - which test
# author: jonathan
# revised: 10/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/2-sample/comparing2proportions.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/2-sample/comparing2proportions.Rmd")
```

Comparing means is not the only way to compare two samples. Here is an example where **proportions** or **rates** may be used to compare groups. Let'ssee that with a quick example:

> You run an experiment where mice dispatched in 2 groups (a control group and an experimental group) are tested for their capacity to escape from a maze in a given amount of time. You carefully note the number of successes for each group and end up with the following results: 5 mice (out of 20) from group A made it in time whereas 9 mice (out of 27) from group B made it in time.

The “success rates” in group A and B being 25% and 33% respectively, it looks like the mice in group B performed better in this test. But did this result happen by chance? What if we do this test again and again…? How likely is it that group B performs better, again?

The function `prop.test(x, y)` runs a test for equality of proportions. It compares the proportions or probabilities of success, the null hypothesis `H0` stating that these proportions are equal. `x` is a vector that contains the number of success for each group and `y` is a vector that contains the total numbers of individuals in each group).

Let’s run this test on our example.

```{r test}
x <- c(5, 9)
y <- c(20, 27)
prop.test(x, y)
```

Looking at the test output, you can see that the obtained p-value is quite high. The null hypothesis is thus accepted. The apparent difference in success rate between group A and group B is likely to have occurred only by chance. Mice in group B are not faster or more clever than in group A, based on this test.


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
