---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: comparing two means - Student's paired t-test
# author: jonathan
# revised: 13/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/2-sample/student-paired.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/2-sample/student-paired.Rmd")
date <- c(5:15)
GFI <- c(8.2, 6.0, 6.1, 4.9, 4.8, 6.7, 7.5, 8.0, 7.2, 5.7, 3.5)
Yr <- c(8.6, 6.6, 6.6, 5.5, 5.4, 6.8, 7.9, 8.4, 7.3, 6.0, 4.2)
data.source <- rep(c("GFI", "Yr"), each = 11)
temperature <- c(GFI, Yr)
df <- data.frame(date, data.source, temperature)
library(ggplot2)
```
Student’s paired t-test may be used to compare the means of two dependent samples. As for [Student’s t-test for independent samples](http://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"}, the assumptions are that the samples are normally distributed, and have equal variances.

The function in R is `t.test(x, y, paired = TRUE)` where `x` and `y` are the vectors containing the samples and `paired = TRUE` the argument indicating that samples are dependent (note that `paired = FALSE` is the argument by default in `t.test()`, which explains why one does not need to (but can) write `paired = FALSE` whenever running the test on independent samples).

Let’s take an example. We have access to weather records for the period April 5-15th, 2016 at Florida, close to Bergen City Centre. Daily mean temperatures were recorded by two stations relatively close to each other. The data from the first station is provided by [GFI](https://veret.gfi.uib.no/){target="_blank"} and data from the second station is provided by [Yr](https://www.yr.no/){target="_blank"}. Our whole data set thus consists of *pairs of measurements* for the same location and the same period. We may thus run Student’s paired t-test, but we still need to check for normal distribution ([Shapiro-Wilk test](http://biostats.w.uib.no/test-for-normality-shapiro-wilks-test/){target="_blank"}) and equal variances ([Fisher’s F test](http://biostats.w.uib.no/1-comparing-two-variances/){target="_blank"}):
```{r normality-variance, echo=FALSE}
shapiro.test(GFI)
shapiro.test(Yr)
var.test(GFI, Yr)
```
All p-values are greater than 0.05; the samples thus have equal variances and originate from a normal distribution.

Let's visualize the data with a boxplot:
```{r boxplot, echo=FALSE}
ggplot(df, aes(data.source, temperature)) +
  geom_boxplot()
```
<br /><br />
Here, the medians are relatively close to each other but not equal, and the spread is rather similar. 
Let's look at the data with a line plot:
```{r lineplot, echo=FALSE}
ggplot(df, aes(date, temperature, color = data.source)) +
  geom_line(size = 1.25) +
  geom_point(size = 2.5)
```
<br /><br />
It seems like there is a rather small, but constant gap between the lines. Is that enough to declare that the 2 samples are significantly different? 

Let's test it with Student's paired t-test:

```{r student, warning=FALSE}
t.test(GFI, Yr, paired = TRUE)
```
So, the test results in a p-value *less* than 0.05, thus indicating that the null hypothesis (the sample means are not different) may be rejected. 

Note: If you wonder whether the argument `paired = TRUE` makes a difference or not, let's try the test without it:
```{r student2, warning=FALSE}
t.test(GFI, Yr)
```
This time, the p-value is *greater* than 0.05 and `H0` cannot be rejected anymore... So it is clearly important to know about the dependency of your samples; the conclusion of the test relies on that.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
