---
title: "Descriptive Statistics"
output:
  bookdown::html_document2:
    highlight: tango
    toc: true
    toc_float: true
    css: css/style-chapters.css
---

```{r setup, eval=TRUE, include=FALSE}
# add all packages that need loading
library(tidyverse)
library(knitr)
# source figure settings
source("../Templates/biostats_theme.R")
# set warning to be off for all chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

### Before you start {.facta .toc-ignore}

You should be familiar with basic statistical theory and the basics of R.

</br>



## Introduction

Descriptive statistics summarize or characterize a given sample and are the foundation of statistical tests and data visualizations.
In this section, we will define a number of descriptive statistics and learn how to calculate them in R. 

</br>



### 1. Mean

The *mean* (or average) is the most commonly used measure of central tendency of a sample. 
A measure of central tendency describes a typical value within a sample. 
The mean is often called the average and constitutes one of the measures of the central tendency of your sample.
The mean is the sum of all the values in your sample divided by the number of values in the sample set. 
Mathematically described as: 

$$\overline{x} = \frac{\sum_{i=1}^n x_i}{n}$$ 

Where $\overline{x}$ is the mean of the sample equal to the sum ($\sum$) of all observations, _i_, divided by the total number of observations, _n_. 

In R, the function that calculates the mean is called `mean()`:

```{r mean, echo=TRUE}
x <- c(12, 6, 3) # create a sample vector
mean(x) # take the mean of the sample vector
```

</br>



### 2. Median

The *median* is another measure of the central tendency of your sample. 
The median is the number “in the middle” of the sample, or, the middle value of a series of values sorted from smallest to largest.

Provided that the sample contains an odd number _n_ of entries, the median is the value of the $(\frac{(n+1)}{2})$ entry in the sample. 
If the sample contains an even number _n_ of entries, then the median is the mean of the values of the $(\frac{n}{2})$ and the $(\frac{n}{2} + 1)$ entries.

In R, the function that calculates the median is called `median()`:

```{r median odd, echo=TRUE}
x <- c(12, 6, 3) # odd number in sample vector
median(x) # takes the median
```

```{r median even, echo=TRUE}
y <- c(12, 6, 3, 1) # even number in sample vector
median(y) # takes the median
```

</br>



### 3. Minimum, maximum, range, and midpoint

The *minimum* and *maximum* of a sample are the smallest and largest entries, respectively. 

In R, the functions that calculate the minimum and maximum  are called `min()` and `max()`, respectively:

```{r min and max, echo = TRUE}
x <- c(12, 6, 3, -1) # sample vector
min(x) # find the minimum
max(x) # find the maximum
```

The *range* defines the spread of the data and is the difference between the minimum and maximum values.
Note, a range may be expressed as a single value (the actual difference between maximum and minimum) or by writing minimum-maximum. 
For instance, the range of the series (1, 2, 3, 4, 5, 6, 7, 8, 9, 10) can be expressed as 9 or 1-10.

In R, the function that calculates the range is `range()`. 
However, `range()` returns a vector that contains the minimum and the maximum, respectively.
To calculate the difference between these values, you may use `diff(range())` or `max()-min()`:

```{r range, echo = TRUE}
x <- c(12, 6, 3, -1) # sample vector
range(x) # find the range of the sample as a vector
diff(range(x)) # find the range as difference between minimum and maximum values with diff()
max(x)-min(x) # find the range as difference between minimum and maximum values
```

The *midpoint* of a sample is the value equidistant from the minumum and the maximum. 
It can be calculated by taking taking the mean of the minumum and maximum value of a sample.

In R, you can calculate the midpoint of a sample with: 

```{r midpoint, echo = TRUE}
x <- c(12, 6, 3, -1) # sample vector
(max(x) + min(x))/2
```

</br>



### 4. Variance and standard deviation

The *variance* (var or $\sigma^2$) corresponds to the mean of the squared differences from the mean. 
Mathematically, variance ($\sigma^2$) of the data (x) is: 

$$\sigma^2 (x) = \frac{\sum_{i=1}^n (x_i - \overline{x})^2}{n}$$

Where we take an entry in the sample, $x_i$, and calculate the difference between its value and the mean, $\overline{x}$, then square it. 
Do the same for all the entries, _n_, in the sample. 
Finally, calculate the mean of these squared differences (divide by _n_) and you get the variance. 
Note-variance is always positive because of the squared difference.

In R, the function that calculates the variance is `var()`:

```{r variance, echo=TRUE}
x <- c(12, 6, 3, -1) # sample vector
var(x) # finds the variance of the sample
```

The standard deviation (sd or $\sigma$) is the square root of the variance. 
Unlike the variance, the standard deviation has the same unit as the mean, the median or any entry in the sample. 
This makes it very useful as it can be used in combination with the mean to define how values are spread around the mean. 
Often, you will find this expressed as mean +/- sd.

In R, the standard deviation is obtained via the function `sd()`. 
Note that you will obtain the same result by calculating “manually” the square root of the variance with `sqrt(var())`:

```{r standard deviation}
x <- c(12, 6, 3, -1) # sample vector
sd(x) # finds the standard deviation of the sample
sqrt(var(x)) # alternative method to find the standard deviation
```

</br>



### 5. Quartiles

Take a sample, sort the data from smallest to largest and split it in four equal subsets.
The quartiles are the values of the sample that cut it off in four.

Quartiles are called:
Q1: the first quartile under which the first 25% of the data in the set can be found,
Q2: the second quartile under which the first 50% of the data in the set can be found (Q2 is also the median),
Q3: the third quartile under which the first 75% of the data in the set can be found.

In addition, one refers to:
Q0: the minimum value in the set,
Q4: the maximum value in the set.

Note that the set of data between Q1 and Q3 (which contains the middle 50% of the data) is the interquartile range (IQR).

In R, quartiles may be obtained using the function `quantile()` (NB: this is not a typo, quantile refers to the intervals (0, 25, 50, 75, 100%) at which the quartiles, the values in the data, are assessed). 
If you choose to use only `quantile()` with no other argument than the vector containing the sample, R returns Q0, Q1, Q2, Q3, and Q4. 
If you need only Q1, Q2 and Q3, you use an extra argument (see the following example):

```{r quartiles, echo = TRUE}
x <- c( 21 ,15, 12, 6, 3, -1, -4, -9) # sample vector
quantile(x) # Find the Q0, Q1, Q2, Q3, and Q4
quantile(x, probs = c(0.25, 0.5, 0.75)) # Find the Q1, Q2, and Q3 by defining which quartiles (in proportions) using the probs argument
```

Quartiles can be visualized with `ggplot()` and `geom_boxplot()`:

```{r quartiles boxplot, echo=TRUE}
plot_vector <- data.frame(x) # make the vector a data frame for ggplot
ggplot(plot_vector, aes(y=x))+geom_boxplot() # assign the vector "x" as the y variable in the 'aes' argument and pass it to 'geom_boxplot' 
```

</br>



### Further reading {.facta .toc-ignore}


### Contributors {- .contributors .toc-ignore}

```{r contributors, eval=FALSE}
* Joshua Lynn
* Kwaku P. Adjei
```

