```{r setup, include=FALSE}
library(palmerpenguins)
library(tidyverse)
source("../Templates/biostats_theme.R")
```

# Cross-referencing {#sec-cross-referencing}

Cross-referencing lets you refer to figures and tables in your document. Once set up, this happens automatically, so you don't need to re-number all the figures when you add or delete one.

## Cross-referencing figures and tables

The block that made the figure or table needs to have a label that starts with `fig-` or `tbl-`, respectively. The figure or table also need a caption.

```{r}
#| echo: fenced
#| label: fig-penguin-dens
#| fig-cap: Histogram of penguin body mass
#| warning: false

penguins |>
  ggplot(aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.7) +
  scale_fill_brewer(palette = "Set1")
```

To cross-reference the figure made by the above chunk, we need to use this notation.

``` markdown
Penguins are cool (@fig-penguin-dens)
```

[Penguins are cool (@fig-penguin-dens)]{style="background-color: Beige;"}

If you type an <kbd>\@</kbd> in the visual editor it will make suggestion.

To refer to a table made by a block labelled `tbl-penguin-results`, you would use

``` markdown
Penguins are cool (@tbl-penguin-results))
```

If you get the chunk name wrong, there will be a warning when the document in knitted, and question marks shown where the reference should be.

``` markdown
Figure @fig-no-exist) does not exist
```

[Non-existing figure (@fig-no-exist))]{style="background-color: Beige;"}

## Cross-referencing equations

To cross-reference an equation, we need to give it a label.

``` markdown
$$
\frac{\sum_{i=1}^{n}{x_i}}{n}
$${#eq-mean}
```

Now the label can be used.

``` markdown
The mean (equation @eq-mean).
```

$$
\frac{\sum_{i=1}^{n}{x_i}}{n}
$$ {#eq-mean}

[The mean (equation @eq-mean).]{style="background-color: Beige;"}

## Cross-referencing document sections {#sec-cross-referencing-document-sections}

You can also make cross-references to other sections of the document. First, you need to set up a identifyier for the section you want to link to. In the visual editor, this can be done by clicking on the three dots to the right of the header. The ID should start with `#sec-` and be in lower case. Now you can use this ID to make a link by typing <kbd>\@</kbd>.

``` markdown
## Cross-referencing document sections {#sec-cross-referencing-document-sections}

Cross-referencing (@sec-cross-referencing-document-sections)
```

[Cross-referencing (@sec-cross-referencing-document-sections)]{style="background-color: Beige;"}

Note that the reference must be written in lower case and spaces must be replaced by hyphens. Punctuation is deleted.
