```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Run a one-sample t-test
# author: jonathan
# revised: 9/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/1-sample/ttest.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/1-sample/ttest.Rmd")
scores <- c(123, 135, 142, 127, 117, 134, 123, 133, 111, 130, 113, 118, 105, 126, 125, 122, 147, 106, 117, 129, 144, 112, 126, 140, 129, 115, 118, 129, 125, 137, 135, 138, 134, 124, 128, 135, 124, 137, 121, 120)
```

With regards to **t-test**, the function `t.test()` in R may be used. This is a rather simple function which performs both one- and two-sample t-tests (it is thus likely that we will meet that function [elsewhere](https://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"} in this website).

Assuming that you have stored your sample data in the variable called `scores`, the command to use is `t.test(scores, alternative="ALT", mu = Y)` where:

+ `ALT` shall be replaced by either `greater` or `less` or `two.sided` depending of your alternative hypothesis `Ha`. The null hypothesis `H0` states that the sample mean is NOT different from the population mean. Your alternative hypothesis `Ha` is one of the following:
  - the sample mean is **greater** than the population mean (in that case, use `greater`),
  - the sample mean is **less than** the population mean (in that case, use `less`),
  - the sample mean is **either smaller or greater** than the population mean (in that case, use `two.sided`).
  
+ `Y` shall be replaced by the value of the population mean.

Using our previous example, this looks like:
```{r ttest}
t.test(scores, alternative = "greater", mu = 120)
```
 
R returns several lines of text. One of them provides a **p-value** while the next line states the **alternative hypothesis** which depends on the parameter `alternative` that you have entered in the `t.test()`. This alternative hypothesis `Ha` is considered valid when the p-value is less than 0.05.



 

Read more about `t.test()` and find more options by clicking [here](http://www.stat.columbia.edu/~martin/W2024/R2.pdf){target="_blank"} or [there](http://www.statmethods.net/stats/ttest.html){target="_blank"} or by simply typing `?t.test` in the R console.
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
