---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a 2D histogram
# author: jonathan
# revised: 17/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/2D-histogram.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/2D-histogram.Rmd")
library(ggplot2)
# variable 1
values1 <- rnorm(4000, mean=50, sd=10)
# variable 2
values2 <- rnorm(4000, mean=5, sd=1)
```
When a scatter plot becomes so dense that the symbols overlap massively, the plot lacks clarity and loses its purpose. This is called overplotting. In such a case, there are a few alternatives, one of them being the 2D histogram. While a regular one-dimensional histogram uses bars to display the count of observations by intervals (or bins) on the X-axis, the 2D histogram displays a map of squares made by the intersection of intervals (or bins) on both the X- and Y-axis. Each square is thus defined by an interval on the X- and Y-axis, and a color (or shade) which describes the local count. Note that there exists a variant of this plot called [Hexbin plot](https://biostats.w.uib.no/creating-a-hexbin-plot/){target="_blank"}.

Let’s take the following example where the variable to be plotted along the X-axis (`values1`) and the variable to be plotted along the Y-axis (`values2`) both contain 4000 points (!). Here is the dataframe:
```{r dataframe}
# dataframe
df <- data.frame(values1, values2)
str(df)
```
<br/><br/>

If we attempt to draw a regular scatter plot with `geom_point()`, we obtain this unreadable figure:
```{r scatterplot}
ggplot(df, aes(values1, values2)) +
  geom_point()
```
<br/><br/>

In such a case of overplotting, it is logical to switch to a different plot type.
To draw a 2D histogram, we will use `geom_bin2d()`:
```{r 2D}
ggplot(df, aes(values1, values2)) +
  geom_bin2d()  
```

A color bar guide named `count` automatically appears to the right, showing the different shades of blue (color by default) used in the figure to help you define approximately the count for each of the squares.
<br/><br/>

As for the regular histogram, there is a possibility to change the value of the bins, either by defining the binwidth with `binwidth=` or by defining the number of bins with `bins=`:
```{r bins}
ggplot(df, aes(values1, values2)) +
  geom_bin2d(bins = 70)  
```
The size of the squares changes and the scale of the color bar guide is redefined accordingly.
<br/><br/>

If you feel like changing the colors used in the color bar guide, you can use `scale_fill_gradient()` or `scale_fill_viridis_c()` for example:
```{r colors}
ggplot(df, aes(values1, values2)) +
  geom_bin2d(bins = 70) + 
  scale_fill_viridis_c()
```
Read [this page](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"} to learn more about *color palettes*.
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
