---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: Creating a bubble plot
# author: jonathan
# revised: 17/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/bubble-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/bubble-plot.Rmd")
library(ggplot2)
# variable 1
values1 <- c(rnorm(100, mean = 50, sd = 15), rnorm(100, mean = 75, sd = 15), rnorm(100, mean = 100, sd = 15))
# variable 2
values2 <- c(rnorm(100, mean = 20, sd = 10), rnorm(100, mean = 30, sd = 10), rnorm(100, mean = 40, sd = 10))
# variable 3
values3 <- runif(300, min = 1, max = 6)
```
A regular scatter plot represents with dots (or any other symbols) the relationship between two given variables. One can plot a third variable to any scatter plot by the mean of a gradient of colors applied onto the dots and thus obtain a [color-graded scatter plot](https://biostats.w.uib.no/creating-a-color-graded-scatter-plot/){target="_blank"}. But one can also plot this third variable by the mean of the size of the dots or circles representing the data. The results looks thus like bubbles in a chart.

We will see how to use `ggplot()` to code for such a bubble plot. Here is the dataframe with the three variables:
```{r dataframe}
# dataframe
df <- data.frame(values1, values2, values3)
str(df)
```
<br/><br/>

We use `geom_point()` as for any scatter plot, but we need to add `aes(size=)` to represent the values of the variable `values3` as a set of bubbles of varying size:
```{r plot}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3))
```
<br/><br/>

With the argument `alpha=`, we can add transparency to the bubbles:
```{r bubbles}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3), alpha = .3)
```
<br/><br/>

And as always, it is possible to add a bit of color to the plot:
```{r color}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3), color = "blue", alpha = .3)
```
<br/><br/>

But it is also possible to color-grade the bubbles as a function of the variable `values3` with `color=` (in the same way as with `size=`). For aesthetic reasons, we will use the palette viridis brought in by `scale_color_viridis_c()`:
```{r colors}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3, color = values3), alpha = .5) +
  scale_color_viridis_c()
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
