---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: Creating a line plot
# author: jonathan
# revised: 18/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/line-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/line-plot.Rmd")
library(ggplot2)
library(gridExtra)
```
A line plot represents a data set in the form of straight lines joining the data points or markers (which are themselves often represented  as dots, circles, squares, etc.).

Let's use `ggplot()` to draw the line plot for a simple dataset representing the solar irradiance (registered every 30 minutes during 24 hours) in Østerbø on July 22nd, 2017. Here are the variables and dataframe:
```{r dataframe}
# ID
timepoint.ID <- 1:48
# variable1
solar.rad <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 10, 22, 36, 50, 68, 58, 44, 96, 121, 190, 250, 280, 325, 349, 357, 389, 322, 328, 274, 232, 192, 108, 184, 214, 301, 350, 328, 259, 258, 213, 246, 148, 61, 36, 24, 9, 3, 0)
# dataframe
df <- data.frame(timepoint.ID, solar.rad)
```
<br/><br/>

Now, let's map the data from the variable `solar.rad` by typing `ggplot(df, aes(timepoint.ID, solar.rad))` and use `geom_line()` to draw the line:
```{r plot}
ggplot(df, aes(timepoint.ID, solar.rad)) +
  geom_line()
```

We can improve the look of this plot by adding a few arguments to set a specific color and thickness of this line. It is also possible to define what type of line (dotted, solid, dashed, twodash, longdash, dotdash or blank) will be used:
```{r plot2}
ggplot(df, aes(timepoint.ID, solar.rad)) +
  geom_line(colour = "blue", size = 1, linetype = "dotted")
```
<br/><br/>

We can also add a layer that will display the data points as large red dots by the mean of `geom_point()` and the necessary arguments:
```{r plot3}
ggplot(df, aes(timepoint.ID, solar.rad)) +
  geom_line(colour = "blue") +
  geom_point(size = 3, colour = "red", shape = 16)
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
