---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Post Hoc Tests - multiple comparisons in linear mixed effect models with glht()
# author: jonathan
# revised: 11/4/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/post-hoc tests/post-hoc-tests-lme.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/post-hoc tests/post-hoc-tests-lme.Rmd")
library(multcomp)
library(ggplot2)
library(nlme)
```

You have been running a [repeated measures ANOVA](https://biostats.w.uib.no/repeated-measures-anova/){target="_blank"} with `lme()` and `anova()`, and the F-test has revealed the existence of a significant difference between some of the tested groups. But which groups? To answer that question, you will need to run the appropriate *post-hoc* tests to assess the significance of differences between pairs of group means. The function `glht()` in the package `multcomp` will help you do this.

We will reuse the example introduced [here (repeated measures ANOVA)](https://biostats.w.uib.no/repeated-measures-anova/){target="_blank"}.
To create the corresponding dataframe, use the following code:
```{r dataframe}
# response variable
rat.weight <- c(164,164,158,159,155,220,230,226,227,222,261,275,264,280,272,306,326,320,330,312)

# predictor variable
time.point <- as.factor(c(rep("week08",5), rep("week12",5), rep("week16",5), rep("week20",5)))

# individual ID
rat.ID <- as.factor(rep(c("rat1","rat2","rat3", "rat4", "rat5"),4))

# dataframe
my.dataframe <- data.frame(rat.ID,time.point,rat.weight)
```
<br/><br/>

For recall, here is the corresponding plot:
```{r plot, echo=FALSE}
ggplot(my.dataframe, aes(x = time.point, y = rat.weight, group = rat.ID, color = rat.ID)) +
  geom_line(size = 1.25) +
  geom_point(size = 2) +
  scale_color_viridis_d()
```
<br/><br/>

### Running the test with `glht()`

First we need to activate the package `multcomp`:
```{r package, eval=FALSE}
library(multcomp)
```
<br/>

Running the test with `glht()` involves using two other functions:

+ `lme()` which fits a linear mixed effect model to your data, and which syntax is `lme(response ~ predictor, random=~1|subjects, data = my.dataframe)`
+ `summary()` which computes and displays the result table.

<br/>
First, we need to fit the linear model with `lme()`, which you have certainly already done. Then we run the model in `glht()`. The syntax is `glht(model, linfct=mcp(predictor='Tukey'))` where `predictor` is the predictor variable which levels have to be compared pairwise.
Let's apply this to our example:
```{r glht}
# fit the model
model <- lme(rat.weight ~ time.point, random=~1|rat.ID, data=my.dataframe)

# run glht()
post.hoc <- glht(model, linfct=mcp(time.point='Tukey'))

# get the result table
summary(post.hoc)
```
The output shows all six pairwise comparisons, p-values and corresponding stars, and all pairwise comparisons are accompanied by a p-value less than 0.05. 

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
