---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: TukeyHSD
# author: jonathan
# revised: 11/4/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/post-hoc tests/tukeyHSD.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/post-hoc tests/tukeyHSD.Rmd")
library(multcomp)
```

**Tukey’s Honest Significant Difference (HSD)** test is a *post hoc* test commonly used to assess the significance of differences between pairs of group means. Tukey HSD is often a follow up to [one-way ANOVA](https://biostats.w.uib.no/one-way-anova/){target="_blank"}, [factorial ANOVA](https://biostats.w.uib.no/factorial-design/){target="_blank"}, when the F-test has revealed the existence of a significant difference between some of the tested groups.

The null hypothesis `H0` states that the means of the tested groups are equal. 

Here we will described 2 alternatives for running a Tukey HSD test. The first one involves the function `TukeyHSD()` which does not require any additional package installation, and the second one involves the function `glht()` which is part of the package `multcomp`. 

We will reuse the example introduced [here (one-way ANOVA)](https://biostats.w.uib.no/one-way-anova/){target="_blank"}.
To create the corresponding dataframe, use the following code:
```{r dataframe}
# response variable
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22)

# predictor variable
location <- as.factor(c(rep("ForestA",10), rep("ForestB",10), rep("ForestC",10)))

# dataframe
my.dataframe <- data.frame(size,location)
```
<br/><br/>

### Assumptions

The assumptions are:

+ the observations are **independent**,
+ **normality** of distribution,
+ **homogeneity of variance**.

NB: these are indeed the same assumptions as for one-way ANOVA or factorial ANOVA; in other words, if you were allowed (or if you allowed yourself) to conduct an ANOVA test, then it is ok to run Tukey’s test.

<br/><br/>

### Running the test with `TukeyHSD()`

The syntax is `TukeyHSD(aov(response ~ predictor), conf.level)` where `response` is the response variable, `predictor` is the predictor variable, and `conf.level` is the confidence level that you want to define (usually fixed at 0.95).
Let's apply it to our example:
```{r TukeyHSD}
TukeyHSD(aov(size ~ location), conf.level=0.95)
```

The output displays the results of all pairwise comparisons among the tested groups (here three groups: `ForestA`, `ForestB` and `ForestC`, thus 3 comparisons).  You’ll find the actual difference between the means under `diff` and the adjusted p-value (`p adj`) for each pairwise comparison. Looking at the last column, the only significant difference to be reported in the present test is between the means of the groups `ForestB` and `ForestC`.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
