---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: Creating a multiple line plot
# author: jonathan
# revised: 28/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Multiple/multiple-lines.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Multiple/multiple-lines.Rmd")
library(ggplot2)
time <- rep(0:23, 2)
location <- c(rep("Østerbø", 24), rep("Lygra", 24))
solarrad <- c(
  0, 0, 0, 0, 0, 0, 14, 53, 183, 308, 323, 558, 509, 609, 693, 685, 702, 651, 514, 393, 253, 104, 33, 1,
  0, 0, 0, 0, 0, 12, 43, 82, 260, 396, 506, 624, 727, 780, 519, 544, 284, 229, 557, 425, 148, 38, 23, 0
)
```

A (multiple) line plot represents two or more data series from a data set in the form of straight lines joining the points or markers (which are themselves represented often as dots, circles, squares, etc.). 

Before going any further, if you are not so familiar with line plots, have a quick look at this page:

+ [Creating a line plot](https://biostats.w.uib.no/creating-a-line-plot/){target="_blank"}.

Let’s use `ggplot()` to draw the line plot for a dataset representing the solar irradiance (response variable `solarrad`) registered every hour during 24 hours (predictor variable `time`) at two locations: Østerbø, Sogn og Fjordane and Lygra, Hordaland (predictor variable `location`) on July 23rd, 2017.

The dataframe for this tutorial is as follows:
```{r dataframe}
# dataframe
df <- data.frame(time, location, solarrad)
# structure of the dataframe
str(df)
```
<br/><br/>

The function to use for drawing the lines is `geom_line()`. We must map the variables with the function `aes()` which will look more or less like this: `aes(time, solarrad, location)`. Our plan is to:

+ plot the predictor variable `time` on the X-axis, 
+ plot the response variable `solarrad` on the Y-axis,
+ use `color=` to split the data for each level of `location`,
+ draw the lines with `geom_line()`.

The code for the plot is as follows:

```{r plot}
ggplot(df, aes(x = time, y = solarrad, color = location)) +
  geom_line()
```
Here the data for each location is represented by a colored line.

Alternatively, it is possible to represent the data for each location in the form of a dotted line, a plain line, a dashed line, etc, instead of a specific color. To do so, we replace `color=` with `linetype=`:

```{r plot2}
ggplot(df, aes(x = time, y = solarrad, linetype = location)) +
  geom_line()
```
<br/><br/>

And it is possible to combine bith color and linetype to represent the locations. Here we simply cumulate `linetype = location` and `color = location` in `aes()`:
```{r plot3}
ggplot(df, aes(x = time, y = solarrad, color = location, linetype = location)) +
  geom_line()
```
<br/><br/>

As usual, you may tune the thickness of the lines with `size=` to make them more visible:
```{r plot4}
ggplot(df, aes(x = time, y = solarrad, color = location, linetype = location)) +
  geom_line(size = 1.3)
```
<br/><br/>

It is possible to add shapes to the lines to mark the data points. To do so, we must add a layer with the function `geom_point()`. Nothing else is needed as all the necessary info (mapping, grouping) is already mentioned in `aes()`:

```{r plot5}
ggplot(df, aes(x = time, y = solarrad, color = location, linetype = location)) +
  geom_line(size = 1) +
  geom_point(size = 2)
```
<br/><br/>

And if you want to use various shapes (triangle, square, circles, etc) for the markers, add `shape = location` to `aes()`:

```{r plot6}
ggplot(df, aes(x = time, y = solarrad, color = location, shape = location)) +
  geom_line(size = 1) +
  geom_point(size = 2)
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
