---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: facet_grid
# author: jonathan
# revised: 2/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Faceted/facet_grid.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Faceted/Faceted/facet_grid.Rmd")
library(ggplot2)
library(gridExtra)
# predictor variables
groups <- c(rep("A", 500), rep("B", 500), rep("C", 500), rep("D", 500), rep("E", 500), rep("F", 500))
categories <- as.factor(floor(runif(3000, min=1, max=4)))
# response variable
values <- c(rnorm(500, mean=50, sd=10), rnorm(500, mean=30, sd=5), rnorm(500, mean=40, sd=10), rnorm(500, mean=60, sd=20), rnorm(500, mean=70, sd=10), rnorm(500, mean=55, sd=15))
# dataframe
df <- data.frame(groups, categories, values)
```

`facet_grid()` allows for creating a multi-faceted plot. In this plot, each of the panels displays the data matching the combination of 2 levels, one for each of two categorical predictor variables. This results often in a more readable figure than a larger, unique figure where all the data point overlap.

Let’s illustrate this with the following histogram of distribution. 
```{r hist, echo=FALSE}
ggplot(df, aes(values)) +    
  geom_histogram(bins=50)
```

The dataframe for drawing this histogram is described below. It contains 3000 entries described by one of 6 levels (`A`, `B`, `C`, `D`, `E` and `F`) in the categorical variable `groups` and by one of 3 levels (`1`, `2` and `3`) in the categorical variable `categories`:

```{r variables and dataframe2, eval=FALSE}
# predictor variables
groups <- c(rep("A", 500), rep("B", 500), rep("C", 500), rep("D", 500), rep("E", 500), rep("F", 500))
categories <- as.factor(floor(runif(3000, min=1, max=4)))
# response variable
values <- c(rnorm(500, mean=50, sd=10), rnorm(500, mean=30, sd=5), rnorm(500, mean=40, sd=10), rnorm(500, mean=60, sd=20), rnorm(500, mean=70, sd=10), rnorm(500, mean=55, sd=15))
#dataframe
df <- data.frame(groups, categories, values)
```
<br/><br/>


Here are 2 color-coded histograms of distribution, one showing the levels of `groups` (left plot) and the other one showing the levels of `categories` (right plot):
```{r histcolors, eval=FALSE}
ggplot(df, aes(values, fill=groups)) +        # left plot, levels of groups
  geom_histogram(bins=50) +
  scale_fill_viridis_d() 
  
ggplot(df, aes(values, fill=categories)) +    # right plot, levels of categories
  geom_histogram(bins=50) +
  scale_fill_viridis_d()
```
```{r histcolors2, echo=FALSE, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values, fill=groups)) +        # left plot, levels of groups
    geom_histogram(bins=50) +
    scale_fill_viridis_d(),
  ggplot(df, aes(values, fill=categories)) +    # right plot, levels of categories
    geom_histogram(bins=50) +
    scale_fill_viridis_d(),
  ncol=2)
```
<br/><br/>

Using `facet_grid()`, you can immediately get an overview over all the groups and categories in the form of a 6 x 3 matrix. you just need to indicate `groups` and `categories` between the parentheses, and separated by `~`:
```{r facet}
ggplot(df, aes(values)) +
  geom_histogram(bins=50) +
  facet_grid(categories~groups)
```
<br/><br/>

Note that the order of the elements between the parentheses in `facet_grid()` determines the order of the matrix:
```{r facet2}
ggplot(df, aes(values)) +
  geom_histogram(bins=50) +
  facet_grid(groups~categories)
```

If you are not interested in combinations of groups/categories, you may choose to display a matrix based only on one element. Here you may choose to display only the groups in rows. Use `rows = vars()` in the following way:

```{r rowvar}
ggplot(df, aes(values)) +
  geom_histogram(bins=50) +
  scale_fill_viridis_d() +
  facet_grid(rows = vars(groups))
```
<br/><br/>

Same request, but this time displayed in columns with `cols = vars()`:

```{r colvar}
ggplot(df, aes(values)) +
  geom_histogram(bins=50) +
  facet_grid(cols = vars(groups))
```
<br/><br/>

By default in `facet_grid()`, the scales of the X- and Y-axes are fixed. This means that all the mini-plots will have the same scales, making them easy to compare. You may override that setting with `scales =`. This argument has 4 options: 

+ `"free"`: all the plots will have different scales on both axes based on the data content of the individual plots,
+ `"free_x"`: only the scale of the Y-axis will be fixed and common to all plots,
+ `"free_y"`: only the scale of the X-axis will be fixed and common to all plots,
+ `"fixed"`: both the X- and Y-axis are common to all plots (*default*).


Here are two examples, with `free_x` to the left and `free` to the right:
```{r free, eval=FALSE}
ggplot(df, aes(values)) +    # left plot, free_x
  geom_histogram(bins=50) +
  facet_grid(categories~groups, scales = "free_x") 

ggplot(df, aes(values)) +    # right plot, free
  geom_histogram(bins=50) +
  facet_grid(categories~groups, scales = "free")
```
```{r free2, echo=FALSE, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values)) +    # left plot, free_y
    geom_histogram(bins=50) +
    facet_grid(categories~groups, scales = "free_x") ,
  ggplot(df, aes(values)) +    # right plot, free
    geom_histogram(bins=50) +
    facet_grid(categories~groups, scales = "free"),
  ncol=2)
```


<br/><br/>



***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
