---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Grid of clustered/grouped barplots
# author: jonathan
# revised: 14/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Faceted/grid-clustered-bars.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Faceted/Faceted/grid-clustered-bars.Rmd")
library(ggplot2)
library(gridExtra)

# predictor variables
location <- rep(c("Lygra", "Østerbø"), each = 36)
year <- factor(rep(rep(2017:2019, each = 12), 2))
month <- factor(rep(month.abb, 6), levels = c(month.abb))

# predictor variables
precipitations <- c(135.8, 88.4, 91.0, 111.7, 31.0, 120.2, 69.6, 142.2, 83.8, 208.6, 229.4, 141.4,
                    109.8, 52.8, 37.7, 69.8, 50.8, 47.3, 30.7, 186.3, 294.6, 245.9, 57.6, 46.4,
                    33.8, 62.6, 111.2, 12.8, 46.0, 125.4, 95.2, 180.8, 148.6, 132.0, 21.6, 224.4,
                    34.2, 22.8, 61.0, 32.0, 29.6, 44.7, 42.4, 36.8, 61.8, 88.2, 92.6, 80.2,
                    48.0, 11.2, 5.6, 24.2, 17.2, 31.6, 43.4, 75.3, 144.0, 103.0, 12.4, 36.6,
                    29.4, 15.4, 57.6, 3.2, 25.2, 52.4, 65.0, 67.4, 79.0, 18.2, 7.8, 92.0)

```

In this tutorial, we will see how to make a grid of clustered/ grouped bar plots using `facet_wrap()`. Such a grid may be useful when your data set contains several categorical predictor variables, and displaying the data in a single graph makes it hardly comprehensible. Compared to a grid of non-grouped bar plots (introduced [HERE](https://biostats.w.uib.no/grid-bar-plots/){target="_blank"}), it allows for side-by-side comparison of related groups in the form of color-coded clusters. 


If you are not so familiar with bar plots, clustered bar plots or `facet_wrap()`, have a quick look at these pages:

+ [Creating a bar plot](https://biostats.w.uib.no/creating-a-bar-plot/){target="_blank"},
+ [Clustering/grouping bars](https://biostats.w.uib.no/clustering-grouping-bars/){target="_blank"},
+ [face_wrap()](https://biostats.w.uib.no/facet_wrap/){target="_blank"}.

<br/>

We will plot the precipitations recorded monthly in 2017, 2018 and 2019 at two Norwegian locations: Lygra and Østerbø. We will thus have three categorical variables: `month`, `year` and `location`, and one response variable `precipitations`. Here is the code for the dataframe:

```{r dataframe}
# dataframe
df <- data.frame(location, year, month, precipitations)
# structure of the dataframe
str(df)
```
<br/><br/>

Our plan is to make a grid displaying 2 panels, each of which is a clustered bar plot. In these bars plot, the predictor variable `month` and the response variable `precipitations` shall be plotted on the X- and Y-axis, respectively. Three color-coded bars shall display the monthly precipitations for the recorded years in clusters defined by the variable `year`. Finally, the grid shall show the two locations on top of each other (two panels displayed in a single column). To obtain this grid, we must:

+ map the data for the bar plots by setting up the æsthetics in the following manner: `aes(x = month, y = precipitations)`,
+ draw the bars by using `geom_col()`,
+ define the color-coded clusters as a function of the variable `year` with `aes(fill = year), position = "dodge"`
+ create the grid by setting up `facet_wrap()` like this: `facet_wrap(~location, ncol=1)`.

Here is the code, and the corresponding faceted plot:

```{r plot}
ggplot(df, aes(x = month, y = precipitations)) +
  geom_col(aes(fill = year), position = "dodge") + 
  facet_wrap(~location, ncol=1)
```
<br/><br/>

If the plan was to set up a grid with `location` in a single row instead of a single column, we should have used `facet_wrap(~location, nrow=1)`:
```{r plot2}
ggplot(df, aes(x = month, y = precipitations)) +
  geom_col(aes(fill = year), position = "dodge") + 
  facet_wrap(~location, nrow=1)
```
However, in this particular case, the design is not so attractive since the bars become very thin and the labels of the X-axes quite close to each other. 
<br/><br/>

### `facet_wrap()` vs `facet_grid()`  

Here we have made use of `facet_wrap()`, but we could have written the code with `facet_grid()` to achieve the same results. `facet_wrap()` is easier to use when making a grid based on one variable (here `location`); on the opposite, `facet_grid()` requires the use of two variables, unless overridden by:

+ `rows = vars( )` which shows the levels of the given variable as rows,
+ `cols = vars( )` which shows the levels of the given variable as columns.

```{r plot3, eval=FALSE}
ggplot(df, aes(x = month, y = precipitations)) +          # left plot, levels in rows
  geom_col(aes(fill = year), position = "dodge") + 
  facet_grid(rows = vars(location))

ggplot(df, aes(x = month, y = precipitations)) +          # right plot, levels in columns
  geom_col(aes(fill = year), position = "dodge") + 
  facet_grid(cols = vars(location))
```
```{r plot4, echo=FALSE, fig.height=4, fig.width=10}
grid.arrange(
  ggplot(df, aes(x = month, y = precipitations)) +          # left plot, levels in rows
    geom_col(aes(fill = year), position = "dodge") + 
    facet_grid(rows = vars(location)),
  ggplot(df, aes(x = month, y = precipitations)) +          # right plot, levels in columns
    geom_col(aes(fill = year), position = "dodge") + 
    facet_grid(cols = vars(location)),
  ncol=2)
```
<br/><br/>

### Improving the look

You may improve the look of a grid by tuning the labels of the matrix. This is further explained [HERE](https://biostats.w.uib.no/modifying-labels-in-faceted-plots/){target="_blank"}.  
Since colors might be important for the interpretation of the data, have a look at [this page](https://biostats.w.uib.no/color-scale-for-discrete-variables/){target="_blank"} which shows how to color frames and/or boxes as a function of a variable, and [this page](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"} that tells you more about color palettes.
<br/><br/>

### Alternative plot

This data set may be alternatively plotted in the form of a *grid of non-grouped bar plots*. [HERE](https://biostats.w.uib.no/grid-bar-plots/){target="_blank"} is a tutorial for making such a plot.
```{r plot5, echo=FALSE}
ggplot(df, aes(x = month, y = precipitations)) +
  geom_col() +
  facet_grid(year~location) 
```




<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
