---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Grid of bar plots
# author: jonathan
# revised: 5/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Faceted/grid-bar.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Faceted/Faceted/grid-bar.Rmd")
library(ggplot2)
```

In this tutorial, we will see how to make a grid of bar plots using `facet_grid`. Such a grid may be useful when your data set is relatively complicated (with several categorical predictor variables), and displaying the data in a single graph makes it hardly comprehensible. If you are not so familiar with bar plots or `facet_grid()`, have a quick look at these two pages:

+ [Creating a bar plot](https://biostats.w.uib.no/creating-a-bar-plot/){target="_blank"},
+ [face_grid()](https://biostats.w.uib.no/facet_grid/){target="_blank"}.

<br/><br/>

We will represent and compare the precipitations recorded monthly in 2017, 2018 and 2019 at two Norwegian locations: Lygra and Østerbø. We will thus have three categorical variables: `month`, `year` and `location`, and one response variable `precipitations`. Here is the code for these variables and the dataframe:

```{r variables and dataframe}
# predictor variables
location <- rep(c("Lygra", "Østerbø"), each = 36)
year <- factor(rep(rep(2017:2019, each = 12), 2))
month <- factor(rep(month.abb, 6), levels = c(month.abb))

# predictor variables
precipitations <- c(135.8, 88.4, 91.0, 111.7, 31.0, 120.2, 69.6, 142.2, 83.8, 208.6, 229.4, 141.4,
                    109.8, 52.8, 37.7, 69.8, 50.8, 47.3, 30.7, 186.3, 294.6, 245.9, 57.6, 46.4,
                    33.8, 62.6, 111.2, 12.8, 46.0, 125.4, 95.2, 180.8, 148.6, 132.0, 21.6, 224.4,
                    34.2, 22.8, 61.0, 32.0, 29.6, 44.7, 42.4, 36.8, 61.8, 88.2, 92.6, 80.2,
                    48.0, 11.2, 5.6, 24.2, 17.2, 31.6, 43.4, 75.3, 144.0, 103.0, 12.4, 36.6,
                    29.4, 15.4, 57.6, 3.2, 25.2, 52.4, 65.0, 67.4, 79.0, 18.2, 7.8, 92.0)

# dataframe
df <- data.frame(location, year, month, precipitations)
```
<br/><br/>

Our plan is to make a matrix displaying 6 panels, each of which is a bar plot. In these bars plot, the predictor variable `month` and the response variable `precipitations` shall be displayed on the X- and Y-axis, respectively. Eventually the matrix will show `location` in columns and `year` in rows. To obtain this matrix, we must:

+ map the data for the bar plots by setting up the æsthetics in the following manner: `aes(x = month, y = precipitations)`,
+ draw the bars by using `geom_col`,
+ create the matrix by setting up `facet_grid()`: `facet_grid(year~location)`.

Here is the code, and the corresponding faceted plot:

```{r plot}
ggplot(df, aes(x = month, y = precipitations)) +
  geom_col() + 
  facet_grid(year~location)
```
<br/><br/>

If the plan was to set up a matrix with `location` in rows and `year` in columns, we shoulld have inverted the variables in `facet_grid()`:
```{r plot2}
ggplot(df, aes(x = month, y = precipitations)) +
  geom_col() + 
  facet_grid(location~year)
```
However, in this particular case, the X-axis is overloaded and the labels are unreadable. 
<br/><br/>

### Improving the look

You may improve the look of a grid by tuning the labels of the matrix. This is further explained [HERE](https://biostats.w.uib.no/modifying-labels-in-faceted-plots/){target="_blank"}.
<br/><br/>

### Alternative plot

This dataframe could be been alternatively plotted in the form of a *grid of clustered/grouped bars*. [HERE](https://biostats.w.uib.no/grid-clustered-barplots/){target="_blank"} is a tutorial for making such a plot.
```{r plot3, echo=FALSE}
ggplot(df, aes(x = month, y = precipitations, fill=year)) +
  geom_col(position = "dodge") + 
  facet_grid(location~.) + 
  scale_fill_viridis_d()
```




<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->

