```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: So, what about our example?
# author: jonathan
# revised: 10/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/1-sample/example.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/1-sample/example.Rmd")
scores <- c(123, 135, 142, 127, 117, 134, 123, 133, 111, 130, 113, 118, 105, 126, 125, 122, 147, 106, 117, 129, 144, 112, 126, 140, 129, 115, 118, 129, 125, 137, 135, 138, 134, 124, 128, 135, 124, 137, 121, 120)
```

So, what about our university teacher and his bunch of clever students? First, let’s check what is known about the sample:

1. List of the things you should know:

    + the mean/reference of the population, `µ` (here µ=120)
    + the number of observations in your sample, `n` (here n=40)
    + the sample mean, M (here `M`=126.35)
    + the standard deviation for the sample, `sd` (here sd=10.14)
    + the standard deviation for the population, `SD` (unknown in our example)

2. Assumptions:

    + the variable is *continuous*
    + the data are *independent* (no relationship between observations)
    + the variable has a *normal distribution*
  
3. The null hypothesis to be tested:

    + `H0`: the sample mean M equals the population mean µ.
  
4. Our alternative hypothesis:

    + `Ha`: the sample mean M is *greater* than the population mean µ (our teacher has the feeling that the student are clever, meaning that their IQ is *higher* than average).
    
  
**We do not know the standard deviation of the population, SD**. Thus we have to run a **one-sample t-test**.
In R, we had previously stored our data in `scores`. We just need to enter the following command:
```{r ttest}
t.test(scores, alternative="greater", mu = 120)
```

As you may read above here, a p-value less than 0.05 (0.0001547) has been “delivered” by `t.test()`. This means that our alternative hypothesis `Ha` is accepted. Our teacher’s feeling about his students was correct, assuming that using IQ score for this purpose makes sense…

 

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
