```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Run a one-sample z-test
# author: jonathan
# revised: 10/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/1-sample/ztest.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/1-sample/ztest.Rmd")
scores <- c(123, 135, 142, 127, 117, 134, 123, 133, 111, 130, 113, 118, 105, 126, 125, 122, 147, 106, 117, 129, 144, 112, 126, 140, 129, 115, 118, 129, 125, 137, 135, 138, 134, 124, 128, 135, 124, 137, 121, 120)
install.packages("TeachingDemos")
library(TeachingDemos)


Assuming that you have stored your sample data in the variable `scores`, the command to use is `z.test(scores, mu = Y, stdev = W, alternative="ALT")` where:

+ Y shall be replaced by the value of the population mean,
+ W shall be replaced by the standard deviation of the population (since it is one of the conditions to run the test, see [here](https://biostats.w.uib.no/2-lets-choose-and-run-a-one-sample-test/){target="_blank"}),
+ `ALT` shall be replaced by either `greater` or `less` or `two.sided` depending on your alternative hypothesis `Ha`. The null hypothesis `H0` states that the sample mean is NOT different from the population mean. Your alternative hypothesis `Ha` is one of the following:
  - the sample mean is **greater** than the population mean (in that case, use `greater`),
  - the sample mean is **less than** the population mean >> (in that case, use `less`)
  - the sample mean is **either smaller or greater** than the population mean (in that case, use `two.sided`).
More info about `TeachingDemos` is available here.

Considering our previous example, this would look like:
```{r ztest, eval=FALSE, warning=FALSE}
z.test(scores, alternative="greater", mu=120, stdev=UNKNOWN)
```
However, as stated in the code, the *standard deviation is unknown*. Therefore, a z.test cannot be used.

**IF** the standand deviation was previously know and equal to 15, this would have been the code and the corresponding output:
```{r ztest2}
z.test(scores, alternative="greater", mu=120, stdev=15)
```

R returns several lines of text. One of them provides a **p-value** while the next line states the **alternative hypothesis** which depends on the parameter `alternative` that you have entered in the `z.test()`. This alternative hypothesis `Ha` is considered valid when the p-value is less than 0.05.



 

Read more about `z.test()` by simply typing `?z.test` in the R console.
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
