---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: comparing two means - Mann-Whitney U test
# author: jonathan
# revised: 10/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/2-sample/mann-whitney.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/2-sample/mann-whitney.Rmd")
Location_A <- c(6.3, 6.3, 8.1, 8.9, 11.6, 10.4, 6.8)
Location_B <- c(6.9, 7.7, 7.9, 9.1, 11.3, 11.8, 14.2)
```
The Mann-Whitney U test (also known as Wilcoxon rank-sum test) allows for comparison of two groups which do not necessarily originate from normal distributions. This means that you may use this test even if the samples have a normal distribution. This is thus a good alternative to [Student’s t-test](http://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"} when the assumption of normality might constitute a problem.

The function to use in R is `wilcox.test(x, y)` where `x` and `y` are the vectors containing the samples. Let’s use the same example as we have use for [Student’s t-test](http://biostats.w.uib.no/comparing-two-means-t-test/){target="_blank"}:

```{r student}
wilcox.test(Location_A, Location_B)
```
So, the test results in a p-value *greater* than 0.05, thus indicating that the null hypothesis is not to be rejected. The sample means are not significantly different.



<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
