---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a circular line chart
# author: jonathan
# revised: 4/4/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Multiple/circular-line.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Multiple/circular-line.Rmd")
library(ggplot2)
week.day <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
day <- factor(rep(week.day, each = 24), levels=week.day)
time <- factor(rep(1:24, 7), levels=1:24)
temperature <- c(3.7,1.8,1.2,1.3,0.8,0.8,0.9,4.5,6.9,8.6,10.2,9.8,10.6,11.8,10.6,10.5,9.8,9.8,8.8,7.6,5.6,4.8,4.6,3.7,4.0,3.7,1.8,2.4,2.2,0.6,0.7,2.4,7.5,10.9,11.6,12.6,12.6,13.7,13.9,13.9,13.8,13.1,11.9,11.4,10.5,9.2,8.9,8.1,7.3,6.0,5.7,5.7,5.8,4.6,4.6,5.2,6.8,7.3,9.3,12.0,11.8,12.0,12.6,12.8,13.4,13.8,13.3,12.8,11.4,10.2,9.3,8.8,8.5,8.0,7.8,7.4,7.2,6.8,7.1,7.9,9.4,10.6,12.3,13.8,14.7,15.3,15.6,15.4,15.1,15.0,13.9,13.0,12.2,9.8,8.9,7.8,8.2,8.4,8.2,7.3,6.9,8.0,6.4,9.3,10.6,13.9,15.0,16.2,15.9,16.1,16.3,16.2,15.7,15.2,15.3,14.6,13.1,11.7,10.7,11.0,11.1,10.7,9.2,9.2,8.7,8.4,8.7,9.8,12.1,13.1,13.8,13.6,14.1,14.1,14.9,15.6,15.4,14.2,13.7,12.9,12.2,10.6,10.7,10.2,9.7,9.1,9.0,8.1,7.8,7.5,7.3,7.4,7.9,8.8,9.5,10.5,12.1,13.5,14.6,14.7,14.7,13.2,12.4,12.3,12.9,12.3,11.2,11.8)
```

A circular line chart may be useful when plotting cyclic dataseries resulting from the combination of a measurement variable and a time-related numerical/continuous variable. It usually helps highlighting a cyclic pattern, a repetition (or absence of repetition) in time. In principle, a circular line chart is a regular line chart which is projected into a polar coordinate system.

Before going any further, if you are not so familiar with line plots or polar coordinate systems, have a quick look at these two pages:

+ [Creating a line plot](https://biostats.w.uib.no/creating-a-line-plot/){target="_blank"},
+ [Creating a pie chart](https://biostats.w.uib.no/creating-a-pie-chart/){target="_blank"}.

Here we will plot the temperatures (response variable) recorded hourly (predictor variable 1) at Lygra, Vestland, between Apr. 15^th^ and 22^nd^, 2019 (predictor variable 2). 

The dataframe for this tutorial is as follows:
```{r dataframe}
# dataframe
df <- data.frame(day, time, temperature)
# structure of the dataframe
str(df)
```
<br/><br/>

Using the coding strategy presented [here](https://biostats.w.uib.no/creating-a-multiple-line-plot/){target="_blank"}, we can start by building a multiple line chart. The function to use for drawing the lines is `geom_line()`. We must map the variables with the function `aes()` which will look more or less like this: `aes(time, temperature, day)`.  Our plan is thus to:

+ plot the predictor variable `time` on the X-axis, 
+ plot the response variable `temperature` on the Y-axis,
+ use both `group=` and `color=` to display the data in colored lines based on the predictor variable `day`,
+ draw the lines with `geom_line()`.

The code for the plot is as follows:
```{r plot}
ggplot(df, aes(x = time, y = temperature, group = day, color = day)) +
  geom_line()
```
<br/><br/>

Then, to transform this regular line plot into a circular line plot, we project it into a polar coordinate system with `coord_polar()`:
```{r plot2}
ggplot(df, aes(x = time, y = temperature, group = day, color = day)) +
  geom_line() +
  coord_polar()
```
<br/><br/>

This plot is almost acceptable, but the gap between days (which is materialized by the absence of line between 24 and 1 every day) is suboptimal. One solution is to use `geom_polygon()` instead of `geom_line()` as `geom_polygon()` links the first and the last data points of a series with a segment of corresponding color. Note that `geom_polygon()` must (in our case) be used with the argument `fill = NA` otherwise the polygone shape resulting from that code will not allow to see the overlapping series correctly:

```{r plot3}
ggplot(df, aes(time, temperature, group=day, color=day)) +
  geom_polygon(fill = NA) + 
  coord_polar() 
```
<br/><br/>

Finally we may improve the global look of the plot by increasing the thickness of the lines with `size=` and by applying a nicer palette of colors such as viridis:
```{r plot4}
ggplot(df, aes(time, temperature, group=day, color=day)) +
  geom_polygon(size=1.2, fill = NA) + 
  scale_color_viridis_d() +
  coord_polar() 
```

<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
