---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Clustering/grouping boxplots
# author: jonathan
# revised: 28/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Multiple/clustering-boxplots.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Multiple/clustering-boxplots.Rmd")
library(ggplot2)
ID <- 1:400
values <- c(rnorm(100, mean=25, sd=5), rnorm(100, mean=30, sd=4), rnorm(100, mean=15, 3), rnorm(100, mean=40, sd=5))
category1 <- c(rep("A", 100), rep("B", 100), rep("C", 100), rep("D", 100))
category2 <- rep(c("1","2"), 100)
```

When designing a boxplot for a data set with two or more categorical variables, one may need to group/cluster some of the boxes by category. Such a clustered (grouped) boxplot is very easy to create if you know already how to draw boxplots.

Before going any further, if you are not so familiar with boxplots, have a quick look at this page:

+ [Creating a (multiple) boxplot](https://biostats.w.uib.no/creating-a-multiple-boxplot/){target="_blank"}.

Here we will take the following example where `values`is the response variable, and `category1` and `category2` the categorical predictor variables. The dataframe for this tutorial is as follows:
```{r dataframe}
# dataframe
df <- data.frame(values, category1, category2)
# structure of the dataframe
str(df)
```


As you may guess from the structure of the dataframe above, `category1` has 4 levels (`A`, `B`, `C` and `D`) and `category2` has only 2 levels (`1` and `2`). 

As for any boxplot, the function to use for drawing the bars is `geom_boxplot()`. Since we have two categorical variables and the response variable to map, the function `aes()` will look more or less like this: `aes(values, category1, category2)`. However, we have to order properly the variables and ask ggplot to group and color the boxes according to one of categories. We will use `fill=` to do so. Our plan is to:

+ plot `values` on the Y-axis,
+ plot `category1` on the X-axis, 
+ cluster the `category2` levels with `fill=`, 
+ draw the plot with `geom_boxplot()`.

Here is the code:
```{r plot}
ggplot(df, aes(x = category1, y = values, fill = category2)) +
  geom_boxplot()
```
<br/><br/>

Alternatively we may replace `fill=` with `color=`. While `fill=` colors the entire boxes, `color=` changes the color of the box frames and lines only:
```{r plot2}
ggplot(df, aes(x = category1, y = values, color = category2)) +
  geom_boxplot()
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->

