---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Adding a smoother line to a scatter plot
# author: jonathan
# revised: 21/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Multiple/smoother-scatterplot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Multiple/smoother-scatterplot.Rmd")
library(ggplot2)
values_x <- 1:600
values_y <- c(rnorm(200, mean=40, sd=15), rnorm(200, mean=60, sd=15), rnorm(200, mean=50, sd=15))
```

Scatter plots that display a huge amount of data tend to be very confusing, and it is often difficult to define whether there is a pattern or trend in the data distribution. One strategy to better visualize a trend is to add a smoother, a line that removes the noise or inherent variation in the data set.

Before going any further, if you are not so familiar with scatter plots, have a quick look at this page:

+ [Creating a scatter plot](https://biostats.w.uib.no/creating-a-scatter-plot/){target="_blank"}.

Adding a smoother is simply done by adding an extra layer in the form of the function `geom_smooth()` to the code of the scatter plot. Let’s take the following example. The dataframe for this example is as follows:

```{r dataframe}
# dataframe
df <- data.frame(values_x, values_y)
# structure of the dataframe
str(df)
```

And the corresponding scatter plot is:
```{r scatterplot}
ggplot(df, aes(values_x, values_y)) +
  geom_point()
```
<br/><br/>

To add the smoother, we add `geom_smooth()` to the code:
```{r scatterplot with smoother}
ggplot(df, aes(values_x, values_y)) +
  geom_point() +
  geom_smooth()
```
As you may see on the plot above, the plot displays now a smoother (the blue line) AND the standard error (grey area surrounding the smoother). A warning message also appears in the console. It lists the default parameters that `geom_smooth()` has used, since we have not given it any information or arguments.
<br/><br/>

It is possible to adjust the shape of the smoother by playing with the *wiggliness* of the line. To do so, use the argument `span=` and give it a value between 0 and 1:

```{r ajusted smoother}
ggplot(df, aes(values_x, values_y)) +
  geom_point() +
  geom_smooth(span = .2)
```
<br/><br/>

If you are interested in the smoother but not in the standard error, you can ask `geom_smooth()` to omit it. Just add the argument `se=FALSE`:
```{r smoother-noSE}
ggplot(df, aes(values_x, values_y)) +
  geom_point() +
  geom_smooth(se=FALSE)
```
<br/><br/>

`geom_smooth()` can also be used to overlay a *line of best fit* . For that, we use the argument `method="lm"` in `geom_smooth()`:
```{r lm}
ggplot(df, aes(values_x, values_y)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```
<br/><br/>

Finally, you may adjust the thickness, color and linetype with `size=`, `color=` and `linetype=`, respectively:
```{r style}
ggplot(df, aes(values_x, values_y)) +
  geom_point() +
  geom_smooth(size=2, color="orange", linetype="dashed")
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->

