# Installing packages {#sec-installing-packages}

```{r setup, include=FALSE}

source("R/setup.R")
```

::: callout-note
## In this chapter, you will

- learn how packages extend what R can do
- install packages from CRAN and from GitHub
- load packages so their functions/data are available
:::

Packages are add-ons to base R (the R base package) that expand the computing possibilities of R by adding new functions, classes, documentation, data sets, etc.
When installing RStudio for the first time, a long list of packages comes along.
You will find this list in the tab `Packages` (see @sec-packages-tab).

This section will show you how to install additional packages.
Every time you install a new one, R imports all necessary files into a local *library*, but does not activate it.
You will have to remember to activate the new package with `library()` every time your project require items or functions from that package.

## Packages published on CRAN

If you want to install a package published on [CRAN](https://cran.r-project.org/){target="_blank"} repository, you may use the function `install.packages()`.
Simply type its name and add quotation marks `"` `"`.
Here is an example with the package `tidyr`:

```{r install-packages, echo = TRUE, eval=FALSE}
install.packages("tidyr")
```

You may use the exact same code to update the package later on.

If you want to update all packages at once, use the following code:

```{r update-packages, echo = TRUE, eval=FALSE}
update.packages()
```

## Packages published on GitHub

If you want to install a package published on [GitHub](https://github.com/){target="_blank"}, you may use the function `remotes::install_github()`.
First you need to install the package `remotes`.
Type first the name of the repository, then `\` and the name of the package.
Here also, you must add quotation marks `"` `"`.

## Recommended packages

The present website refers to the use of several packages that we recommend, not only for learning R, but also for working with your course assignments and projects.
These packages are:

-   `tidyverse`
-   `biostats.tutorials`,
-   `palmerpenguins`.

These packages will be used throughout the website, so get ready to install them now.
Here is the procedure.

### The tidyverse

The tidyverse is an amazing toolbox which contains a growing, evolving collection of R packages for data science.
The packages are developed on the same philosophy and are fully compatible with each other.
Some of the tidyverse packages help you read files and import data, some packages let you draw plots and make figures, others help you rearrange your data set.
Have a look at [the tidyverse webpages](https://www.tidyverse.org/){target="_blank"} to explore the collection.

Install and activate the tidyverse with the following lines:

```{r recommended-tidyverse, echo=TRUE, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

### palmerpenguins

`palmerpenguins` is a package that provides you with two *real* data sets.
They contain measurements from three penguin species found in the Palmer archipelago, Antarctica.
Several variables such as species, island, and body mass are included, and no less than 344 observations are found in the table.
These data sets will be used in the upcoming sections of this website.

Install and activate the package with these lines:

```{r install-the-penguins, echo=TRUE, eval=FALSE}
install.packages("palmerpenguins")
library(palmerpenguins)
```

Type this following line in the console if you want to know more about `palmerpenguins`:

```{r citation-penguins, echo=TRUE, eval=FALSE}
citation("palmerpenguins")
```

### biostats.tutorials {#biostatstutorials}

Along with the present website, our team has developed a package with tutorials called `biostats.tutorials`.
This package will help you learn and practice R functions and concepts.
Several of the upcoming chapters refer to this package.
Install `biostats.tutorials` with this code:

```{r install-the-tutorial, echo=TRUE, eval=FALSE}
#install.packages("remotes")
remotes::install_github("biostats-r/biostats.tutorials")
```

Now load the package with `library(biostats.tutorials)` and the tutorials should appear in the tab `Tutorial` (see @sec-tutorial-tab).
Exercises will tell you when to run a tutorial.
Click the "Start Tutorial" button to start the tutorial (this may take a few seconds to start the first time).

### Exercise {.unnumbered .facta .toc-ignore}

Install the `tidyverse`, `palmerpenguins` from CRAN and `biostats.tutorials` from GitHub using the code above.
