---
format:
  html:
    css: include/webex.css
    include-after-body: include/webex.js
---

# Set-up quiz



```{r}
#| label: setup-quiz
#| file: !expr here::here("Templates/webex.R")
#| include: false
```

## Quiz

`r total_correct(elem = "h3", args = "style='color:#003366;'")`


```{r}
#| label: questions
#| echo: false
#| output: asis
questions <- list(
    list(
    question = "Why should you set RStudio option to never restore .RData on start-up",
    choice = c(
      "to make starting RStudio faster",
      "to make your analysis more reproducible",
      "reduce the probability of bugs in your code",
      answer = "all of the above"
    ),
    hint = "See @sec-customise-rstudio"
  ),
    list(
    question = "Your R console shows '+' instead of the usual '>' prompt. What does this mean?",
    choice = c(
      "R is broken and need to be re-installed",
      answer = "the current command is unfinished. Perhaps a bracket or quote mark is missing.",
      "R is now a calculator and is waiting for a number"
    ),
    hint = "See @sec-console-tab. Usually best to press {{< kbd escape >}}. "
  ),
  list(
    question = "Why should you use RStudio projects for your work?",
    choice = c(
      "because Richard will hassle you until you do",
      answer = "becase they make R so much easier to work with"
    ),
    hint = "See #sec-working-project"
  ),
    list(
    question = "You have installed the `tidyverse` package. Which command do you need to use to activate it?",
    choice = c(
      "`install.packages(\"tidyverse\")`",
      "`libraries(tidyverse)`",
      answer = "`library(tidyverse`&#8288;`)`",
      "`package(tidyverse)`",
      "`Library(tidyverse)`"
    )
  )
)

print_multichoice(questions)


```






## Checker

Now we can use the checker package to check that your computer has the correct versions of R and RStudio installed, the required packages, and that the recommended RStudio options are set.

First install the `checker` package.

```{r}
#| label: checker-1
#| eval: false
install.packages("checker")
```

Now run

```{r}
#| label: checker-2
#| eval: false
checker::chk_requirements("https://raw.githubusercontent.com/biostats-r/biostats/main/checker/basic.yaml")
```

Please address any issues `checker` finds before continuing to the next chapter.

