# Working with single tables in `dplyr`

```{r setup, include=FALSE}
# continue knitting after errors
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
options(tibble.print_min = 3)
# packages
library("tidyverse")

source("../Templates/biostats_theme.R")
```

Data frames are usually the most convenient objects for storing, plotting or analysing data in R. 
We also need to be able to manipulate data in data frames.

This tutorial will show you how to manipulate data frames using the `dplyr` package, part of `tidyverse`.

Everything that can be done with `dplyr` can be done with base R. 
The `dplyr` solution is usually easier to write and understand, and can be faster.

There are many function in `dplyr`, this tutorial focuses on the functions I find most useful.

### Load the package

You can load `dplyr` with 

```{r load-dplyr}
library("dplyr")
```

But is is usually more convenient to load `tidyverse` so that you get `ggplot2`, `dplyr`, `readr` and other useful packages with one command.

```{r load-packages, eval = FALSE}
library("tidyverse")
```


### More feedback `tidylog`

We can get more feedback on what `dplyr` functions have done to the data by loading the `tidylog` package.

```{r tidylog, echo = TRUE, message = FALSE}
library("tidylog")
# if you are using conflicted, you will need to set conflict_prefer() for all dplyr function
# use purrr::map to iterate over tidylog functions 
# library(conflicted)
# map(getNamespaceExports("tidylog"), ~conflict_prefer(.x, winner = "tidylog"))
```

More than once, `tidylog` has helped identified bugs in my code.

## The `penguins` dataset

This tutorial will use the `penguins` dataset from the [`palmerpenguins` package](https://github.com/allisonhorst/palmerpenguins).
This dataset includes measurements of three species of penguin.

```{r penguins}
# Load the data
data("penguins", package = "palmerpenguins")
# Show the data
penguins
```


## Selecting columns with `select()`

You can choose which columns of the data frame you want with `select()`.

The first argument is the data, which is supplied by the pipe `|>`, the next arguments are the names of the columns you want. 
The names do not need quote marks.

```{r select1}
#select species, bill_length_mm & bill_depth_mm
penguins %>% 
  select(species, bill_length_mm, bill_depth_mm)
```


This is equivalent to the base R code

```{r select1.5}
#select species, bill_length_mm & bill_depth_mm
penguins[, c("species", "bill_length_mm", "bill_depth_mm")] 
```
Remember that if you want to use the output of this code in a further analysis, you need to assign it to an object name with `<-`.

### `select()` helpers  

Sometimes we don't want to write out the names of all the columns we want to select.
We might not even know them all in advance. 
Fortunately there are some helper functions.

If you want to `select()` adjacent columns, you can use the notation `first:last`.

```{r select2}
#select species to bill_depth_mm
penguins %>% select(species:bill_depth_mm)
```

Sometimes it is easier to remove the columns you don't want. 
You can do this by putting a `-` in front of the column name.

```{r select3}
#select everything but year and sex
penguins %>% select(-year, -sex)
```

If there is a pattern to the column names that we want to select (or remove), there are some helper functions.
For example, to select columns that start with "bill", we can use `starts_with()`.

```{r select4}
#select bill_length_mm & bill_depth_mm" 
penguins %>% select(starts_with("bill"))
```

Conversely, if we want to select all columns that end with "mm", we can use `ends_with()`.
`contains()` is more flexible and `matches()` is the most powerful of the helper functions, using regular expressions to identify the columns (see the regular expression tutorial).

Sometimes, you might want to select all the columns of a certain type. 
For example, to select all the numeric columns we can use the `is.numeric` function inside `select()` with the helper `where()`.

```{r select5}
penguins %>% select(where(is.numeric)) # No brackets on the function
```

Other `is.*` functions exist, for example, `is.character` for text, `lubridate::is.Date` for dates.

You can also select columns by number (1 being the first column), but this is generally a bad idea because it makes the code difficult to understand and if a new column is added, or the column order is changed, the code will break.

Which of these strategies works best is context dependent.

## Renaming columns with `rename`

You can use `rename()` to rename columns

```{r rename1}
penguins %>% rename(Species = species)
```

The syntax is `new_name = current_name`.

You can also rename a column when selecting.
This is convenient if you are using `select()` anyway.

```{r rename2}
penguins %>% select(Species = species)
```


## Moving columns with `relocate()`

Sometimes it is useful to reorder the columns. 
This is never necessary for data analysis or plotting, but can be needed when making a table for presentation.

```{r relocate}
penguins %>% relocate(island)
```

The default is to move the named column first, the `.before` and `.after` arguments let you move the column into any position. 

## Filtering rows with `filter()`

Filtering rows that meet some condition is a very common task.

For example, to filter rows of `penguins` that have a bill length greater than 40 mm, we can use 

```{r filter1}
penguins %>% filter(bill_length_mm > 40)
```

This will filter out each row where the condition is `TRUE`.

The base R equivalent of this is 

```{r filter_base}
penguins[penguins$bill_length_mm > 40, ]
```

Generally, `filter` makes it easier to understand, especially with more complex criteria.


Other tests include 

- `==` exactly equals. Often a bad idea to use with numeric data 
- `near` safe function for testing equality of numeric data as it has a tolerance for rounding errors.

```{r filter-near}
sqrt(2) ^ 2 == 2 # should be true, but rounding errors 
sqrt(2) ^ 2 - 2 # the difference
near(sqrt(2) ^ 2, 2) # safe alternative
```

- `!=` not equal to
- `<` less than
- `<=` less than or equal to
- `>` greater than
- `>=` greater than or equal to
- `is.na()` for filtering by missing values.
- `between()` for filtering values with a range
- `%in%` is used when you want to test if a value is in a vector

```{r filter-in}
penguins %>% 
  filter(species %in% c("Adelie", "Chinstrap"))
#equivalent to 
penguins %>% 
  filter(species == "Adelie" | species == "Chinstrap") # with many alternatives, this gets long
```

### Filtering on multiple criteria

If we want to filter on multiple criteria, we need to decide whether we want all criteria to be `TRUE` (AND in Boolean logic), or for one or more to be `TRUE` (OR in Boolean logic).

If we want all criteria to be `TRUE`, we can separate them by a comma (or by an `&` if you want to be explicit).

```{r filter-AND}
penguins %>% 
  filter(bill_length_mm > 40, bill_depth_mm > 18)
```

If we want rows where any of the criteria is `TRUE`, we can separate them by a `|`.

```{r filter-OR}
penguins %>% 
  filter(bill_length_mm > 40 | bill_depth_mm > 18)
```

We can negate a criterion by putting `!` in front of it.
So to filter rows that do not have bills longer than 40 mm we can use

```{r filter-NOT}
penguins %>% filter(!bill_length_mm > 40)
```

Of course, in this example, we could also use `<=` as the test.

### Common errors

The commonest error is to use a single `=` rather than `==`.
Only the latter is a test of equality.
If you do this, the error message is quite helpful.

```{r filter-error, error=TRUE}
penguins %>% filter(species = "Chinstrap")
```

Another common error is to forget to quote any strings.

```{r filter-error2, error=TRUE}
penguins %>% filter(species == Chinstrap)
```

## Slicing the data with `slice()`

Sometimes it is useful to extract rows by row number. 

```{r slice}
penguins %>% slice(3:7)
```

You can use negative numbers to remove rows. 
Be careful using `slice()` as if the data change, different rows may be returned. 


## Distinct rows with `distinct()`

If there are duplicates in the data, we can remove these with `distinct()`.
`distinct()` with no extra arguments will remove duplicate rows. 

```{r}
penguins %>% distinct()
```

If we are only interested in some of the columns, we can supply the names of these columns.

```{r}
penguins %>% distinct(island)
```

Other columns will be removed unless the argument `.keep_all = TRUE` is used.

## Random rows

Sometimes you want to sample rows at random from a data.frame. 
This can be done with `slice_sample()`. 
This can either sample a constant _n_ rows or constant fraction of the rows depending on whether the `n` or `prop` argument is used.

```{r sample-n}
penguins %>% slice_sample(n = 10)
```

## Mutating and adding columns with `mutate`

The function `mutate()` can add an new column or replace an existing one.

To make a new column called `body_mass_kg` we can use

```{r}
penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000)
```

There are lots of functions that are useful to use with mutate. 
Any function that returns either a single value or as many values as are in the data can be used.

`mutate()` is very useful when cleaning data.

- See text manipulation tutorial for cleaning text with the `stringr` package.
- See date and time tutorial for cleaning dates and times with the `lubridate` package.


## Summarising data with `summarise()`

`summarise()` lets us summarise data.
We can use it if we want to calculate a summary statistic of the data. 
Remember to separate arguments with a comma.

```{r summarise1}
penguins %>% summarise(
  flipper_len_mean = mean(flipper_length_mm, na.rm = TRUE), 
  flipper_len_sd = sd(flipper_length_mm, na.rm = TRUE)
  )
```

Only the columns created in the `summarise()` and any grouping columns (see below) will be kept.

### Summarising multiple columns

Sometimes you want to summarise multiple columns at the same time.
This can be done with the `across()` helper function. 
`across()` needs to be told which columns to process and what function or functions to use.

```{r across}
penguins %>%
   summarise(across(c(bill_length_mm, bill_depth_mm), .fns = mean, na.rm = TRUE))

#using a list of functions
penguins %>%
  summarise(across(.cols = starts_with("bill"), .fns = list(sd = sd, mean = mean), na.rm = TRUE))
```

You can also use `across()` with `mutate()` to mutate several columns at the same time.


## Grouping data with `group_by`

`group_by()` changes the way that many of the `dplyr` functions work.
Instead of working on the entire dataset, they now work on each group in the data

To find the mean flipper length for each species, we need to `group_by()` species and then `summarise()`.

```{r group-by}
penguins %>% 
  group_by(species) %>% 
  summarise(mean_flipper_length = mean(flipper_length_mm))
```


Grouped data can be ungrouped with `ungroup()`.
This can help prevent surprises!


## Sorting with `arrange()`

To sort the data frame by one or more of the variables we can use `arrange()`.

```{r arrange1}
penguins %>% arrange(bill_length_mm, bill_depth_mm)
```

This will sort smallest first. 
To reverse the sort order, use `desc()`

```{r arrange2}
penguins %>% arrange(desc(bill_length_mm), desc(bill_depth_mm))
```

 
## Counting rows with `count()` and `n()`

The function `n` can count how many rows there are in the each group (or the entire data frame if it is not grouped).
It can be used with either `mutate()` or `summarise()`.

```{r n}
penguins %>% 
  group_by(species) %>% 
  summarise(n = n())
```

Or with `count()`

```{r count}
penguins %>% 
  count(species)
```

`count()` is more concise, but `group_by()` and `summarise()` is useful when you need to calculate more variables (for example mean and standard deviation). 



## Common problems

### Non standard names

Ideally column names should follow the standard rules for naming objects in R - UPPER and lower case letters, numbers, "." and "_" with the first character being a letter (or a dot if you want an invisible object).
Sometimes when you import data, it has non-standard names with spaces or extra characters.
If you need to refer to a column name that doesn't follow the rules, you need to enclose it with back-ticks.
```{r non-standard-names}
df <- tibble(`Region/Country` = "Norway", value = 42)
df
df %>% rename(region_country = `Region/Country`)
```
It is sometimes best to rename these columns to make them easier to refer to. 
`janitor::clean_names()` is very efficient for making easy-to-use names.

## Resources

- [stat545](https://stat545.com/dplyr-intro.html)
- [R for data science](https://r4ds.had.co.nz/) (Online copy of book)
- [`dplyr` cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

### Contributors {- .contributors .toc-ignore}

Richard J. Telford