---
execute: 
  warning: false
---

``` {r setup, include=FALSE}
source("R/setup.R")
```

# Combining plots {#sec-combining-plots}

We often need to combine plots to make a multipart figure in a manuscript or thesis.

First, check you actually need to combine plots and cannot use facets instead.
Typically, use facets if the x-axis variable is the same for all plots and the plots use the same geoms.

Plots can be combined using the `patchwork` package.

```{r patchwork-load, message=FALSE}
library(patchwork)
```

Start by making some plots

```{r some-plots}
p1 <- ggplot(penguins, aes(x = species, y = bill_length_mm)) +
  geom_boxplot()
p2 <- ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point()
p3 <- ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  facet_wrap(facets = vars(species))
p4 <- ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

The simplest way to use `patchwork` is to `+` to combine the plots.
`patchwork` will try to make the combined figure square.

```{r patchwork1}
#| fig-alt: Several plots joined together into a single figure with patchwork.
p1 + p2 + p3 + p4
```
With `|` (side by side) `\` (over under), you can have more control.

```{r patchwork2}
#| fig-alt: >
#|   Several plots joined together into a single figure with patchwork, 
#|   using | and / to control placement.
(p1 | p2 | p4) / p3
```

`plot_layout` gives you control over the relative size of the plot and whether the legends should be combined if possible.

```{r patchwork-layout}
#| fig-alt: >
#|   Several plots joined together into a single figure with patchwork,
#|   with control over the size of each plot and the legends combined.
(p1 | p2 | p4) / p3 +
  plot_layout(heights = c(2, 1), guides = "collect")
```

If you want to change all the plots in the combined figure, you can add a ggplot2 function with `&`.

```{r patchwork-ampersand}
#| fig-alt: >
#|   Several plots joined together into a single figure with patchwork. 
#|   All plots are modified to use a new theme.
(p1 | p2 | p4) / p3 +
  plot_layout(heights = c(2, 1), guides = "collect") &
  theme(panel.grid = element_blank())
```

You can also overlay figures, which might be useful to show an inset map.

```{r inset-map}
#| fig-alt: >
#|   Map of Svalbard, with inset map showing location of Svalbard in
#|   Northern Europe.
library(rnaturalearth)
library(sf)

# map data inset
europe <- ne_countries(scale = 110, continent = "Europe", returnclass = "sf")

# map data main map
svalbard <- ne_countries(scale = 50, country = "Norway", returnclass = "sf") |>
  st_crop(c(xmin = 0, xmax = 34, ymin = 76, ymax = 81))

# bounding box main map
bb <- svalbard |>
  st_bbox()

# inset map
euro_map <- ggplot() +
  geom_sf(data = europe) +
  # annotate bounding box
  annotate(
    geom = "rect",
    xmin = bb$xmin, xmax = bb$xmax, ymin = bb$ymin, ymax = bb$ymax,
    fill = "red", alpha = 0.3
  ) +
  coord_sf(xlim = c(-25, 40), y = c(50, 82)) +
  ggthemes::theme_map() +
  theme(
    # give white background and black border
    panel.background = element_rect(fill = "white", colour = "black"),
    plot.margin = margin() # remove margins
  )

# main map
svalbard_map <- ggplot() +
  geom_sf(data = svalbard)

# combined map
svalbard_map +
  inset_element(euro_map,
    left = 0.7,
    right = 0.99,
    top = 0.4,
    bottom = 0.01
  )
```

::: callout-note
## Further reading

[patchwork package](https://patchwork.data-imaginist.com/)
:::

::: {.column-margin}
### Contributors {.unlisted .unnumbered}

-   Richard Telford
:::
