``` {r setup, include=FALSE}
source("R/setup.R")
```

# Combining plots {#sec-combining-plots}

We often need to combine plots to make a multipart figure in a manuscript or thesis.

First, check you actually need to combine plots and cannot use facets instead.
Typically, use facets if the x-axis variable is the same for all plots and the plots use the same geoms.

Plots can be combined using the `patchwork` package.

```{r patchwork-load, message=FALSE}
library(patchwork)
```

Start by making some plots

```{r some-plots}
p1 <- ggplot(penguins, aes(x = species, y = bill_length_mm)) +
  geom_boxplot()
p2 <- ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point()
p3 <- ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  facet_wrap(~ species)
p4 <- ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

The simplest way to use `patchwork` is to `+` to combine the plots.
`patchwork` will try to make the combined figure square.

```{r patchwork1}
p1 + p2 + p3 + p4
```
With `|` (side by side) `\` (over under), you can have more control.

```{r patchwork2}
(p1 | p2  | p4) / p3
```

`plot_layout` gives you control over the relative size of the plot and whether the legends should be combined if possible.

```{r patchwork-layout}
(p1 | p2  | p4) / p3 + 
  plot_layout(heights = c(2, 1), guides = "collect")
```

If you want to change all the plots in the combined figure, you can add a ggplot2 function with `&`.

```{r patchwork-ampersand}
(p1 | p2  | p4) / p3 + 
  plot_layout(heights = c(2, 1), guides = "collect") &
  theme(panel.grid = element_blank())
```

You can also overlay figures, which might be useful to show an inset map.

```{r inset-map}
library(rnaturalearth)
library(sf)

# map data inset
europe <- ne_countries(scale = 110, continent = "Europe", returnclass = "sf")

# map data main map
svalbard <- ne_countries(scale = 50, country = "Norway", returnclass = "sf") |>  
   st_crop(c(xmin = 0, xmax = 34, ymin = 76, ymax = 81))

# bounding box main map
bb <- svalbard |> 
  st_bbox()

#inset map
euro_map <- ggplot() +
  geom_sf(data = europe) +
  # annotate bounding box
  annotate(geom = "rect", 
           xmin = bb$xmin, xmax = bb$xmax, ymin = bb$ymin, ymax = bb$ymax, 
           fill = "red", alpha = 0.3) +
  coord_sf(xlim = c(-25, 40), y = c(50, 82)) +
  ggthemes::theme_map() +
  theme(
  #give white background and black border
    panel.background = element_rect(fill = "white", colour = "black"), 
    plot.margin = margin() #remove margins
  )

# main map
svalbard_map <- ggplot() +
  geom_sf(data = svalbard)

# combined map
svalbard_map + 
  inset_element(euro_map, 
                left = 0.7, 
                right = 0.99, 
                top = 0.4, 
                bottom = 0.01)
```

::: callout-note
## Further reading

[patchwork package](https://patchwork.data-imaginist.com/)
:::

::: {.column-margin}
### Contributors {.unlisted .unnumbered}

-   Richard Telford
:::
