---
editor: visual
execute:
  message: false
---

# Making maps in R

```{r}
#| label: tidyverse
#| include: false
library(tidyverse)
library(patchwork)
source("R/setup.R")
```

Biologists need almost as many maps as geographers, perhaps to show where our field sites are, the range of a species, or the migration path of a bird.

Typically, the data we want to show is added to a base map.
These can be added to different types of base map

-   vector map showing the coastline, and perhaps political borders, rivers and other features

-   A map made of downloaded ties (similar to how Google maps works)

-   A raster (made of pixels)

::: callout-tip
## Terminology

Small scale vs large scale?
The terminology is confusing!

Map scales are given as a ratio.
A world map might have a scale of 1:100000000.
1 cm on the map represents 100000000 cm on the ground - a ratio of 1/100000000.
A small number hence a small-scale map.

Conversely, a map of a city might have a scale of 1:25000.
1 cm on the map represents 25000 cm on the ground - a ratio of 1/25000.
A large number hence a large-scale map.

Large-scale maps are usually high resolution and cover a small spatial extent.
:::

## Vector base maps

### `rnaturalearth`

The `rnaturalearth` package makes [Natural Earth](http://www.naturalearthdata.com/) data available.
Natural Earth [features include](https://www.naturalearthdata.com/features/) coastlines, rivers, bathymetry, political borders, roads and railways at scales 1:10m, 1:50m, and 1:110 million.
The `rnaturalearth` package includes coastline and country data at the small and medium scale.
A companion package `rnaturalearthhires` has the large scale (1:10 million) data.
Other datasets from Natural Earth can be downloaded directly from the [website](https://www.naturalearthdata.com/downloads/) or with `ne_download()`

```{r}
#| label: rnaturalearth
#| results: hide
library(rnaturalearth)
#install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev")
world <- ne_countries(scale = 110, returnclass = "sf") 
small_scale_map <- ggplot() +
  geom_sf(data = world) +
  coord_sf(xlim = c(-20, 50), ylim = c(33, 80)) +
  ggtitle("Europe")

europe <- ne_countries(scale = 50, returnclass = "sf", continent = "Europe") 
medium_scale_map <- ggplot() +
  geom_sf(data = europe) +
  coord_sf(xlim = c(5, 30), ylim = c(55, 71)) +
  ggtitle("Norden")

norway <- ne_countries(scale = 10, returnclass = "sf", country = "Norway") 

large_scale_map <- ggplot() +
  geom_sf(data = norway) +
  coord_sf(xlim = c(4, 9), ylim = c(59, 62)) +
  ggtitle("Vestland")

# combine maps with patchwork
library(patchwork)
small_scale_map + medium_scale_map + large_scale_map
```

`coord_sf()` is used to show only part of the map.

::: callout-tip
## `sf` and `sp` packages

`sf` and `sp` are both packages for geospatial data.
`sf` is the newer package that supports the "simple features" standard and is what I recommend, but you will find examples using the older `sp` package online.
Choose one, and don't mix them together.
:::

### `ggOceanMaps`

`ggOceanMaps` is, as the name suggests, focused on ocean map, with coastlines, bathymetry and also glaciers.
`ggOceanMaps` requires `ggOceanMapsData`, which needs to be installed separately

```{r}
#| label: install-ggOceanMapsData
#| eval: false
remotes::install_github("MikkoVihtakari/ggOceanMapsData")
```

Now `ggOceanMaps` is ready to use.

```{r}
#| label: ggOceanMaps
#| message: false
library(ggOceanMaps)
#limits are given longitude min/max, latitude min/max
basemap(limits = c(-30, 30, 50, 80),
        bathymetry = TRUE,
        glaciers = TRUE)
```

::: callout-note
## Exercise

Make a map of the Nordic Seas using either `rnaturalearth` or `ggOceanMaps`.
:::

### Other vector files

The maps in `rnaturalearth` and `ggOceanMaps` are good and the global and regional scale, but lack resolution for local scale maps, and may lack features we are interested in.

For such maps we need to find alternative resources.
These could be a shapefile, GeoJSON or GeoPackage file, all of which can be imported with `sf::st_read()`.

::: callout-tip
## Shapefiles

A "shapefile" is not one file but collection of several files in the same directory, only of which has the extension ".shp".
:::

Good sources of data for Norway include:

-   [GeoNorge](https://www.geonorge.no/)
-   [Miljødirektoratet](https://kartkatalog.miljodirektoratet.no/)

This is a map of the fylke of Norway

```{r}
#| label: fylker
library(sf)

fylker <- st_read("data/fylker2021.json")

ggplot(fylker) + 
  geom_sf()
```

::: callout-tip
## Coordinate reference systems

Most geographic data are given with latitude and longitude, but sometimes, especially for local-regional maps, the data are given as [Universal Transverse Mercator (UTM)](https://en.wikipedia.org/wiki/Universal_Transverse_Mercator_coordinate_system) coordinates instead.

UTM coordinates are a projection of the spherical Earth onto one of 60 flat surfaces.

Most modern latitude-longitude data will use the WGS84 geodetic standard.
Older data might use other standards.

You can find the coordinate system of a `sf` class object with `sf::st_crs()`.

```{r}
sf::st_crs(fylker)
```

This gives a lot of information, the most important is that the coordinate reference systems is UTM zone 33N.

If we need to change a coordinate reference systems, we can do that with `sf::st_transform()`.
You need to know the EPSG code of the target reference system.
The code for WGS84 is 4326.

```{r}
fylker2 <- sf::st_transform(fylker, crs = 4326)
```

`geom_sf()` will automatically transform coordinate systems (if they are specified).
:::

::: callout-note
## Exercise

Get a `sf` object from `rnaturalearth::ne_coastline()` and find out what coordinate reference system it uses.
:::

## Tiled basemaps

Tiled basemaps can be used with either `ggspatial` or `ggmap` packages. 
I would generally recommend using `ggspatial` as it is consistent with the other mapping tools used here.
If you use a tiled map background, you should attribute the source (e.g., "Copyright OpenStreetMap contributors" when using an OpenStreetMap-based tiles).

## `ggspatial`

We can add a tiled-basemap to a plot with `annotation_map_tile()`.
Here, we need to use `coord_sf()` to set the map extent and coordinate reference system as we have not added any `sf` layers with `geom_sf()`.
Downloaded tiles will be stored in the maps directory (which you may need to make first).

```{r}
#| fig-width: 8
#| message: false

ggplot() +
  annotation_map_tile(
    type = "osm", 
    cachedir = "maps/", 
    zoomin = -1) + # sets the zoom level relative to the default
  coord_sf(
    xlim = c(4, 8), 
    ylim = c(59, 62),
    crs = 4326)


```
Several different types of maps are available (see `rosm::osm.types()`) and more can be added.

```{r}
#| echo: false
#| fig-width: 8
#| message: false
# rosm::osm.types()
c("osm", "stamenwatercolor", "cartodark", "cartolight") |> 
  map(~{
    ggplot() +
      annotation_map_tile(
        type = .x, 
        cachedir = "maps/", 
        zoomin = -1) +
      coord_sf(
        xlim = c(5.24, 5.36), 
        ylim = c(60.37, 60.41),
        crs = 4326) +
      ggtitle(.x) +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(), 
            title = element_text(size = 10), 
            plot.margin = margin(1, 1, 1, 1))
}) |> 
  patchwork::wrap_plots()

```

## `ggmap`

The `ggmap` package lets you use Google Maps and other similar maps as a basemap.

::: callout-tip
## googlemaps

`ggmap` can use maps and satellite image from Google, but you need to [register](https://mapsplatform.google.com/) for an API key.
You shouldn't be charged unless you make a lot of maps (more than 20000 per month).
:::

```{r}
#| label: ggmap
#| message: false

library(ggmap)

bergen <- get_map(
  location = c(5.24, 60.37, 5.36, 60.41), #  left/bottom/right/top
  source = "stamen"           
)
ggmap(bergen)

```

Maps made with `ggmap` can appear cluttered with unnecessary information.

::: callout-note
## Exercise

Make a tiled map that shows your favourite holiday destination.
:::

## Raster basemaps

Rasters can be used to show maps of continuous data, for example, elevation or sea surface temperature.

### `terra`

The `terra` package can import raster images in several formats, including GeoTIFF.

::: callout-tip
## `terra` vs `raster` vs `stars` packages

The `terra` package is an update to the widely-used `raster` package.
It should be faster and easier to use.

`stars` is designed for spatio-temporal arrays.
There are some things it cannot do that `terra` can (and vice versa) but has better integration with `sf` and `ggplot2`.
:::

Rasters imported with `terra` are easy to plot with the base R `plot()` function, but if we want to use `ggplot()`, we have to first convert to an `sf` object.

```{r}
#| label: terra
#| message: false
library(terra)

# import digital elevation model
# data from https://topotools.cr.usgs.gov/gmted_viewer/viewer.htm
norway_dem <- rast("data/50N000E_20101117_gmted_med300.tif")

# crop to vestland and convert to sf object
vestland_extent <- ext(4.5, 9, 59, 62)
vestland_dem <- crop(norway_dem, vestland_extent) |> 
  as.points() |> 
  st_as_sf() 

#rename the data layer
names(vestland_dem)[1] <- "Elevation"

# plot

ggplot(vestland_dem) +
  geom_sf(aes(colour = Elevation))

```

If the raster is in UTM coordinates, it is possible to convert to a data.frame and then plot with `geom_raster()`.
Then `ggplot()` just treats it as regular plot and does not know that it is a map.

<!-- ```{r} -->

<!-- library(stars) -->

<!-- dem <- read_stars("data/50N000E_20101117_gmted_med300.tif") -->

<!-- #crop dem -->

<!-- bbox <- st_bbox(c(xmin = 4.5, xmax = 9, ymax = 62, ymin = 59), crs = st_crs(4326)) -->

<!-- dem_cropped <- st_crop(x = dem, y = bbox) -->

<!-- ggplot() + -->

<!--   geom_stars(data = dem_cropped) + -->

<!--   # geom_sf(data = st_crop(norway, bbox), fill = NA, colour = "white") + -->

<!--   labs(fill = "Elevation m")  -->

<!-- ``` -->

## Adding data to the basemap

After deciding what type of base map to draw, we can add the data we want to show with the map.
This can be

-   points, line, and polygons

-   Shaded political units (a cloropleth map)

-   A grid of values (raster)

### points/lines/polygons

Points lines and polygons can be added to the base map.
If the data are already a `sf` object they can be plotted with `geom_sf()`.

```{r}
#| label: second-sf
#| message: false
# aquaculture sites downloaded from Barentswatch.no/fiskinfo
aquaculture <- st_read("data/flate-ihht-akvakulturregisteret20220928.geojson")

# with rnaturalearth
ggplot() +
  geom_sf(data = europe) +
  geom_sf(data = aquaculture, colour = "red") +
  coord_sf(xlim = c(5, 30), ylim = c(55, 71))  

# with ggOceanMaps

basemap(limits = c(-30, 30, 50, 80)) +
  geom_sf(data = aquaculture, colour = "red")

# with ggmap
vestland <- get_map(
  location = c(4,  59, 8, 62),
  source = "stamen"           
)

# needs inherit.aes = FALSE
ggmap(vestland) + 
 geom_sf(data = aquaculture, colour = "red", inherit.aes = FALSE)
```

Alternatively, you can make/import a tibble with the geographic data and add them to the basemap with the relevant spatially aware geom.
So `geom_spatial_point()` rather than `geom_point()`

```{r}
#| label: geom-spatial-points
library(ggspatial)

random_points <- tibble(
  lat = runif(n = 10, 59, 62), 
  lon = runif(n = 10, 5, 8)) 



# ggOceanMaps

basemap(limits = c(-30, 30, 50, 80)) +
  geom_spatial_point(aes(x = lon, y = lat), 
             data = random_points,
             colour = "red") +
  geom_spatial_path(aes(x = lon, y = lat),
            data = random_points, 
            colour = "red")


# with ggmaps. just use `geom_point()` Don't need aes as it expects lat, lon
ggmap(vestland) + 
  geom_point(data = random_points) +
  geom_path(data = random_points)
```

`geom_spatial_point()` is assuming that the data are latitude-longitude coordinates.
It they are UTM, you will need to use the `crs` argument with the correct [EPSG code](https://epsg.io/?q=UTM).

::: callout-tip
## geom_path() vs geom_line()

`geom_path()` draws a line from the first point in the dataset to the second and so on.
This is useful for plotting on maps (and ordinations).

`geom_line()` draws a line from the left-most point to the next left-most point in the dataset.
This is useful for plotting timeseries.
:::

::: callout-tip
## Degrees minutes and seconds

For latitude-longitude data, we recommend using decimal degrees (Bergen is at 60.3807°N, 5.3323°E).
But archived data can be in all sorts of unhelpful formats, such as degrees minutes and seconds (Bergen is at 60° 22' 50.52" N 5° 19' 56.28" E).
If you get data like this, you need to convert it to decimal degrees.
The `parzer` package can help (it is like `lubridate` for latitude-longitude data).
For example:

```{r}
#| label: parzer

parzer::parse_lat("60° 22' 50.52''N")

```
:::

::: callout-note
## Exercise

Download and import the [data](https://setac.onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1002%2Fetc.5064&file=etc5064-sup-0002-Supplementary_dataset.xlsx) for [Andvik et al (2021)](https://doi.org/10.1002/etc.5064) on contaminants in orca.

Make a map to show the hexabromobenzene (HBB) concentrations in the blubber.
:::

### Cloropleth maps

Cloropleth maps are useful for plotting data that have been aggregated to a geographic unit (kommune, fylke, country etc).
A `sf` object is a special type of data frame that we can `filter()`, `mutate()` or `left_join()` to other data frames.
We need a tibble with the data that we can join to the `sf` object with the geographic units.
Here, I generate some random data and join it by Fylkesnummer (Fylkesnavn column in fylke is complicated), and plot it by setting `fill` in the `aes`.

```{r}
#| label: cloropleth
random_fylke <- tibble(
  Fylkesnummer = c("38", "42", "30", "11", "18", "15", "46", "03", "50", "54", "34"),
  random = runif(11))

fylker |> left_join(random_fylke) |> 
  ggplot() +
  geom_sf(aes(fill = random))
```

::: callout-note
## Exercise

With `rnaturalearth` data, make a world map that shows the population (column pop_est) of each country.
:::

### raster

Raster data, perhaps the output of a model, can also be added to maps, using code similar to that for plotting a raster basemap.

## Scalebars, north pointer etc

Scalebars and north pointers can be added with the `ggspatial` package or the `ggsn` package.
North points are not very useful if the map has gridlines as these already indicate north.
A scalebar can be useful, especially for large scale map.
On small scale maps, they can be inaccurate as the scale varies.

```{r}
#| label: annotations
# with rnaturalearth
ggplot() +
  geom_sf(data = norway) +
  coord_sf(xlim = c(4, 9), ylim = c(59, 62)) +
  annotation_scale(location = "br") # br = bottom right
```

## Hints for maps

Keep it simple.
Remove unnecessary features (do you really need to show the bathymetry?) and use appropriate scale data for the base map (too high resolution takes a long time to plot and can look worse).

Use facets as necessary (different species, different years etc).

If you need multiple colour scales, the `ggnewscale` package can help.
Use inset maps (@sec-combining-plots) to show your location in context.

### Projections

The Earth is an oblate sphere and needs projecting to plot in two dimensions.
This inevitably leads to distortions, especially for maps with a large extent.
[Different projections](https://proj.org/operations/projections/index.html) have different properties and may be suitable for different purposes or regions.
Map projections can be set using `coord_sf()`.

```{r}
#| label: projections
#| results: hide
world <- ne_countries(scale = "medium", returnclass = "sf")

default <- ggplot(world) + geom_sf()

mollweide <- ggplot(world) + 
  geom_sf() + 
  coord_sf(crs = "+proj=moll")


lambert <-  ggplot(world) + 
  geom_sf() + 
  coord_sf(crs = "+proj=laea")
 
default/mollweide/lambert
```

Many of the projections have optional parameters to change, for example, the projection centre.

::: callout-note
## Exercise

Change the projection of one of the small-scale maps you made previously.
:::

<!-- ## More map types -->

<!-- ### leaflets -->

<!-- ### Interactive maps -->

<!-- projections -->

<!-- ### Cartograms -->

<!-- A cartogram is a map with a distorted geometry to illustrate a variable according to the area of a region. -->

<!-- These can be made with the `cartogram` package. -->

::: callout-note
## Further reading

-   [Spatial Data Science](https://r-spatial.org/book/)
-   [Visualizing geospatial data](https://clauswilke.com/dataviz/geospatial-data.html)
:::

::: column-margin
### Contributors {.unlisted .unnumbered}

-   Richard Telford
:::
