---
fig-height: 2
filters:
  - webr
webr:  
  show-startup-message: true    # Display status of webR initialization
  packages: ['ggplot2', 'tibble'] # Pre-install dependency
  cell-options:
    fig-height: 2
---

# Mathematical notation for plots

``` {r}
#| label: setup
#| include: false
source("R/setup.R")
library(gt)
```

```{webr-r}
#| label: webr-setup
#| context: setup
theme_set(theme_bw())

dat <- tibble(x = 1:5, y = x^2)
```


## Introduction

When making a publication ready plot, we often need to use special characters and mathematical notation  for the axes and other labels.

 - $TP~concentration~\mu g~l^{-1}$
 - $Conductivity~\mu S~cm^{-1}$
 - $\sqrt{Standardised~residuals}$
 - $\delta^{15}N$
 
The functions `expression()` and `bquote()` are powerful tools for annotating figures with mathematical notation in R. 
This functionality is not immediately obvious from their respective help files. 
`demo(plotmath)` and `?plotmath` show the huge potential of `expression()`, but do not have many examples relevant to biologists with graphs to label.

Expressions are comprised of symbols. 
The expression $TP ~ concentration ~\mu g~l^{-1}$ includes the symbols $TP$, $concentration$, $\mu$, $g$, $l$, and $^{-1}$. 
In this chapter, you will learn how to generate these symbols and combine them with `expression()`.

The key to getting expressions to work is to start with something simple, test it and edit until it works, then add the next symbol and repeat. 
This is the approach we are going to use here.

## `expression()`

We are going to use `expression()` with `labs()`, but the same expressions can be used with some other functions that write text on a plot (including base plots). 
Other functions take a slightly different approach (see section on parsing below).

The simplest use of expression is take a symbol that is a single word or number.

```{r}
#| label: expression1
# sample data
dat <- tibble(x = 1:5, y = x^2)

# base plot
g <- ggplot(dat, aes(x = x, y = y)) +
  geom_point() 

g +  labs(x = expression(TP))

```

Notice that `expression()` does not require quote marks except to enclose spaces or punctuation. 
This first example of  `expression()` is entirely pointless except as a starting point for something more complex.

If we want two symbols with a space between them, we need to separate them with a tilde `~`. 
If we don't want a space, we can separate symbols with `*` (if you actually want a `*`, use `%*%`).

```{r}
#| label: expression2
g + labs(x = expression(TP~concentration))
```

Some symbols have special meanings, for example, to get a Greek letter, write its name in lower case. 
Capitalise the first letter to get an upper-case Greek letter. 
If for some reason you want to have "beta" written on your plot, it must be in quotes.

```{r}
#| label: expression3
g + labs(x = expression(TP~concentration~mu*g))
```

Subscript or superscript can be added to a string using `^` and `[]` notation respectively. 
If you want more than one symbol to be in superscript, you need to enclose them in braces `{}`.

```{r}
#| label: expression4
g + labs(x = expression(TP~concentration~mu*g~l^-1))
```

To start an `expression()` with a superscript (or subscript), put `""` first.

```{r, eval = FALSE}
#| label: expression5
g + labs(x = expression(""^14*C~years~BP))
```



```{r}
#| label: tbl-common-notation
#| tbl-caption: Some common notation
#| echo: false

tribble(
~What, ~How, ~Output,         
"Lower-case Greek letters", "delta", "$$\\delta$$",
"Upper-Case Greek Letters", "Delta", "$$\\Delta$$",
"Subscript", "CO[2]", "$$CO_{2}$$",
"Superscript", "m^-2", "$$m^{-2}$$",
"Italics", "italic(Navicula)", "$$Navicula$$"
) |> 
  gt()
```


If you already have the expression as a string, you can use `str2expression()` (or `parse()` with the `text` argument) to convert it into an expression.

```{r}
#| label: str2expression

lab <- "Wind~speed~m~s^-1"
str2expression(lab)

```


::: callout-note
## Exercise

1) Make the x-axis label read $Area~m^2$. 
The data are already loaded.

::: {.panel-tabset}
## Code
```{webr-r}
ggplot(dat, aes(x = x, y = y)) +
  geom_point() + 
  labs(x = ___)
```
## Hint
```{r expression1-solution}
#| eval = FALSE

ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  labs(x = expression(Area~m^2))
```
:::

2) Make the x-axis label read $Flux~g~m^{-2}~yr^{-1}$.

::: {.panel-tabset}
## Code
```{webr-r}
ggplot(dat, aes(x = x, y = y)) +
  geom_point() + 
  labs(x = ___)
```

## Hint
```{r expression2-solution}
#| eval = FALSE

ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  labs(x = expression(Flux~g~m^-2~yr^-1))
```
:::

3) Make the x-axis label read $\delta^{15}N$.

::: {.panel-tabset}
## Code
```{webr-r}
ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  labs(x = ___)
```

## Hint
```{r expression3-solution}
#| eval = FALSE

ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  labs(x = expression(delta^15*N))
```
:::

4) Make the x-axis label read _Aulacoseira granulata_ var. _angustissima_.

::: {.panel-tabset}
## Code
```{webr-r}
ggplot(dat, aes(x = x, y = y)) +
  geom_point() + 
  labs(x = ___)
```

## Hint

```{r expression4-solution}
#| eval = FALSE

ggplot(dat, aes(x = x, y = y)) +
  geom_point() +
  labs(x = expression(italic(Aulacoseira~granulata)~var.~italic(angustissima)))
```
:::
:::

## Parsing

Some functions such as `geom_text()`, `facet_wrap()` and `scale_x_discrete()` do not accept expressions directly. 
Instead, they parse a string and convert it to an expression. 
Each function does this a different way!
Usually it helps to use `mutate()` to add columns with the parsable labels if necessary.
Here is a minimal example that adds subscripts and superscripts to text, axis labels and facet strips. 

```{r}
#| label: parsed
#| fig-cap: Composition of Lake Aranguadi, Ethiopia (Wood and Talling 1988)
#| fig-height: 5
#| fig-width: 7

# collate data
dat <- tibble(
  species = c("Na+", "K+", "Ca2+", "Mg2+", "Cl-", "SO42-", "HCO3-", "CO32-"),
  conc = c(368, 59, 7, 70, 142, 17, 1231, 85),
  mol_mass = c(
    22.99, 39.10, 40.08, 24.31, 35.45, 96.06, 61.02, 60.01
  ),
  moles = conc/mol_mass
) |> 
  select(-mol_mass) |>
  pivot_longer(cols = c(conc, moles), names_to = "unit", values_to = "value")

# add a column that can be parsed as formatted labels
dat <- dat |> 
  mutate(labels = case_match(species,
    "Na+" ~ "Na^'+'",
    "K+" ~ "K^'+'",
    "Ca2+" ~ "Ca^{2*'+'}",
    "Mg2+" ~ "Mg^{2*'+'}",
    "Cl-" ~ "Cl^'-'",
    "SO42-" ~ "SO[4]^{2*'-'}",
    "HCO3-" ~ "HCO[3]^'-'",
    "CO32-" ~ "CO[3]^{2*'-'}"
  )) |> 
  # convert into factor to ensure plotting order
  mutate(labels = factor(labels)) |> 
  # make units column parsable as expression
  mutate(unit = case_match(unit,
    "conc" ~ "mg %.% l^{-1}",
    "moles" ~ "mmol %.% l^{-1}"
  )) 



ggplot(dat, aes(x = labels, y = value)) +
  geom_col() +
  # use parse argument in geom_text()
  geom_text(aes(label = labels), 
            parse = TRUE, 
            vjust = -0.1 # move text up slightly
            ) +  
  # use parse function to make expressions for scale_x_discrete()
  scale_x_discrete(labels = parse(text = levels(dat$labels))) +
  # use labeller argument in facet_wrap to parse strip labels
  facet_wrap(vars(unit), labeller = "label_parsed", scale = "free_y") +
  labs(x = "Ion", y = "Concentration") 

```





## Troubleshooting

Expressions, are quite tricky to get right. 

Here are some examples of errors and how to interpret them.

```{r}
#| error: true

expression(Wind speed m%.%s^-1)
```

You cannot have spaces in expressions unless you quote them.
You can use a `~` instead of a space.

```{r}
#| error: true

expression(Wind~speed~m%.%s^-1)
# or
expression("Wind speed m"%.%s^-1)
```


```{r}
#| error: true

expression(Ca^2+)
```
The `+` is a mathematical symbol that expects another symbol after it. 
To make it work here, quote the `+` or the `2+`.

```{r}
#| error: true
expression(Ca^2*'+')
# or
expression(Ca^"2+")

```


```{r}
#| error: true
expression(italic(Betula~nana)~%)
```

The `%` is a special character, so it need to be quoted. 

```{r}
expression(italic(Betula~nana)~'%')
```

::: callout-note
## Exercise

Fix the following broken expressions

::: {.panel-tabset}
## Code
```{webr-r}
expression(Wet mass g m^-2)
```
## Hint

No spaces allowed in expressions. Quote or replace spaces with `~`.  
:::

::: {.panel-tabset}
## Code
```{webr-r}
ggplot() + 
  labs(x = expression(italics(Dyras~octopetala)~"%"))
```
## Hint

Be careful with spelling! italic not italics.

:::
:::

## Other tips

- `bquote()` (or `label_bquote()` for facet labels) can be useful if you want to include an R object in the expression. 
- `demo(plotmath)` and `?plotmath` are very useful once you have understood the basics.
- Some symbols are not defined by `plotmath` but can be included using [unicode](http://www.unicode.org/charts/). 
For example a per mil symbol (‰) - `\u2030` 
```{r}
#| eval: false
expression(delta^15*N~('\u2030'))
```

- non-alphanumeric symbols (e.g. %) may need quoting
- [LaTeX equations](https://biostats-r.github.io/biostats/rmarkdown/figures-tables-and-equations.html#equations) can be converted to expressions with `latex2exp::TeX`.



If you understand these examples, you should be able to use the remainder of the functionality demonstrated by `demo(plotmath)` and at `?plotmath`.




::: {.column-margin}
### Contributors {.unlisted .unnumbered}

-   Richard Telford
:::

