``` {r setup, include=FALSE}
source("R/setup.R")
library(patchwork)
```

# ggplot2 helper packages

`ggplot2` is the heart of a vast network of packages that enhance it.
This a an annotated list of packages that I find useful.


## `directlabels`

The `directlabels` package helps add labels to plots as a alternative to including a legend.

```{r}
#| label: fig-directlabels
#| fig-cap: Reaction time against day of sleep deprivation experiment. Subject ID are on the right of the plot.
#| fig-alt: Line graph showing reaction time against day of sleep deprivation experiment. Subject ID are on the right of the plot next to their respective line and with the same colour as the line.


library(directlabels)
p <- ggplot(lme4::sleepstudy, aes(x = Days, y = Reaction, colour = Subject)) +
  geom_line() +
  labs(y = "Reaction time ms")
direct.label(p, method = "last.qp") # last - go after plot, qp - don't overlap
```



## `GGally`

Among other functions, `GGally` includes `ggpairs()` which makes a matrix of plots, plotting each variable in a dataset against every other variable. 
This is a useful exploratory method.
Different combinations of variable types (continuous vs categorical) get different types of plot.

```{r}
#| label: fig-ggally
#| fig-cap: ggpairs() applied to the penguins dataset
#| fig-alt: matrix of plots that show each variable in the penguins dataset plotted against every other variable.
#| results: hide
#| message: false
#| warning: false


library(GGally)
ggpairs(penguins, mapping = aes(colour = species))
```


## `gganimate`

It won't animate a Disney film, but `gganimate` will help bring your data to life.

```{r}
#| label: fig-gganimate
#| fig-cap: Relationship between life expectancy and GDP per capita over time.
#| fig-alt: Animation of the relationship between life expectancy and GDP per capita over time using the gapminder dataset.
library(gganimate)
library(gapminder) # world health data

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, colour = continent)) +
  geom_point(alpha = 0.7) +
  scale_colour_manual(values = continent_colors) +
  scale_size(range = c(2, 12), guide = "none") +
  scale_x_log10() +
  theme(legend.position = "inside", legend.position.inside = c(0.99, 0.01), legend.justification = c(1, 0)) +
  labs(title = "Year: {frame_time}", x = "GDP per capita US$", y = "Life expectancy yr", colour = "Continent") +
  transition_time(year) # animates ggplot
```
This is not so useful for traditional journals (except perhaps in supplementary material), but some new journal may accept animations and other video. 
It is good for presentations and websites.

## `ggbeeswarm`

`ggbeeswarm` includes `geom_quasirandom()` and `geom_beeswarm()`  which are alternatives to `geom_jitter()` for plotting data with a continuous dependent variable and categorical independent variable.

```{r}
#| label: fig-ggbeeswarm
#| warning: false
#| echo: false
#| fig-cap: "`geom_jitter()` and alternatives from the `ggbeeswarm` package"
#| fig-alt: "Plots of penguin body mass by species made using `geom_jitter()` and alternatives from the `ggbeeswarm` package"

library(ggbeeswarm)
p <- ggplot(palmerpenguins::penguins, aes(x = species, y = body_mass_g)) +
  labs(x = "Species", y = "Body mass, g")

c("geom_jitter()", "geom_beeswarm(cex = 2.5)", "geom_quasirandom()") |>
  map(\(geom){
    p + eval(parse(text = geom)) +
      ggtitle(paste0(str_extract(geom, "geom_[a-z]*"), "()"))
  }) |>
  patchwork::wrap_plots()
```

## `ggfortify`

The `ggfortify` package includes `autoplot()` functions for several types of objects, including (generalised) linear models, time series, and survival analyses.


## `gghighlight`

The `gghighlight` package helps highlight part of the data see @sec-highlight for an example.

## `ggiraph`

The `ggiraph` lets you make interactive plots. 
Use `geom_*_interactive()` to get tooltips.

```{r}
#| label: fig-ggiraph
#| fig-cap: Interactive plot of life expectancy against GDP per capita in 1972. Hover mouse over a point to see which country it represents.
#| fig-alt: Interactive plot of life expectancy against GDP per capita from the gapminder dataset. Hovering the mouse over a point brings up a tool tip that shows which country it represents.

library(ggiraph)
# remove ' from cote d'Ivoire
gapminder <- gapminder |>
  mutate(country2 = str_replace(country, "'", "")) |>
  filter(year == 1972)

p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, colour = continent, size = pop)) +
  geom_point_interactive(aes(tooltip = country, data_id = country2, alpha = 0.7)) +
  scale_colour_manual(values = continent_colors) +
  scale_size(range = c(2, 12), guide = "none") +
  scale_x_log10() +
  theme(legend.position = "none") +
  labs(x = "GDP per capita US$", y = "Life expectancy yr", colour = "Continent")


girafe(
  ggobj = p,
  options = list(opts_hover(css = "fill: red"))
)
```


## `ggpattern`

Bored with solid fills? 
`ggpattern` can give you textured fills with `geom_*_pattern()` where the `*` is any of the geoms that take a fill argument.

```{r}
#| label: fig-ggpattern
#| fig-cap: Bar chart made with `geom_bar_pattern()` showing three different pattern types.
#| fig-alt: Barchart of the penguin data showing three different pattern types and a larger legend than default so the patterns show clearly.


library(ggpattern)
ggplot(penguins, aes(x = species)) +
  geom_bar_pattern(
    aes(
      pattern_fill = species,
      pattern_type = species
    ),
    pattern = "polygon_tiling"
  ) +
  scale_pattern_type_manual(values = c("hexagonal", "rhombille", "pythagorean")) +
  theme(legend.key.size = unit(1.5, "cm")) + # larger legend
  labs(
    title    = "ggpattern::geom_density_pattern()",
    subtitle = "pattern = 'polygon_tiling'"
  )
```


## `ggraph`

The `ggraph` package lets you plot network graphs, including dendrograms.

## `ggrepel`

If you want to label points on a plot, but the labels overlap, `ggrepel` can help.

```{r}
#| label: fig-ggrepel
#| fig-cap: "`geom_text()` and `geom_text_repel()`"
#| fig-alt: Plots of milage against weight for cars. Left plot labels points with geom_text(), with a lot of overlapping text. Right plot uses geom_text_repel() to avoid overlapping.
#| warning: false
library(ggrepel)

# base plot
p <- ggplot(mtcars, aes(x = wt, y = mpg, label = rownames(mtcars))) +
  geom_point() +
  labs(x = "Car weight", y = "Milage mgp")

# add labels (use geom_label/geom_label_repel to add background rectangle to text)
p1 <- p + geom_text() + ggtitle("geom_text()")
p2 <- p + geom_text_repel() + ggtitle("geom_text_repel()")

# combine with patchwork
p1 + p2
```


## `ggspatial`

The `ggspatial` package helps plot maps. See @sec-map-making for some examples using this package.

## `ggtext`

The `ggtext` package lets you use markdown wherever there is text in a plot. 
This lets you add formatting (colour, bold, italics etc) to text (with `geom_richtext()`) or axis labels and titles by using `element_markdown()` in the theme.
This can be an alternative to using a legend.


```{r}
#| label: fig-ggtext
#| warning: false
#| fig-cap: Coloured text in the title indicates which species is which, so a legend is not required.
#| fig-alt: Plot of penguins flipper-length against body-mass. Plot title has coloured text to indicate the species, so a legend is not required.

library(ggtext)

ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", aes(fill = after_scale(colour)), alpha = 0.5) +
  scale_color_manual(
    values = c(Adelie = "#0072B2", Chinstrap = "#D55E00", Gentoo = "#55aa55"),
    guide = "none" # one way to drop legends
  ) +
  labs(
    x = "Body mass g",
    y = "Flipper length mm",
    # title has formatting
    title = "<span style = 'font-size:14pt; font-family:Helvetica;'>Flipper-body mass relationship similar between penguin species</span><br>
<span style = 'color:#0072B2;'>Adelie</span> and <span style = 'color:#D55E00;'>Chinstrap</span>
are smaller than <span style = 'color:#55aa55;'>Gentoo</span> penguins,
but the flipper length-body mass relationship is similar."
  ) +
  theme(
    text = element_text(family = "Times"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 10, lineheight = 1.2)
  )
```


## `ggvegan`

`ggvegan` (install from [GitHub](https://github.com/gavinsimpson/ggvegan)) has `autoplot()` functions to plot ordinations and other objects from the `vegan` package.


## `patchwork`

`patchwork` is used to combine plots into a single figure. 
See @sec-combining-plots for information on how to use this package.


::: {.column-margin}
### Contributors {.unlisted .unnumbered}

-   Richard Telford
:::
