---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")

# title: Creating a 2D density plot
# author: jonathan
# revised: 17/3/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/2D-density.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/2D-density.Rmd")
library(ggplot2)
library(gridExtra)
# variable 1
values1 <- rnorm(4000, mean = 50, sd = 10)
# variable 2
values2 <- rnorm(4000, mean = 5, sd = 1)
```
When a scatter plot becomes so dense that the symbols overlap massively, the plot lacks clarity and loses its purpose. This is called overplotting. In such a case, there are a few alternatives, one of them being a 2D density plot (the 2D version of the density plot described [HERE](https://biostats.w.uib.no/creating-a-density-plot/){target="_blank"}). The 2D density plot brings an extra dimension to the scatter plot by allowing to visualize the density of observations with a colored scale.

Let's take the following example where the variable to be plotted along the X-axis (`values1`) and the variable to be plotted along the Y-axis (`values2`) both contain 4000 points (!). Here is the dataframe:
```{r dataframe}
# dataframe
df <- data.frame(values1, values2)
str(df)
```
<br/><br/>

If we attempt to draw a regular scatter plot with `geom_point()`, we obtain this unreadable figure:
```{r scatterplot}
ggplot(df, aes(values1, values2)) +
  geom_point()
```
<br/><br/>

In such a case of overplotting, it is logical to switch to a different plot type.
To draw a 2D density plot, we will either use `stat_density_2d()` or `geom_density_2d()` with a series of necessary arguments. Here is an example with `stat_density_2d()`:
```{r 2D stat}
ggplot(df, aes(values1, values2)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon")
```

And here is the same example with `geom_density_2d()`:
```{r 2D geom}
ggplot(df, aes(values1, values2)) +
  geom_density_2d(aes(color = ..level..))
```
A color bar guide named `level` automatically appears to the right thanks to the parameter `..level..`. It shows the different shades of blue (color by default) used in the figure to help you define approximately the density level in the figure.
<br/><br/>

If you feel like changing the colors used in the color bar guide, you can use `scale_fill_gradient()`, `scale_color_gradient()`, `scale_fill_viridis_c()` or `scale_color_viridis_c()`. 
```{r 2Dviridis, eval=FALSE}
ggplot(df, aes(values1, values2)) + # left plot with stat_density_2d
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  scale_fill_viridis_c()
ggplot(df, aes(values1, values2)) + # right plot with geom_density_2d
  geom_density_2d(aes(color = ..level..)) +
  scale_color_viridis_c()
```
```{r 2Dviridis2, echo=FALSE, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values1, values2)) + # left plot with stat_density_2d
    stat_density_2d(aes(fill = ..level..), geom = "polygon") +
    scale_fill_viridis_c(),
  ggplot(df, aes(values1, values2)) + # right plot with geom_density_2d
    geom_density_2d(aes(color = ..level..)) +
    scale_color_viridis_c(),
  ncol = 2
)
```
Read [this page](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"} to learn more about *color palettes*.

<br/><br/>

Note that it is possible to highlight the borders between the levels in the `stat_density_2d()` plot by adding contours. To do so, you simply have to add the arguments `color=` to the function `stat_density_2d()`. You may also modulate the thickness of these contour lines by adding `size=`:
```{r borders}
ggplot(df, aes(values1, values2)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "red", size = 1.5) +
  scale_fill_viridis_c()
```
<br/><br/>

The same applies to lines in the `geom_density_2d()` plot:
```{r lines}
ggplot(df, aes(values1, values2)) +
  geom_density_2d(aes(color = ..level..), size = 1.5) +
  scale_color_viridis_c()
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
