---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a density plot
# author: jonathan
# revised: 17/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/density-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/density-plot.Rmd")
library(ggplot2)
library(gridExtra)
# ID
ID <- 1:200
# sample data
values <- rnorm(200, mean=65, sd=15)
```
A density plot, like a histogram of frequency, helps visualize the distribution of a sample. This time, instead of bars, a single curve is drawn to represent that distribution. 

Let's use `ggplot()` to draw the density plot for a data set generated by `rnorm()` (read more [here](https://biostats.w.uib.no/4-generating-random-series/){target="_blank"} about `rnorm()`). Here is the dataframe:
```{r dataframe}
# dataframe
df <- data.frame(ID, values)
str(df)
```
<br/><br/>

We first map the data from the variable `values` by typing `ggplot(df, aes(values))` and then use `geom_density()` to draw the plot:
```{r plot}
ggplot(df, aes(values)) + 
  geom_density()
```

To realize how similar a density plot and a histogram are, we can put them next to each other:
```{r plot2a, eval=FALSE}
ggplot(df, aes(values)) +            # histogram
  geom_histogram(bins = 30)
ggplot(df, aes(values)) +            # density plot
  geom_density()
```
```{r plot2b, echo=FALSE, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values)) +          # histogram
    geom_histogram(bins = 30),
  ggplot(df, aes(values)) +          # density plot
    geom_density(),
  ncol=2
)
```

There isn't much you can do to improve the look of a density plot, but you can alway add some colors with `fill=` and `color=`, or make the line thicker with `size=`:
```{r color}
ggplot(df, aes(values)) +  
  geom_density(fill="red", color="blue", size=1) 
```
<br/><br/>
If this is a bit to bright for your eyes, you may add some transparency with `alpha=`:
```{r transp}
ggplot(df, aes(values)) +  
  geom_density(fill="red", alpha=.2) 
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
