```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: standard deviation
# author: jonathan
# revised: 8/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Descr/sd.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Descr/sd.Rmd")
```

The **standard deviation (sd or $\sigma$)** is the square root of the [variance](https://biostats.w.uib.no/4-variance/){target="_blank"}. Unlike the variance, the standard deviation has the same unit as the mean, the median or any entry in the dataset. This makes it very useful as it can be used in combination with the mean to define how values are spread around the mean. Often, you will find this expressed as `mean +/- sd`.

In R, the standard deviation is obtained via the function `sd()`. Note that you will obtain the same result by calculating “manually” the square root of the variance `sqrt(var())`:

```{r sd}
my.dataset <- c(1,2,3,4,5,6,7,8,9,10)
sd(my.dataset)
sqrt(var(my.dataset))
```
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
