# Package Contents

```{r setup, include=FALSE, eval = TRUE}
library("tidyverse")
data("penguins", package = "palmerpenguins")
```

## Basic workflow

1)  Write tests (@sec-tests)
2)  Write functions (@sec-functions) and documentation (@sec-documentation-with-roxygen2)
3)  Check package (@sec-checking)
4)  Fix errors goto 3
5)  Test drive with `devtools::load_all()`
6)  Build and install (@sec-build-and-install)

Because there are usually lots of errors to fix, it is sensible to build the package slowly, testing it frequently.

## Tests {#sec-tests}

You need to write tests for your package to ensure your functions do what they are supposed to do.
They protect you against breaking your package when you edit your code.
Tests are run when the package is checked.

You can, ideally should, write tests before you write your functions.
We can use the `testthat` package for the tests.

Set up the testing infrastructure with

```{r use-testthat}
usethis::use_testthat()
```

Now set up a test file for one or more related functions with `use_test()`.

```{r use-test}
usethis::use_test(name = "import")
```

This will create and open a file called `test-import.R` which looks like

```{r}
test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

The first argument of `test_that()` is a description, second argument is an expression which contains the test.
More complex tests might need some additional set-up code in the expression.
There are several `expect_*()` functions to test different aspects of the function, including that errors and warnings are thrown as expected.
Each `test_that()` call can test multiple expectations.
You can have multiple `test_that()` calls per file.

::: callout-tip
## tests for import functions

If you are writing a package that imports data from files, you need to test this.
You can save files for testing in the "tests/testthat" directory. 
When you load them you will need to `testhat::test_path()` so the file can be found.

For example, if you wanted to test if function `my_import()` could correctly import a file ("tests/testthat/testdata.csv") with the correct number of rows, we could use

```{r}
#| eval: false
test_that("import works", {
  expect_equal(nrow(my_import(test_path("testdata.csv"))), 10)
})

```

:::

Look for examples of tests on GitHub if you need inspiration.

Functions are much easier to test if the functions do one job.
This is also best practice when writing functions.
For example, if you were writing a package to import, process, and plot logger data, you would make at least three functions to do this, not one function that does everything.

## Functions {#sec-functions}

Functions are made with the keyword function, can have one or more arguments separated by commas, and needs assigning to a name.

```{r function}
my_function <- function(arg1 = 1, arg2){
  arg1 * arg2
}

my_function(3, 4)
```

Functions need to be saved in `R/`.
Related functions can be saved in the same file.

### Well behaved functions

Try not to alter the state of the users R session.
Don't include calls to `library()` or `require()` in functions (see @sec-importing-exporting-functions).
If you need to change the state, then revert it with when the function finishes.
You can do this with `withr::defer()`.
This is better behaved than the base R equivalent `on.exit()`.
Even if the function throws an error, the state of `options` will be reverted to its original state.

```{r on-exit}
my_function <- function(x){
  op <- options(digits = 1) # set options
  withr::defer(options(op)) #next line
  print(x)
}
```

### Data validation - expect the unexpected.

If you are going to release your package, you need to try to make it idiot proof.
Assume users will make mistakes with their data input.
Use code to validate that the data are correct, or else throw an error.
`if` statements and `stop()` are useful here.

```{r eval = TRUE, error = TRUE}
my_function <- function(arg1 = 1, arg2){
  if (!is.numeric(arg1) || !is.numeric(arg2)) {
    stop("Arguments arg1 and arg2 must be numeric")
  }
  arg1 * arg2
}

my_function(3, "4")
```

### S3 classes in R

When you use a generic function in R such as `plot()`, `print()`, `autoplot()` or `summary()`, what happens is that the class of the object in determined and dispatched to the appropriate function, which will have the name of the generic followed by the name of the class, separated by a dot.

So a call to `plot()` with an object of class `cca` will be dispatched to `plot.cca`.

You can find out the class of an object with `class()`

The class of an object can be set with `class()`.

```{r class}
my_function <- function(){
  result <- complex_logic()
  class(result) <- "my_class"
  result
  }
```

If the object already has a class and you want to keep it, you need something like

```{r class2}
  class(result) <- c("my_class", class(result))
```

To make a `print()` or `plot()` method for `my_class`, we simply make a function called `print.my_class` or `plot.my_class`.
The method will be declared automatically when the documentation (@sec-documentation-with-roxygen2) is made.

::: callout-tip
## Going further with classes

-   Make your [own generics](https://adv-r.hadley.nz/s3.html#s3-methods)
-   Use the more formal [S4 or R6 class](https://adv-r.hadley.nz/oo.html) systems
:::


### Ellipses ...

Ellipses can be used in two ways when writing functions.

The first is to pass unknown arguments to a second function (e.g., `plot.cca()`).

If we make a `plot.my_class()` function we can use the ellipses so we don't need to specify all the possible arguments in plot.

```{r ellipses1}
plot.my_class <- function(obj, ...){
  #logic to prepare data for plotting
  x <- obj$x
  y <-obj$y
  plot(x, y, ...)
}
```

Now all of arguments to `plot.default()` can be used.

The second way to use ellipses is when there are a variable number of arguments.
We can capture the `...` using `list()`, and then process it further.

```{r dot-list, eval = TRUE}
dot_to_list <- function(...){
  list(...)
}

dot_to_list(1, 2, "c")
```

### Using `dplyr`, `ggplot2` etc {#sec-tidyverse}

Tidyverse packages such as `dplyr` and `ggplot2` which use Non-Standard Evaluation (NSE) are great for using in a script but a little challenging to use in functions.

We cannot just do something like this as we get an error

```{r tidyselect, error = TRUE, eval = TRUE}
my_select <- function(data, col){
  select(data, col)
}

my_select(penguins, col = species)
```

One solution is to use the curly-curly notation

```{r tidyselect-working, error = TRUE, eval = TRUE}
#select
my_select2 <- function(data, col){
  select(data, {{col}})
}

my_select2(penguins, col = species)

# filter
my_filter <- function(data, col, `%test%`, value){
  filter(data, {{col}} %test% value)
}

my_filter(penguins, col = species, `%test%` = `==`, value = "Adelie")

#mutate Note the := operator.
my_mutate <- function(data, col1, col2, sum_col){
  mutate(data, {{sum_col}} := {{col1}} + {{col2}})
}

my_mutate(penguins, bill_length_mm, bill_depth_mm, sum_col = bill)

#ggplot
my_plot <- function(data, x, y){
  ggplot(data, aes(x = {{x}}, y = {{y}})) + 
    geom_point()
}

my_plot(penguins, x = bill_length_mm, y = body_mass_g)
```

Another is to have the arguments as strings and use the `.data` pronoun from `rlang` (don't forget to import `rlang` (see @sec-importing-exporting-functions).

```{r .data, eval = TRUE}
my_select3 <- function(data, col){
  select(data, .data[[col]])
}

my_select3(penguins, col = "species")
```

NSE can also be a problem for functions that don't take column names as arguments.

```{r NSE}
process_penguin_type_data <- function(data){
  data %>% 
    group_by(species) %>% 
    summarise(bill = mean(bill_length_mm))
} 

```

The this function should work, but will generate a note in the checking stage.
as the check does not recognise `species` or `bill_length_mm`.

    > checking R code for possible problems ... NOTE
      process_penguin_type_data: no visible binding for global variable
        ‘species’
      process_penguin_type_data: no visible binding for global variable
        ‘bill_length_mm’
      Undefined global functions or variables:
        bill_length_mm species

To fix this, either declare these variables as global with this line added to the file (outside the function).

```{r globalvariables}
utils::globalVariables(c('bill_length_mm', 'species'))
```

Or, perhaps better, use the `.data` pronoun.

```{r NSE2}
process_penguin_type_data <- function(data){
  data %>% 
    group_by(.data$species) %>% 
    summarise(bill = mean(.data$bill_length_mm))
} 
```

## Documentation with Roxygen2 {#sec-documentation-with-roxygen2}

The documentation for the package lives in `man/` in .Rd files.
The files are written in a LaTeX like language, that is quite hard to get to right.
Fortunately, the [`roxygen2` package](https://roxygen2.r-lib.org/index.html) takes most of the pain away as the format is much simpler.
Also, some parts of the documentation are automatically generated by inspecting the code, and it is easier to keep the documentation and code in sync because the are in the same file.

The roxygen2 comments sit above the function, and start with `#'` to distinguish them from ordinary comments.

The first sentence of the roxygen becomes the title of the help file.
Then we can use roxygen tags for the rest of the documentation.
Roxygen tags all start with an `@`.
Once you type this, RStudio gives you suggestions.

`@description` One paragraph description of what the function does.

`@param` argument_name followed by a description of the argument All parameters must be documented.

`@details` All the gory details.
Possibly several paragraphs (separate with a blank line).

`@return` A description of the object returned by the function (if any)

`@examples` Working examples which will be run when the package is tested.
Examples should run relatively quickly or they become tedious.
They need to use `library()` to load any packages needed other than the package being developed.
Any packages loaded by `library()` need to be declared in the DESCRIPTION file (see @sec-importing-exporting-functions)

You can use [markdown](https://roxygen2.r-lib.org/articles/markdown.html) to enhance the documentation, including links etc.

### Importing functions {#sec-importing-exporting-functions}

The roxygen comments are also where you import functions from other packages.
This information will be written into the NAMESPACE file so you don't have to (see the [R package book](https://r-pkgs.org/namespace.html#namespace) for more information about this file).

If you want to use a function from any other package (except `base`) you need to add the dependency to the package to the DESCRIPTION file.

```{r use-package}
use_package("dplyr")
#use_dev_package("packageFromGithub")
```

You can now use a function from `dplyr` with `dplyr::mutate`.
This gets messy if you need to use lots of functions from a package.
Then it is better to import them.
We could use

    #' @import dplyr

to import all functions in the package, or

    #' @importFrom dplyr %>%

if we want to just import specific functions.
This is safer as it minimises the risk of conflicts.

Forgetting to import functions or to declare the dependencies in the DESCRIPTION file are very common problems when checking the package, but the message is informative (@troubleshooting).

### Exporting functions

If you want to make your function available to users you normally need to export them. 
You can do this by adding `@export` to the roxygen comments.

Functions extending S3 generics (e.g. `plot.myclass()`) don't need to be exported, but it makes it easier for the user to find your function and access the help file.

Other functions that are not exported are internal to the package, they can be accessed by the user with `mypackage:::unexported()` but they should not need to do this.

### Generate the .Rd files

Convert the roxygen comments to .Rd files with

```{r document}
devtools::document() 
```

View the compiled help files by running

```{r load-all}
devtools::load_all()
```

which simulates loading the package without having to install it properly.
Then you can use `?` as normal to get the help (links won't work)

## Data

Many packages include data, either because they are needed for examples, or because the aim of the package is to distribute a dataset (probably with some functions to access it).

Data are stored in `data/` as `.rda` files and can be loaded into R with the `data()` function.

```{r penguins, eval = FALSE}
data("penguins", package = "palmerpenguins")
```

The best way to add `.rda` files is to use the function `use_data_raw()` with the name of the dataset you want create.

```{r, eval = FALSE}
use_data_raw(name = "dataset1")
```

This will create `data_raw/` and add a file called `dataset1.R` that looks like this.

```{r data_set1, eval = FALSE}
## code to prepare `dataset1` dataset goes here

usethis::use_data(dataset1, overwrite = TRUE)
```

Add the code needed to reproducibly create `dataset1`, perhaps importing and processing data files that you have copied into `data-raw/`.

When you run the entire script, the code will create `dataset1.rda` in `data/`.

### Documenting data

Data in `data/` need to be documented.
Write the documentation for the data using roxygen2 and save it in `R/`.
I usually call this file `data.R` and document all the datasets in it.

In my traitstrap package, one of the data objects is called trait.
Here is the entry from `R/data.R`:

    #' Trait data
    #'
    #' A dataset containing plant traits in control plots on Svalbard from PFCT4
    #' TraitTrain course.
    #'
    #' @format A data frame with 53940 rows and 10 variables:
    #' \describe{
    #'   \item{Taxon}{species name}
    #'   \item{Site}{site name}
    #'   \item{PlotID}{plot name}
    #'   \item{Trait}{trait name with unit}
    #'   \item{Value}{trait value}
    #' }
    #' @source \url{http://https://www.uib.no/en/rg/EECRG/114808/plant-functional-traits-course-4}
    "trait"

The first line gives the title followed by the desciption.
The \@format field lets you describe the data.
The last line has the object name in quotes (note no #' first).

## Adding other files

R gets upset if there are files where they are not supposed to be in the R package.
You cannot put extra files into the package root directory.
The solution is to put extra files into subdirectories in `inst/`.

If you want to include raw data files because your package has functions to process them, they can go in `inst/extdata/`.

When the package is compiled, `inst/` will be removed and `extdata/` will be put into the package root directory.
The path to the file can the be accessed with `system.file()`.

```{r system-file}
path <- system.file("extdata", package = "readr")
list.files(path)
```

Tutorials and other resources can also be put into suitable directories in `inst/`.

## Checking {#sec-checking}

Now we are ready to check the package compiles correctly.
The check function is slow (it checks many aspects of the package), so before running it, run some functions that check different aspects of the package.

```{r pre-check}
# build the documentation
devtools::document()

# check the examples work
devtools::run_examples()

# check the tests work
devtools::test()
```

After fixing any problems, check the entire package either with the `Check` button in the Build tab (in the same panel as the Environment tab), or with `devtools::check()`.
They do exactly the same work, but the Check button leaves the console available.

Check is very thorough.
It will almost certainly report errors, warnings or notes the first times you run it.
Identify the problems, fix them and run check again.

Now use

```{r load-all2}
devtools::load_all()
```

and give your functions a test drive.

::: callout-caution
## Don't `source()` functions

You must resist the temptation to `source()` your functions (or otherwise load them through the RStudio interface). 

This will cause problems for `devtools::load_all()` as the sourced functions will mask the versions loaded by `load_all()`, and may not work properly. 
Either delete any functions you have sourced, or restart your R session.

:::

## Build and install {#sec-build-and-install}

When you are happy with your package, use the "Install" button in the build tab to install your package.
Now you can load it with `library()` as you would any other package.
