---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: Post Hoc Tests - multiple comparisons in linear models [factorial design]
# author: jonathan
# revised: 14/4/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/post-hoc-tests/post-hoc-tests-lm2.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/post-hoc-tests/post-hoc-tests-lm2.Rmd")
library(ggplot2)
library(emmeans)
```

You have been running a [two-way ANOVA](https://biostats.w.uib.no/factorial-design/){target="_blank"} with `lm()` and `anova()`, and the F-test has revealed a main effect of at least one of the factors, and possibly an interaction. You would like to investigate further to find out which group means are significantly different from each other. To do so, you will need to run the appropriate *post-hoc* tests to assess the significance of differences between pairs of group means. The function `emmeans()` will help you do this.

We will reuse the example introduced [here (factorial design (two-way ANOVA))](https://biostats.w.uib.no/factorial-design/){target="_blank"}.
To create the corresponding dataframe, use the following code:
```{r dataframe}
# response variable
size <- c(25, 22, 28, 24, 26, 24, 22, 21, 23, 25, 26, 30, 25, 24, 21, 27, 28, 23, 25, 24, 20, 22, 24, 23, 22, 24, 20, 19, 21, 22, 24, 27, 26, 24, 25, 27, 22, 28, 25, 24, 27, 29, 26, 27, 25, 27, 28, 24, 24, 26, 21, 23, 25, 20, 25, 23, 25, 19, 22, 21)

# predictor variables
location <- as.factor(c(rep("ForestA", 10), rep("ForestB", 10), rep("ForestC", 10), rep("ForestA", 10), rep("ForestB", 10), rep("ForestC", 10)))
year <- as.factor(c(rep("2005", 30), rep("2015", 30)))

# dataframe
my.dataframe <- data.frame(size, location, year)
```
<br/><br/>

For recall, here is the corresponding plot:
```{r boxplot, echo=FALSE}
ggplot(my.dataframe, aes(x = location, y = size, fill = year)) +
  geom_boxplot()
```
<br/><br/>

and the model that was fitted with `lm()` was:
```{r model}
# linear model stored in the object 'model'
model <- lm(size ~ location + year, data = my.dataframe)
```
The analysis of variance had shown that there is a main effect of `location`, but not `year`, and no interaction `location:year` (hence the use of `+` instead of `*`).
<br/><br/>

### Running the test with `emmeans()`

`emmeans()` is part of the package `emmeans`, which we first need to activate:
```{r package, eval=FALSE}
library(emmeans)
```
<br/>

The next step consists in "feeding" the linear model to `emmeans()`. The syntax is `emmeans(model, pairwise ~ predictor1 * predictor2)` where `predictor1` and `predictor2` are the predictor variables which levels have to be compared pairwise:
```{r emmeans}
# running emmeans()
emmeans(model, pairwise ~ location * year)
```
The lower part of the table shows all 15 pairwise comparisons and the corresponding p-values.
<br/><br/>

However, this full analysis may only be done if the model originally includes the interaction term `location:year`. In our analysis, there was no interaction found, only a main effect of `location`. We may thus use the syntax `emmeans(model, pairwise ~ predictor1 | predictor2)` to obtain pairwise comparisons of all levels of `predictor1` presented separately for each level of `predictor2`:
```{r emmeans2}
# running emmeans()
emmeans(model, pairwise ~ location | year)
```
This new table shows all three pairwises comparisons (and corresponding p-values) of `ForestA`, `ForestB` and `ForestC` first for `2005`, and then for `2015`.

Another option to obtain pairwise comparisons is to run the result of `emmeans(model, ~ predictor1 * predictor2)` through the function `pairs()` with the argument `simple = "each"`. This creates a table with pairs of means grouped by each single factor. The syntax is `pairs(result, simple = "each")`: 
```{r emmeans3}
# running pairs()
emm <- emmeans(model, ~ location * year)
pairs(emm, simple = "each")
```
<br/><br/>

If you are not interested in the full table, but only the part showing the breakdown based on the levels of a specific factor, run `pairs(emm, simple = "predictor")` instead:
```{r emmeans4}
# running pairs()
emm <- emmeans(model, ~ location * year)
pairs(emm, simple = "location")
```

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
