---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a 2D density plot
# author: jonathan
# revised: 21/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/2D-density.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/2D-density.Rmd")
library(ggplot2)
library(gridExtra)
```
When a scatter plot becomes so dense that the symbols overlap massively, the plot lacks clarity and loses its purpose. This is called overplotting. In such a case, there are a few alternatives, one of them being a 2D density plot (the 2D version of the density plot described [HERE](https://biostats.w.uib.no/creating-a-density-plot/){target="_blank"}). The 2D density plot brings an extra dimension to the scatter plot by allowing to visualize the density of observations with a colored scale.

Let's take the following example where the variable to be plotted along the X-axis (`values1`) and the variable to be plotted along the Y-axis (`values2`) both contain 4000 points (!). Here are the variables and the dataframe:
```{r dataframe}
# variable 1
values1 <- rnorm(4000, mean=50, sd=10)
# variable 2
values2 <- rnorm(4000, mean=5, sd=1)
# dataframe
df <- data.frame(values1, values2)
```
<br/><br/>

If we attempt to draw a regular scatter plot with `geom_point()`, we obtain this unreadable figure:
```{r scatterplot}
ggplot(df, aes(values1, values2)) +
  geom_point()
```
<br/><br/>

In such a case of overplotting, it is logical to switch to a different plot type.
To draw a 2D density plot, we will either use `stat_density_2d()` or `geom_density_2d()` with a series of necessary arguments. Here is an example with `stat_density_2d()`:
```{r 2D stat}
ggplot(df, aes(values1, values2)) +                       
  stat_density_2d(aes(fill = ..level..), geom = "polygon") 
```

And here is the same example with `geom_density_2d()`:
```{r 2D geom}
ggplot(df, aes(values1, values2)) +                       
  geom_density_2d(aes(color = ..level..))
```
A color bar guide named `level` automatically appears to the right thanks to the parameter `..level..`. It shows the different shades of blue (color by default) used in the figure to help you define approximately the density level in the figure.
<br/><br/>

If you feel like changing the colors used in the color bar guide, you can use `scale_fill/color_gradient()` or `scale_fill/color_viridis_c()`:
```{r 2Dviridis, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values1, values2)) +                         # left plot with stat_density_2d
    stat_density_2d(aes(fill = ..level..), geom = "polygon") +
    scale_fill_viridis_c(), 
  ggplot(df, aes(values1, values2)) +                         # right plot with geom_density_2d
    geom_density_2d(aes(color = ..level..)) +
    scale_color_viridis_c(),
  ncol=2)
```
<br/><br/>

Note that it is possible to highlight the borders between the levels in the `stat_density_2d` plot by adding contours. To do so, you simply have to add the arguments `color=` to the function `stat_density_2d()`. You may also modulate the thickness of these contour lines by adding `size=`:
```{r borders}
ggplot(df, aes(values1, values2)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour="red", size = 1.5) +
  scale_fill_viridis_c()
```
<br/><br/>

The same applies to lines in the `geom_density_2d` plot:
```{r lines}
ggplot(df, aes(values1, values2)) +
  geom_density_2d(aes(color = ..level..), size=1.5) +
  scale_color_viridis_c()
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
