---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a bubble plot
# author: jonathan
# revised: 20/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/bubble-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/bubble-plot.Rmd")
library(ggplot2)
```
A regular scatter plot represents with dots (or any other symbols) the relationship between two given variables. One can plot a third variable to any scatter plot by the mean of a gradient of colors applied onto the dots and thus obtain a [color-graded scatter plot](https://biostats.w.uib.no/creating-a-color-graded-scatter-plot/){target="_blank"}. But one can also plot this third variable by the mean of the size of the dots or circles representing the data. The results looks thus like bubbles in a chart.

We will see how to use `ggplot()` to code for such a bubble plot. Here are the three variables and the dataframe:
```{r dataframe}
# variable 1
values1 <- c(rnorm(100, mean=50, sd=15), rnorm(100, mean=75, sd=15), rnorm(100, mean=100, sd=15))
# variable 2
values2 <- c(rnorm(100, mean=20, sd=10), rnorm(100, mean=30, sd=10),rnorm(100, mean=40, sd=10))
# variable 3
values3 <- runif(300, min = 1, max = 6)
# dataframe
df <- data.frame(values1, values2, values3)
```
<br/><br/>

We use `geom_point()` as for any scatter plot, but we need to add `aes(size=)` to represent the values of the variable `values3` as a set of bubbles of varying size:
```{r plot}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3))
```
<br/><br/>

With the argument `alpha=`, we can add transparency to the bubbles:
```{r bubbles}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3), alpha = .3)  
```
<br/><br/>

And as always, it is possible to add a bit of color to the plot:
```{r color}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3), color = "blue", alpha = .3)  
```
<br/><br/>

But it is also possible to color-grade the bubbles as a function of the variable `values3` with `color=` (in the same way as with `size=`). For aesthetic reasons, we will use the palette viridis brought in by `scale_color_viridis_c()`:
```{r colors}
ggplot(df, aes(values1, values2)) +
  geom_point(aes(size = values3, color = values3), alpha = .5) +
  scale_color_viridis_c()
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->

