---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a frequency polygon
# author: jonathan
# revised: 18/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/frequency-polygon.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/frequency-polygon.Rmd")
library(ggplot2)
library(gridExtra)
```
A frequency polygon and a histogram are very much alike. Both help visualize the distribution of a data series, the former using bars to represent counts and the latter using lines.

Let’s use `ggplot()` to draw the frequency polygon for a data set generated by `rnorm()` (read more [here](https://biostats.w.uib.no/4-generating-random-series/){target="_blank"} about `rnorm()`). Here are the variables and dataframe:
```{r dataframe}
# ID
ID <- 1:200
# sample data
values <- rnorm(200, mean=65, sd=15)
# dataframe
df <- data.frame(ID, values)
```
<br/><br/>

We first map the data from the variable `values` by typing `ggplot(df, aes(values))` and then use `geom_freqpoly()` to draw the plot:
```{r plot}
ggplot(df, aes(values)) + 
  geom_freqpoly()
```
<br/><br/>

To realize how similar a frequency polygon and a histogram are, we can put them next to each other:
```{r plot2, fig.width=10, fig.height=4}
grid.arrange(
  ggplot(df, aes(values)) +          # histogram
    geom_histogram(bins = 30),
  ggplot(df, aes(values)) +          # frequency polygon
    geom_freqpoly(bins = 30),
  ncol=2
)
```
The shapes of these two plots are not strictly identical but they clearly show the same pattern of distribution. 
<br/><br/>

As for any histogram, we can modify the binwidth or the number of bins using `binwidth=` or `bins=` in `geom_freqpoly()`. Here are two examples:
```{r plot3, fig.width=10, fig.height=4} 
grid.arrange(
  ggplot(df, aes(values)) + 
    geom_freqpoly(bins = 60),          # left plot, changing bins
  ggplot(df, aes(values)) + 
    geom_freqpoly(binwidth = 10),      # right plot, changing binwidth
  ncol=2)
```
<br/><br/>

As usual, you may change the look of the line with `size=`, `color=` and `linetype=`:
```{r plot4}
ggplot(df, aes(values)) + 
  geom_freqpoly(bins = 60, size = 1.5, color = "darkblue", linetype = "dotted")
```

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
