---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a horizontal bar plot
# author: jonathan
# revised: 18/3/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Visualizing/Simple/horizontal-bar-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Visualizing/Simple/horizontal-bar-plot.Rmd")
library(ggplot2)
```
A bar plot represents the relationship between a measurement variable and a categorical variable. In most cases, each of the bars will display the mean of a specific group, which will allow for visualizing the difference between groups in an experiment.

There are two variants of the bar plot: horizontal and [vertical](https://biostats.w.uib.no/creating-a-bar-plot/){target="_blank"}. One would prefer to use a horizontal bar plot when the categorical variable is nominal (labels/names), while the vertical plot is preferred when the categorical variable is ordinal (numbers, series, dates). Here, we will see how to make a horizontal bar plot.

In the following example, we will use `ggplot()` to draw a bar plot that shows the total precipitations registered in 2018 at two locations (Lygra, Hordaland and Østerbø, Sogn og Fjordane). Here are the variables and dataframe:
```{r dataframe}
# variable 1
location <- c("Lygra", "Østerbø")
# variable 2
precipitations <- c(1229.8, 515.9)
# dataframe
df <- data.frame(location, precipitations)
```
<br/><br/>
We start by drawing a regular, vertical bar plot. We may use either `geom_col()` or `geom_bar()` to create the plot with `ggplot()`, as previously discussed [here](https://biostats.w.uib.no/creating-a-bar-plot/){target="_blank"}. Here we use `geom_col()`:
```{r plot}
ggplot(df, aes(location, precipitations)) +     
  geom_col()
```
<br/><br/>

To obtain a horizontal bar plot, we will add the function `coord_flip()`. This function flips the whole plot so that the X-axis becomes the Y-axis and vice versa:
```{r flip}
ggplot(df, aes(location, precipitations)) +      
  geom_col() +
  coord_flip()                  
```
<br/><br/>

You may bring colors to the bars using `color=` and `fill=`:
```{r color}
ggplot(df, aes(location, precipitations)) +     
  geom_col(color = "blue", fill = "white", width = .5) +
  coord_flip()       
```
<br/><br/>

Finally you may adjust the width of the bars with the argument `width=`:
```{r width}
ggplot(df, aes(location, precipitations)) +     
  geom_col(width = .25) +
  coord_flip()       
```
<br/><br/>

### Adding plot title, axis titles, ticks, labels and other essential elements

In [this section](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}, you will learn how to set/modify all the necessary elements that make a plot complete and comprehensible. Such elements are:

+ [plot title](https://biostats.w.uib.no/title/){target="_blank"}, 
+ [axis title](https://biostats.w.uib.no/axis-titles/){target="_blank"},
+ [axis scale](https://biostats.w.uib.no/axis-scale/){target="_blank"},
+ [axis ticks](https://biostats.w.uib.no/axis-ticks/){target="_blank"},
+ [category labels](https://biostats.w.uib.no/category-labels/){target="_blank"},
+ [legend](https://biostats.w.uib.no/legend/){target="_blank"},
+ [secondary Y-axis](https://biostats.w.uib.no/secondary-y-axis/){target="_blank"},
+ [colors](https://biostats.w.uib.no/choosing-colors-for-your-plot/){target="_blank"},
+ [etc](https://biostats.w.uib.no/topics/data-visualization-with-ggplot2/essential-elements/){target="_blank"}.

<br/><br/>
***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
