---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a pie chart
# author: jonathan
# revised: 21/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/pie-chart.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/pie-chart.Rmd")
library(ggplot2)
library(gridExtra)
```
There is no such thing as a `geom_pie()` function that creates a pie chart in ggplot. Instead of that, one must use a "weird" strategy which consists in building first a *single stacked bar plot* from a dataframe, and then displaying it in a *polar coordinate system*. The stacked bar will thus be stretched to the point of becoming a pie. This reminds in a way of the strategy used when one wants to create a miniature planet from a panorama picture using the "[little planet effect](https://www.photographymad.com/pages/view/little-planet-photos-5-simple-steps-to-making-panorama-worlds){target="_blank"}".

Let's take a simple example with this data set (blood type distribution in Norway, data retrieved [here](https://en.wikipedia.org/wiki/Blood_type_distribution_by_country){target="_blank"}, original source: GiBlod.no). Here are the variables and the dataframe:
```{r dataframe}
# variable 1
type <- c("O+", "A+", "B+", "AB+", "O-", "A-", "B-", "AB-")
# variable 2
proportion <- c(33.2,	41.6,	6.8,	3.4,	5.8,	7.4,	1.2,	0.6)
# dataframe
df <- data.frame(type, proportion)
```
<br/><br/>

In this example, we will decompose the procedure into four steps:

+ Building the stacked bar chart from the dataframe
+ Changing the coordinate system
+ Getting rid of the useless axis
+ Getting a better set of colors.

<br/><br/>

### 1. Building the stacked bar chart from the dataframe

We use `geom_col()` to draw the bar plot and map the variable with `aes()`. Here, `x=` is set but not named by the mean of `""`, while `y =` is set to `proportion`. Finally we use `fill=` to attribute a specific color to each blood type:
```{r plot}
ggplot(df, aes(x="", y=proportion, fill=type))+
  geom_col()
```
<br/><br/>

### 2. Changing the coordinate system

Now we need to exchange the cartesian coordinate system in a polar coordinate system. In this way, the stacked bar will be stretched around a central point of focus. For that we add `coord_polar("y")` to the code:


```{r polar}
ggplot(df, aes(x="", y=proportion, fill=type))+
  geom_col() +
  coord_polar("y") 
```
<br/><br/>

### 3. Getting rid of the useless axis

The circular axis around the chart has very little meaning; it may thus be taken away. One way to do it is to remove the whole *theme*, which means that the grey background disappears as well. To do so, we just add `theme_void`:
```{r theme}
ggplot(df, aes(x="", y=proportion, fill=type)) +
  geom_col() +
  coord_polar("y") +
  theme_void()
```

<br/><br/>

### 4. Getting a better set of colors.

To be honest, the set of colors which comes automatically in `ggplot()` is not very appetizing. Let's use `scale_fill_viridis_d()` to exchange the 8 original colors against 8 colors from the palette viridis:
```{r colors}
ggplot(df, aes(x="", y=proportion, fill=type))+
  geom_col() +
  coord_polar("y") +
  theme_void() +
  scale_fill_viridis_d()
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
