---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a Q-Q plot
# author: jonathan
# revised: 20/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/QQplot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/QQplot.Rmd")
library(ggplot2)
```
A Q-Q plot (or Quantile-Quantile plot) is used to compare quantile distribution of a sample to a theoretical quantile distribution. Often, this is used to visually estimate whether a sample distribution is (close to) normal, in which case the quantiles are nicely aligned in the plot. Note that this type of visual estimation does not replace a proper statistical [test for normality](https://biostats.w.uib.no/test-for-normality-shapiro-wilks-test/){target="_blank"}.

We will see how to use `ggplot()` to code for a Q-Q plot representing a sample which is normally distributed. Here is the sample and dataframe:
```{r dataframe}
# ID
ID <- 1:300
# sample
values <- rnorm(300, mean=50, sd=5)
# dataframe
df <- data.frame(ID, values)
```
<br/><br/>

In ggplot2, Q-Q plots are drawn by the functions `stat_qq()` and `geom_qq()`. Here we use `stat_qq()`:
```{r plot}
ggplot(df, aes(sample=values)) +
  stat_qq() 
```
<br/><br/>

It is of course possible to tune the look of the plot with, for example, `size=` and `color=`:
```{r tune}
ggplot(df, aes(sample=values)) +
  stat_qq(size = 2, color = "blue")   
```
<br/><br/>
To better verify the alignment of the dots, one can add a *quantile-quantile line* by the mean of `stat_qq_line()`. This line can also be tuned with similar arguments:
```{r line}
ggplot(df, aes(sample=values)) +
  stat_qq(size = 2, color = "blue") + 
  stat_qq_line(size=1.25, color="red")  
```

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
