---
output:
  html_document:
    css: custom.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Creating a scatter plot
# author: jonathan
# revised: 16/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Simple/scatter-plot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Simple/scatter-plot.Rmd")
library(ggplot2)
```
A scatter plot is probably one of the simplest forms of plot that can be drawn. The data is represented in the form of dots (alternatively squares, crosses, circles or other simple geometrical shapes), one per data point, providing that the data can be described by two variables, one plotted along the X-axis and the other along the Y-axis.

Let's take a simple example to see how to build a scatter plot with `ggplot()`.
```{r dataframe}
# variable 1
x <- 1:100
# variable 2
y <- x*(rnorm(100, mean=1, sd=.1))
# dataframe
df <- data.frame(x,y)
```
<br/><br/>
Let's map the data from the variable `x` and `y` by typing `ggplot(df, aes(x, y))` and use `geom_point()` to draw the scatter plot:
```{r plot}
ggplot(df, aes(x,y)) +
    geom_point()
```
<br/><br/>
You will certainly say that this scatter plot is rather boring; it could have colors and larger dots for example. We can add a few arguments such as `size=` and `colour=` to `geom_point()` to make the dots larger and blue. 
```{r plot2}
ggplot(df, aes(x,y)) +
    geom_point(size=2, colour="blue")
```
<br/><br/>
You can also change the symbols by introducing the argument `shape=`. Check [this table](https://www.statmethods.net/advgraphs/images/points.png){target="_blank"} and pick the number that matches the shape you want. 
```{r plot3}
ggplot(df, aes(x,y)) +
  geom_point(size=2, colour="blue", shape=17)
```
<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
