---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: plot title
# author: jonathan
# revised: 11/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Essential/plot-title.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Essential/plot-title.Rmd")
library(ggplot2)
```
Axis titles are important for the clarity of a graph. They give the reader information about the variables which are represented and their unit (if any). A common mistake is to omit the unit, thus leading to misinterpretation of the data and aberrant conclusions. Axis titles should be clear, short but descriptive enough to leave no doubt about the variables.

In ggplot2, there are several ways to assign a title to an axis, and to change its look. Here we will see a couple of them using a simple example.

Let’s see that with the scatter plot below. This plot is the result by the following code and is stored in the object `baseplot` so that we can reuse it throughout the whole tutorial:

```{r variables and plot}
# variable 1
values1 <- rnorm(200, mean=50, sd=10) 

# variable 2
values2 <- c(rnorm(100, mean=20, sd=10), rnorm(100, mean=40, sd=10))

# dataframe
df <- data.frame(values1, values2)

# scatter plot
baseplot <- ggplot(df, aes(values1, values2)) +
  geom_point(size=2) 
baseplot
```
<br/><br/>
As you may see here, `ggplot()` has already taken care of placing the variable names (i.e. `values1` and `values2`) as axis titles. The whole point here is to override this action by default, and replace `values1` by `variable1 (unit)`, and `values2` by `variable2 (unit)`.


### Adding axis titles

Here we use the functions `xlab()` and `ylab()` to add the new titles to the axes:

```{r labs}
baseplot +
  xlab("variable1 (unit)") +
  ylab("variable2 (unit)")
```
<br/><br/>
The functions `scale_x_continuous()` and `scale_y_continuous()` (alternatively `scale_x_discrete()` and `scale_y_discrete()` if your variables are *discrete* instead of *continuous*) may be use instead of *xlab()* and *ylab()*:

```{r scale}
baseplot +
  scale_x_continuous("variable3 (unit)") +
  scale_y_continuous("variable4 (unit)")
```
<br/><br/>

### Tuning the look of the axis titles

If you wish to change the size and/or color of the axis titles, you won’t be able to do it using the previous functions. Instead, you will have to modify the *theme* of the plot itself. To do so, you will have to add an additional line to the code and use the function `theme(axis.title = element_text())`:

```{r theme}
baseplot +
  xlab("variable1 (unit)") +
  ylab("variable2 (unit)") +
  theme(axis.title = element_text(color = "blue", size = 12))
```
<br/><br/>

You can alternatively modify only one of the axis titles, or both in a different manner, by replacing `axis.title` with `axis.title.x` and/or `axis.title.y`:

```{r theme2}
baseplot +
  xlab("variable1 (unit)") +
  ylab("variable2 (unit)") +
  theme(axis.title.x = element_text(color = "blue", face ="bold", size = 12)) +
  theme(axis.title.y = element_text(color = "red", size = 14))
```

<br/><br/>


***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
