---
output:
  html_document:
    css: custom-blockquote.css
--- 

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center')

# title: Choosing colors for your plot
# author: jonathan
# revised: 13/2/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Visualizing/Essential/Choosing-colors.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Visualizing/Essential/Choosing-colors.Rmd")
library(ggplot2)
library(gridExtra)

# data for violin plot
# variable 1
values <- c(rnorm(150, mean=20, sd=5), rnorm(150, mean=30, sd=4), rnorm(150, mean=15, 3), rnorm(150, mean=40, sd=5)) 
# variable 2
category <- c(rep("Gr1", 150), rep("Gr2", 150), rep("Gr3", 150), rep("Gr4", 150))
# dataframe
df <- data.frame(category, values)

## data for density plot
# variable 3
values1 <- rnorm(4000, mean=50, sd=10) 
# variable 4
values2 <- rnorm(4000, mean=5, sd=1)
# dataframe
df2 <- data.frame(values1, values2)
```
There are several ways to bring colors to your plot in ggplot2. You may:

+ let ggplot2 use its own default color palette, 
+ pick a predefined palette, 
+ pick single colors.

We will see how to modify colors in a plot using simple examples. Let’s start with the code for two reference plots, a density plot and a violin plot. These plots are stored in the objects `baseplot1` and `baseplot2` so that we can reuse them throughout the whole tutorial:
```{r code, fig.width=10, fig.height=4}
# code for violin plot
baseplot1 <- ggplot(df, aes(category, values, fill=category)) +
  geom_violin()

# code for density plot
baseplot2 <- ggplot(df2, aes(values1, values2)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon")

```
<br/><br/>

### Color palette by default

ggplot2 automatically generates a palette based on the number of items to color (number of categories, etc.) and a color wheel. You may read more about this palette [HERE](http://applied-r.com/colors-in-ggplot/){target="_blank"}. The aim is to provide you with a “well-balanced” set of colors which can easily be distinguished from each other without drawing to much attention onto one specific items. Here are our reference plots colored by ggplot2:
```{r default, fig.width=10, fig.height=4}
grid.arrange(
  baseplot1,
  baseplot2,
  nrow=1)
```
The colors are “OK” as long as you do not have more than a handful of groups, but probably this reaches its limits in terms of aesthetics when handling a larger number of groups.
<br/><br/>

### The viridis color palettes

The viridis package contains a series a colorful palettes which were created to bring more flashy colors to your plots, while providing better distinction between groups. The palettes were also designed so that the color differences would remain perceivable by colorblind people. You may read more about this palette [HERE](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html){target="_blank"}.
To use viridis in your figures, simply add `scale_fill_viridis_d()` (when handling *discrete* predictor variables) and `scale_fill_viridis_c()` (when handling *continuous* predictor variables). Let's look at our examples colored with viridis:
```{r viridis, fig.width=10, fig.height=4}
grid.arrange(
  baseplot1 + scale_fill_viridis_d(),
  baseplot2 + scale_fill_viridis_c(),
  nrow=1)
```

Of course, if the code uses `color=` instead of `fill=` for grouping, use `scale_fill_viridis_d()` and `scale_fill_viridis_c()`.

Note that the example above results of the *default* viridis palette. [Other palettes are available](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html#the-color-scales){target="_blank"} and you just have to add their name with `option=` to the function to activate them. The following plots use the palette `Inferno`:
```{r inferno, fig.width=10, fig.height=4}
grid.arrange(
  baseplot1 + scale_fill_viridis_d(option="inferno"),
  baseplot2 + scale_fill_viridis_c(option="inferno"),
  nrow=1)
```

### Brewer/Distiller

Brewer and Distiller work in a similar way. They offer many sets of colors referenced [HERE](https://www.sthda.com/sthda/RDoc/images/rcolorbrewer.png){target="_blank"}. Pick a set and add it to your plot with `scale_fill_brewer` or `scale_color_brewer` (when handling discrete variables) and `scale_fill_distiller` or `scale_color_distiller` (when handling continuous variables). Add `palette=" "` and the name of the palette between the parentheses. 

```{r brewer, fig.width=10, fig.height=4}
grid.arrange(
  baseplot1 + scale_fill_brewer(palette="Set3"),
  baseplot2 + scale_fill_distiller(palette="Set3"),
  nrow=1)
```

### Choose your own colors 

If you are really picky about colors and have the time to select every single color for your graph, use `scale_fill_manual()`, `scale_color_manual()` or `scale_fill_gradient()` and then indicate between the parentheses that either the R name or the *hex* code of each of them. 

Note that `scale_fill_manual()` and `scale_color_manual()` are to be used with discrete variables. Always make sure that you have chosen a color for each of the groups or categories, otherwise, ggplot will return a warning and no plot. `scale_fill_gradient()` shall be used with continuous variablees and will create a gradient based on two given colors. Here you will find the [R names](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf){target="_blank"} and a [tool](https://htmlcolorcodes.com/) to provide you with hex codes.

```{r manual, fig.width=10, fig.height=4}
grid.arrange(
  baseplot1 + scale_fill_manual(values=c("darkblue", "aquamarine1", "lightsalmon2", "orange2")),
  baseplot2 + scale_fill_gradient(low= "darkblue", high= "lightsalmon2"),
  nrow=1)
```

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
