---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: One-way ANOVA
# author: jonathan
# revised: 12/4/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/ANOVA/1way.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/ANOVA/1way.Rmd")
library(ggplot2)

```

One-way ANOVA is a parametric test designed to **compare the means of two or more samples**. The null hypothesis `H0` states that the means of all samples to be tested are equal. The test returns an F-statistic and a p-value which will help you decide whether or not to reject the null hypothesis.

One of the options to perform a one-way ANOVA in R is to run the function `lm()`;  followed by `anova()`. This option fits a linear model and will work in virtually all cases. 

Let's take an example. Here, let's say that we want to check whether the average size of blue ground beetles (*Carabus intricatus*) differs depending on their location. We consider 3 different locations, for example 3 forests beautifully named `A`, `B` and `C`. In each location, we measure the size (in millimeters) of 10 individuals.

To create the corresponding dataframe in R, use the following code:
```{r dataframe}
# response variable
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22)

# predictor variable
location <- as.factor(c(rep("ForestA",10), rep("ForestB",10), rep("ForestC",10)))

# dataframe
my.dataframe <- data.frame(size,location)
```
<br/><br/> 

It is always nice and useful to start by visualizing the whole dataset, so let's plot the data:
```{r boxplot, echo=FALSE}
ggplot(my.dataframe, aes(x = location, y = size)) +
  geom_boxplot()
```
<br/><br/> 

### Assumptions

The assumptions are:

+ **independence of observations** (each individual is represented by 1 entry/measurement ONLY)
+ **normality of distribution** (to be tested for each group, for example with the [Shapiro-Wilk test](https://biostats.w.uib.no/test-for-normality-shapiro-wilks-test/){target="_blank"})
+ **homogeneity of variance** (to be tested with, for example, [Levene's test](https://biostats.w.uib.no/test-for-homogeneity-of-variances-levenes-test/){target="_blank"})
+ groups contain [**no outliers**](http://biostats.w.uib.no/finding-outliers/){target="_blank"}.

<br/><br/>

### Running the test

The syntax is `lm(response ~ predictor, data = dataframe)` where `response` is the response variable, `predictor` is the predictor variable or factor (which categorizes the observations) and `dataframe` the name of the dataframe that contains the data. 
We first need to fit the linear model with `lm()` and then we store it in the object `model`. We then compute and display the table for the analysis using `anova()`:
```{r lm}
model <- lm(size ~ location, data = my.dataframe)
anova(model)
```
This output provides you with the F-value (7.1101) and the corresponding p-value (0.003307). The hypothesis `H0` stating that the means of the groups are equal is thus to be rejected.

<br/><br/>

**But this does not tell us anything about the groups which means are significantly different... **

Indeed, the ANOVA needs to be followed by another test if we want to check which of the groups are different from the others. For that, we will need a *post-hoc* test, and these three options will help you do that:

+ [pairwise t-test with the function pairwise.t.test()](http://biostats.w.uib.no/post-hoc-tests-pairwise-comparisons/){target="_blank"},
+ [Tukey’s Honest Significant Difference (HSD) with TukeyHSD()](https://biostats.w.uib.no/post-hoc-tests-tukey-hsd/){target="_blank"},
+ [multiple comparisons in linear models with glht()](https://biostats.w.uib.no/post-hoc-tests-multiple-comparisons-in-linear-models/){target="_blank"}.

<br/><br/>

### Alternative

The non-parametric [Kruskal-Wallis test](http://biostats.w.uib.no/comparing-two-means-kruskal-wallis-test/){target="_blank"} is a good alternative to the one-way ANOVA when the assumption of normality of distribution is violated.

<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
