---
output:
  html_document:
    css: custom-blockquote.css
---

```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: factorial design
# author: jonathan
# revised: 20/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/ANOVA/factorial.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/ANOVA/factorial.Rmd")
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22,24,27,26,24,25,27,22,28,25,24,27,29,26,27,25,27,28,24,24,26,21,23,25,20,25,23,25,19,22,21)
location <- as.factor(c(rep("ForestA",10), rep("ForestB",10), rep("ForestC",10), rep("ForestA",10), rep("ForestB",10), rep("ForestC",10)))
year <- as.factor(c(rep("2005",30), rep("2015",30)))
my.dataframe <- data.frame(size,location,year)
my.dataframe
library(ggplot2)

```

Whereas one-way ANOVA allows for comparison of three and more group means based on the different levels of a single factor, **factorial design** allows for comparison of groups based on **several independent variables and their various levels**. Thus, comparing data (such as tree size, egg sizeâ¦) split in groups according to two variables (for example location and species) may be done via a **two-way ANOVA**; if data are split in groups according to three independent variables (for example location, species and date of measurement), then a **three-way ANOVA** may be usedâ¦ Further below, we will essentially talk about two-way ANOVA , but youâll quickly understand how the functions may be adapted to more factors.

Using multi-way ANOVA, we get to assess **main effects**, in other words the impact of single, independent factors on a given variable (plant size, egg number, individual bodyweightâ¦). We also get to see whether there exist **interactions** between factors, in other words whether the effect of a specific factor is influenced by (one of the) other factors.

As for [one-way ANOVA](http://biostats.w.uib.no/one-way-anova/){target="_blank"}, the function to use is `lm()`, followed by `anova()`.

Letâs see that in an example similar to the one that we looked at in [one-way ANOVA](http://biostats.w.uib.no/one-way-anova/){target="_blank"}. Here, we check whether the average size of blue ground beetles (*Carabus intricatus*) differs depending on their location. We now introduce a new factor: measurements were performed in 2005 and in 2015 at the same 3 forests A, B and C. In each location, we measured the size (in millimeters) of 10 individuals (making it a balanced design). In total, 60 individuals were measured. The two factors are location and year.

Here is the code for creating the dataframe:

```{r dataframe}
size <- c(25,22,28,24,26,24,22,21,23,25,26,30,25,24,21,27,28,23,25,24,20,22,24,23,22,24,20,19,21,22,24,27,26,24,25,27,22,28,25,24,27,29,26,27,25,27,28,24,24,26,21,23,25,20,25,23,25,19,22,21)
location <- as.factor(c(rep("ForestA",10), rep("ForestB",10), rep("ForestC",10), rep("ForestA",10), rep("ForestB",10), rep("ForestC",10)))
year <- as.factor(c(rep("2005",30), rep("2015",30)))
my.dataframe <- data.frame(size,location,year)
```

The assumptions are globally the same as for one-way ANOVA:

+ **independence of observations** (each individual is represented by 1 entry/measurement ONLY)
+ **normality of distribution** (to be tested for each group, for example with the Shapiro-Wilk test)
+ **homogeneity of variance** (to be tested with, for example, [Levene’s test](https://biostats.w.uib.no/test-for-homogeneity-of-variances-levenes-test/){target="_blank"}).
 

We start by visualizing the data with boxplots. Since we have several factors, we’ll group some of the boxes according to these factors.

```{r boxplot}
ggplot(my.dataframe, aes(x = location, y = size, fill = year)) +
  geom_boxplot()
```

Before going further, we need to check for normality of distribution with the [Shapiro-Wilk test](http://biostats.w.uib.no/test-for-normality-shapiro-wilks-test/){target="_blank"}. Since we have 6 groups to check based on two parameters in our dataframe, the syntax is a bit more complicated than usual. For each group , we need to restrict the test to the data matching both a specific `location` and a specific `year`. We write it like this:

```{r normality}
shapiro.test(my.dataframe$size[location=="ForestA" & year=="2005"])
shapiro.test(my.dataframe$size[location=="ForestB" & year=="2005"])
shapiro.test(my.dataframe$size[location=="ForestC" & year=="2005"])
shapiro.test(my.dataframe$size[location=="ForestA" & year=="2015"])
shapiro.test(my.dataframe$size[location=="ForestB" & year=="2015"])
shapiro.test(my.dataframe$size[location=="ForestC" & year=="2015"])
```

None of the tests shows a p-value under 0.05. The null hypothesis stating that the data in the groups are normally distributed is thus accepted.
<br/><br/>

Let’s go on with [Levene’s test](https://biostats.w.uib.no/test-for-homogeneity-of-variances-levenes-test/){target="_blank"} for homogeneity of variance. Do not forget to activate the package `car` prior to the test otherwise `“Error: could not find function “leveneTest”` will be the only answer that you will get:


```{r variances}
library(car)
leveneTest(size~location*year, data=my.dataframe, center=mean)
```
The p-value is largely greater than 0.05. We thus assume that variances are equal.

<br/><br/> 

#### Let’s go for the ANOVA test!

We use the functions `lm()` and `anova()` the following way:

```{r lm}
results.lm <- lm(size ~ location*year)
anova(results.lm)
```

The table above shows the F value and p-value for each of the main effects, `location` and `year`, as well as for the interaction `location:year`. It also conveniently indicates with stars where significance levels are reached. For instance, `size` is significantly affected by the factor `location`, and the p-value is very small (p<0.001, ***).
<br/><br/>

**But this does not tell us anything about the groups which means are significantly different…**

Indeed, this test tells you about main effects and interactions, but does not tell you which groups are significantly different from others. For that, we will need *post hoc* tests such as [Tukey HSD](http://biostats.w.uib.no/post-hoc-tests-tukey-hsd/){target="_blank"}.


<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
