```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: boxplot
# author: jonathan
# revised: 8/1/2020 

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Descr/boxplot.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Descr/boxplot.Rmd")
```

A boxplot is a convenient little plot that brings up the “essentials” about your sample. In R, a boxplot can be built within seconds using the function `boxplot()`.

Here is a simple example:

```{r boxplot}
my.dataset <- c(1,2,3,4,5,6,7,8,9,10)
boxplot(my.dataset)
```

Not sure what this boxplot means? Check out the illustration above. The plot contains a box with whiskers on top and below, as well as a “thick line somehow in the middle”. That thick line represents the [**median**](https://biostats.w.uib.no/2-median/){target="_blank"} of the sample. The bottom and top of the box are defined by the first and third [**quartiles**](https://biostats.w.uib.no/6-quartiles/){target="_blank"} Q1 and Q3, respectively. The whole box thus frames the interquartile range (IQR). The whiskers below and on top of the box correspond to the minimum and maximum values if your sample is free for outliers. Otherwise, the function `boxplot()` defines their length as equal to 1.5 times the IQR. If any, outliers will be identified using open circles (not represented here).
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
