```{r setup&variables, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# title: quartiles
# author: jonathan
# revised: 8/1/2020

WEB <- c("https://bioceed.uib.no/dropfolder/bioSTATS/Rmd/Basic/Descr/quart.html")
GIT <- c("https://github.com/jonathansoule/biostats/blob/master/Basic/Descr/quart.Rmd")
```

Consider any dataset, sort its entries from smallest to largest and split the resulting list in 4 equal subsets. The **quartiles** are the values of the dataset that cut it off in 4.
 
Quartiles are called:

  + **Q1**: the first quartile under which the first 25% of the data in the set can be found,
  + **Q2**: the second quartile under which the first 50% of the data in the set can be found (by the way, Q2 is also the median),
  + **Q3**: the third quartile under which the first 75% of the data in the set can be found.

In addition, one also refers to:

  + **Q0** as the minimum value in the set,
  + **Q4** as the maximum value in the set.
  
For info, note that the set of data between Q1 and Q3 (which contains the middle 50% of the data) is the **interquartile range (IQR)**.

In R, quartiles may be obtained using the function `quantile()` (NB: this is not a typo, it is quaNtile and not quaRtile, there is a good explanation for it, believe me…). If you choose to use only `quantile()` with no other argument than the vector containing the dataset, R returns Q0, Q1, Q2, Q3, Q4 and Q5:

```{r quartiles}
my.dataset <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
quantile(my.dataset)
```
<br/>
If you only need Q1, Q2 and Q3, say it using decimals as follows:
```{r quartiles2}
quantile(my.dataset, c(0.25, 0.5, 0.75))
```
<br/><br/>

***
<!-- UNCOMMENT TO ACTIVATE
Get the Rmarkdown file corresponding to this page on [GitHub](`r GIT`){target="_blank"}.
-->
