---
editor_options: 
  markdown: 
    wrap: sentence
---


### Output files

The results and figures of an analysis are usually presented in a report or presentation or both.
One or more output files, such as a quarto document and presentations can be added to a targets pipeline.

Each **target** from the pipeline can be loaded and used in the quarto document.
The targets that are used in the quarto document need to be loaded into the current environment.
For this we can use `tar_load()` or `tar_read()`.
The first function is used when a target is used several times.
`tar_read()` is useful if a target is only needed once, e.g. to show a figure.

```{r}
#| label: print-target
#| eval: false
#| echo: true

# print model output
tar_load(model)
summary(model)
```


All R packages that are needed to run the quarto file need to be loaded in the .qmd file.
If you are using a R package exclusively in the quarto script, the package can be loaded only in the quarto file, and does not need to be added to the `_target.R` file.
`targets` always need to be loaded in both files.

The `tarchetypes` packages has to be loaded in the `_target.R` file.
It has a number of useful helper functions including `tar_quarto()` which is needed to render quarto files in the targets pipeline.


::: callout-note
## Exercise

Open the `target_quarto.qmd` file.

Load and print the trait figure using `tar_read()` in the third code chunk (fig-trait-figure).

:::


To add the quarto document to the pipeline, the manuscript has to be rendered.
This is done in the `_targets.R` file in the list of targets using `tar_quarto()`.

```{r}
#| label: render-ms
#| eval: false
#| echo: true

# render ms
tar_quarto(name = manuscript, path = "target_quarto.qmd")
```


::: callout-note
## Exercise

**Add the quarto file to the pipeline**

Check that you have loaded the `tarchetypes` package in the `_target.R` file.

Add `tar_quarto()` in the `_targets.R` file to render the quarto file.

Run `tar_make()` to run the whole pipeline.

:::

Voilà, you have run your first targets pipeline including a quarto file.
Have some 🍰!

